---
title: "08062024_LuminexAnalysis"
author: "Rebecca Valls"
date: "2024-08-06"
output: html_document
---
'
#Install libraries
```{r}
# Install and load necessary packages if not already installed
# install.packages("dplyr")
# install.packages("readr")
# install.packages("broom")
# install.packages("purrr")
# install.packages("hrbrthemes")
# install.packages("geomtextpath")
# install.packages("lmerTest")
# install.packages("ggplot2")
# install.packages("tidyverse")

library(readr)
library(dplyr)
library(broom)
library(purrr)
library(ggplot2)
library(lme4)
library(lmerTest)
library(hrbrthemes)
library(geomtextpath)
library(dplyr)


```

#Load data
```{r}
BacteroidesEstablishment_SerumandLungs_2nonCFand2CF_MasterLuminex <- read_csv("~/Dartmouth College Dropbox/Rebecca Valls/Projects/Project_MalevsFemale_nonCFmice_2023/3_Experiment/3_DataAnalysis/Luminex_Aug2024/BacteroidesEstablishment_SerumandLungs__2nonCFand2CF_MasterLuminex.csv")

luminex_df <- as.data.frame(BacteroidesEstablishment_SerumandLungs_2nonCFand2CF_MasterLuminex)

# Factor the "Treatment" column
luminex_df <- luminex_df %>%
  mutate(Treatment = factor(Treatment, levels = c("Baseline", "PBS", "Bacteroides")))

# Factor the "Sex" column
luminex_df <- luminex_df %>%
  mutate(Sex = factor(Sex, levels = c("M", "F")))


luminex_df$Cage <- as.factor(luminex_df$Cage)




#Subset data into each experiment
#nonCF Exp1
nonCF_Exp1_luminex_df<- luminex_df %>%
  filter(Experiment == "nonCF_Exp1")

#nonCF Exp2
nonCF_Exp2_luminex_df<- luminex_df %>%
  filter(Experiment == "nonCF_Exp2")


#CF Exp1
CF_Exp1_luminex_df<- luminex_df %>%
  filter(Experiment %in% c("CF_Baseline", "CF_Exp1"))

#CF Exp2
CF_Exp2_luminex_df<- luminex_df %>%
  filter(Experiment == "CF_Exp2")


```


#Extract serum and lungs KC and plot for nonCF Experiment 1
```{r}
nonCF_Exp1_serum_KC_luminex_df <- nonCF_Exp1_luminex_df %>%
  filter(SampleType == "Serum",
         Cytokine == "KC")


nonCF_Exp1_serum_KC_dotplot <- ggplot(nonCF_Exp1_serum_KC_luminex_df, aes(x = Treatment, y = Cytokine_pg_per_mL, color = Treatment, shape = Sex)) +
  geom_jitter(width = 0.2, size = 3) +  # Adjust width to avoid overlap
    stat_summary(fun = median, geom = "crossbar", aes(group = Treatment), width = 0.5, fatten = 2) +  # Add median crossbars
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate labels at 45 degrees
  labs(y = "KC in Serum (pg/mL)", color = "Treatment Group")

nonCF_Exp1_serum_KC_dotplot

ggsave("nonCF_Exp1_serum_KC_dotplot.pdf", plot = nonCF_Exp1_serum_KC_dotplot, device = "pdf", width = 4, height = 3)


nonCF_Exp1_lungs_KC_luminex_df <- nonCF_Exp1_luminex_df %>%
  filter(SampleType == "Lungs",
         Cytokine == "KC")


nonCF_Exp1_lungs_KC_dotplot <- ggplot(nonCF_Exp1_lungs_KC_luminex_df, aes(x = Treatment, y = Cytokine_pg_per_mL, color = Treatment, shape = Sex)) +
  geom_jitter(width = 0.2, size = 3) +  # Adjust width to avoid overlap
    stat_summary(fun = median, geom = "crossbar", aes(group = Treatment), width = 0.5, fatten = 2) +  # Add median crossbars
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate labels at 45 degrees
  labs(y = "KC in Lungs (pg/mL)", color = "Treatment Group")

nonCF_Exp1_lungs_KC_dotplot

ggsave("nonCF_Exp1_lungs_KC_dotplot.pdf", plot = nonCF_Exp1_lungs_KC_dotplot, device = "pdf", width = 4, height = 3)




```

#Extract serum and lungs KC and plot for Experiment 2
```{r}
nonCF_Exp2_serum_KC_luminex_df <- nonCF_Exp2_luminex_df %>%
  filter(SampleType == "Serum",
         Cytokine == "KC")


nonCF_Exp2_serum_KC_dotplot <- ggplot(nonCF_Exp2_serum_KC_luminex_df, aes(x = Treatment, y = Cytokine_pg_per_mL, color = Treatment, shape = Sex)) +
  geom_jitter(width = 0.2, size = 3) +  # Adjust width to avoid overlap
    stat_summary(fun = median, geom = "crossbar", aes(group = Treatment), width = 0.5, fatten = 2) +  # Add median crossbars
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate labels at 45 degrees
  labs(y = "KC in Serum (pg/mL)", color = "Treatment Group")

nonCF_Exp2_serum_KC_dotplot

ggsave("nonCF_Exp2_serum_KC_dotplot.pdf", plot = nonCF_Exp2_serum_KC_dotplot, device = "pdf", width = 4, height = 3)


nonCF_Exp2_lungs_KC_luminex_df <- nonCF_Exp2_luminex_df %>%
  filter(SampleType == "Lungs",
         Cytokine == "KC")


nonCF_Exp2_lungs_KC_dotplot <- ggplot(nonCF_Exp2_lungs_KC_luminex_df, aes(x = Treatment, y = Cytokine_pg_per_mL, color = Treatment, shape = Sex)) +
  geom_jitter(width = 0.2, size = 3) +  # Adjust width to avoid overlap
    stat_summary(fun = median, geom = "crossbar", aes(group = Treatment), width = 0.5, fatten = 2) +  # Add median crossbars
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate labels at 45 degrees
  labs(y = "KC in Lungs (pg/mL)", color = "Treatment Group")

nonCF_Exp2_lungs_KC_dotplot
ggsave("nonCF_Exp2_lungs_KC_dotplot.pdf", plot = nonCF_Exp2_lungs_KC_dotplot, device = "pdf", width = 4, height = 3)


```
#Extract serum and lungs KC and plot for CF Experiment 1
```{r}

CF_Exp1_serum_KC_luminex_df <- CF_Exp1_luminex_df %>%
  filter(SampleType == "Serum",
         Cytokine == "KC")


CF_Exp1_serum_KC_dotplot <- ggplot(CF_Exp1_serum_KC_luminex_df, aes(x = Treatment, y = Cytokine_pg_per_mL, color = Treatment, shape = Sex)) +
  geom_jitter(width = 0.2, size = 3) +  # Adjust width to avoid overlap
    stat_summary(fun = median, geom = "crossbar", aes(group = Treatment), width = 0.5, fatten = 2) +  # Add median crossbars
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate labels at 45 degrees
  labs(y = "KC in Serum (pg/mL)", color = "Treatment Group")

CF_Exp1_serum_KC_dotplot

ggsave("CF_Exp1_serum_KC_dotplot.pdf", plot = CF_Exp1_serum_KC_dotplot, device = "pdf", width = 4, height = 3)


CF_Exp1_Lungs_KC_luminex_df <- CF_Exp1_luminex_df %>%
  filter(SampleType == "Lungs",
         Cytokine == "KC")


CF_Exp1_lungs_KC_dotplot <- ggplot(CF_Exp1_Lungs_KC_luminex_df, aes(x = Treatment, y = Cytokine_pg_per_mL, color = Treatment, shape = Sex)) +
  geom_jitter(width = 0.2, size = 3) +  # Adjust width to avoid overlap
    stat_summary(fun = median, geom = "crossbar", aes(group = Treatment), width = 0.5, fatten = 2) +  # Add median crossbars
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate labels at 45 degrees
  labs(y = "KC in Lungs (pg/mL)", color = "Treatment Group")

CF_Exp1_lungs_KC_dotplot

ggsave("CF_Exp1_lungs_KC_dotplot.pdf", plot = CF_Exp1_lungs_KC_dotplot, device = "pdf", width = 4, height = 3)



```

#Extract serum and lungs KC and plot for CF Experiment 2
```{r}

CF_Exp2_serum_KC_luminex_df <- CF_Exp2_luminex_df %>%
  filter(SampleType == "Serum",
         Cytokine == "KC")


CF_Exp2_serum_KC_dotplot <- ggplot(CF_Exp2_serum_KC_luminex_df, aes(x = Treatment, y = Cytokine_pg_per_mL, color = Treatment, shape = Sex)) +
  geom_jitter(width = 0.2, size = 3) +  # Adjust width to avoid overlap
    stat_summary(fun = median, geom = "crossbar", aes(group = Treatment), width = 0.5, fatten = 2) +  # Add median crossbars
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate labels at 45 degrees
  labs(y = "KC in Serum (pg/mL)", color = "Treatment Group")

CF_Exp2_serum_KC_dotplot
ggsave("CF_Exp2_serum_KC_dotplot.pdf", plot = CF_Exp2_serum_KC_dotplot, device = "pdf", width = 4, height = 3)


CF_Exp2_Lungs_KC_luminex_df <- CF_Exp2_luminex_df %>%
  filter(SampleType == "Lungs",
         Cytokine == "KC")


CF_Exp2_lungs_KC_dotplot <- ggplot(CF_Exp2_Lungs_KC_luminex_df, aes(x = Treatment, y = Cytokine_pg_per_mL, color = Treatment, shape = Sex)) +
  geom_jitter(width = 0.2, size = 3) +  # Adjust width to avoid overlap
    stat_summary(fun = median, geom = "crossbar", aes(group = Treatment), width = 0.5, fatten = 2) +  # Add median crossbars
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate labels at 45 degrees
  labs(y = "KC in Lungs (pg/mL)", color = "Treatment Group")

CF_Exp2_lungs_KC_dotplot
ggsave("CF_Exp2_lungs_KC_dotplot.pdf", plot = CF_Exp2_lungs_KC_dotplot, device = "pdf", width = 4, height = 3)


```






#Plot KC in serum and lungs from both nonCF experiments
```{r}
nonCF_serum_KC_luminex_df <- luminex_df %>%
  filter(Genotype == "C57BL_6J",
         SampleType == "Serum",
         Cytokine == "KC")

nonCF_serum_KC_dotplot <- ggplot(nonCF_serum_KC_luminex_df, aes(x = Treatment, y = Cytokine_pg_per_mL, color = Treatment, shape = Sex)) +
  geom_jitter(width = 0.2, size = 1.5) +  # Adjust width to avoid overlap
    stat_summary(fun = median, geom = "crossbar", aes(group = Treatment), width = 0.8, fatten = 5) +  # Add median crossbars
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  scale_shape_manual(values = c("F" = 16, "M" = 17)) +  # Set female to circles and male to triangles
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate labels at 45 degrees
  labs(y = "KC in Serum (pg/mL)", color = "Treatment Group")

nonCF_serum_KC_dotplot
ggsave("nonCF_serum_KC_dotplot.pdf", plot = nonCF_serum_KC_dotplot, device = "pdf", width = 6, height = 3)

nonCF_lungs_KC_luminex_df <- luminex_df %>%
  filter(Genotype == "C57BL_6J",
         SampleType == "Lungs",
         Cytokine == "KC")

nonCF_lungs_KC_dotplot <- ggplot(nonCF_lungs_KC_luminex_df, aes(x = Treatment, y = Cytokine_pg_per_mL, color = Treatment, shape = Sex)) +
  geom_jitter(width = 0.2, size = 1.5) +  # Adjust width to avoid overlap
    stat_summary(fun = median, geom = "crossbar", aes(group = Treatment), width = 0.8, fatten = 5) +  # Add median crossbars
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  scale_shape_manual(values = c("F" = 16, "M" = 17)) +  # Set female to circles and male to triangles
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate labels at 45 degrees
  labs(y = "KC in Lungs (pg/mL)", color = "Treatment Group")


nonCF_lungs_KC_dotplot

ggsave("nonCF_lungs_KC_dotplot.pdf", plot = nonCF_lungs_KC_dotplot, device = "pdf", width = 6, height = 3)

```




#Plot KC in serum and lungs from both experiments
```{r}
CF_serum_KC_luminex_df <- luminex_df %>%
  filter(Genotype == "CftrF508del",
         SampleType == "Serum",
         Cytokine == "KC")

CF_serum_KC_dotplot <- ggplot(CF_serum_KC_luminex_df, aes(x = Treatment, y = Cytokine_pg_per_mL, color = Treatment, shape = Sex)) +
  geom_jitter(width = 0.2, size = 3) +  # Adjust width to avoid overlap
    stat_summary(fun = median, geom = "crossbar", aes(group = Treatment), width = 0.5, fatten = 2) +  # Add median crossbars
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate labels at 45 degrees
  labs(y = "KC in Serum (pg/mL)", color = "Treatment Group")

CF_serum_KC_dotplot
ggsave("CF_serum_KC_dotplot.pdf", plot = CF_serum_KC_dotplot, device = "pdf", width = 4, height = 3)

CF_lungs_KC_luminex_df <- luminex_df %>%
  filter(Genotype == "CftrF508del",
         SampleType == "Lungs",
         Cytokine == "KC")

CF_lungs_KC_dotplot <- ggplot(CF_lungs_KC_luminex_df, aes(x = Treatment, y = Cytokine_pg_per_mL, color = Treatment, shape = Sex)) +
  geom_jitter(width = 0.2, size = 3) +  # Adjust width to avoid overlap
    stat_summary(fun = median, geom = "crossbar", aes(group = Treatment), width = 0.5, fatten = 2) +  # Add median crossbars
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate labels at 45 degrees
  labs(y = "KC in Lungs (pg/mL)", color = "Treatment Group")

CF_lungs_KC_dotplot
ggsave("CF_lungs_KC_dotplot.pdf", plot = CF_lungs_KC_dotplot, device = "pdf", width = 4, height = 3)

```



Run stats for just KC of nonCF Exp1 and 2
```{r}
# Factor the "Sex" column
nonCF_serum_KC_luminex_df <- nonCF_serum_KC_luminex_df %>%
  mutate(Sex = factor(Sex, levels = c("M", "F")))

# Factor the "Treatment" column 
nonCF_serum_KC_luminex_df <- nonCF_serum_KC_luminex_df %>%
  mutate(Treatment = factor(Treatment, levels = c("Baseline", "PBS", "Bacteroides")))

#serum

#Subset just PBS and Bacteroides treatment groups to run stats (change between the two to set Baseline or PBS as the reference)
nonCF_serum_KC_luminex_df_PBSvsBt <- nonCF_serum_KC_luminex_df[nonCF_serum_KC_luminex_df$Treatment %in% c("Baseline", "PBS"), ]

#check if data is normal
nonCF_serum_KC_luminex_df_PBSvsBt_normalcy_result <- shapiro.test(nonCF_serum_KC_luminex_df_PBSvsBt$Cytokine_pg_per_mL)
print(nonCF_serum_KC_luminex_df_PBSvsBt_normalcy_result) #data is not normal


#Run Stats on KC cytokine after abx to see if treatment or sex group has an affect on KC 


nonCF_serum_KC_luminex_df_PBSvsBt_model2 <- lm(log2(Cytokine_pg_per_mL) ~ Treatment + Sex + Sex*Treatment + Experiment:Cage, data = nonCF_serum_KC_luminex_df_PBSvsBt)
summary(nonCF_serum_KC_luminex_df_PBSvsBt_model2) #this is the one that is used to discuss paper


#lungs
#Subset just PBS and Bacteroides treatment groups to run stats
nonCF_lungs_KC_luminex_df_PBSvsBt <- nonCF_lungs_KC_luminex_df[nonCF_lungs_KC_luminex_df$Treatment %in% c("PBS", "Bacteroides"), ]

# Factor the "Sex" column
nonCF_lungs_KC_luminex_df_PBSvsBt <- nonCF_lungs_KC_luminex_df_PBSvsBt %>%
  mutate(Sex = factor(Sex, levels = c("F", "M")))

#check if data is normal
nonCF_lungs_KC_luminex_df_PBSvsBt_normalcy_result <- shapiro.test(nonCF_lungs_KC_luminex_df_PBSvsBt$Cytokine_pg_per_mL)
print(nonCF_lungs_KC_luminex_df_PBSvsBt_normalcy_result) #data is not normal

#Run Stats on KC cytokine after abx to see if treatment or sex group has an affect on KC in lungs
nonCF_lungs_KC_luminex_df_PBSvsBt_model <- lm(log2(Cytokine_pg_per_mL) ~ Treatment + Sex + Sex*Treatment + Experiment:Cage, data = nonCF_lungs_KC_luminex_df_PBSvsBt)
summary(nonCF_lungs_KC_luminex_df_PBSvsBt_model)



```

Run stats for just KC of CF Exp1 and 2
```{r}
# serum
#Subset just PBS and Bacteroides treatment groups to run stats
CF_serum_KC_luminex_df_PBSvsBt <- CF_serum_KC_luminex_df[CF_serum_KC_luminex_df$Treatment %in% c("PBS", "Bacteroides"), ]

#check if data is normal
CF_serum_KC_luminex_df_PBSvsBt_normalcy_result <- shapiro.test(CF_serum_KC_luminex_df_PBSvsBt$Cytokine_pg_per_mL)
print(CF_serum_KC_luminex_df_PBSvsBt_normalcy_result) #data is normal, but we have only 20 entries so we will still normalize it


#Run Stats on KC cytokine after abx to see if treatment or sex group has an affect on KC / setting Cage as random affect
CF_serum_KC_luminex_df_PBSvsBt_model <- lm(log2(Cytokine_pg_per_mL) ~ Treatment + Sex + Experiment:Cage, data = CF_serum_KC_luminex_df_PBSvsBt)
summary(CF_serum_KC_luminex_df_PBSvsBt_model)

CF_serum_KC_luminex_df_PBSvsBt_model2 <- lmer(log2(Cytokine_pg_per_mL) ~ Treatment + Sex + (1|Cage), data = CF_serum_KC_luminex_df_PBSvsBt)
summary(CF_serum_KC_luminex_df_PBSvsBt_model)

#lungs

#Subset just PBS and Bacteroides treatment groups to run stats
CF_lungs_KC_luminex_df_PBSvsBt <- CF_lungs_KC_luminex_df[CF_lungs_KC_luminex_df$Treatment %in% c("PBS", "Bacteroides"), ]

#check if data is normal
CF_lungs_KC_luminex_df_PBSvsBt_normalcy_result <- shapiro.test(CF_lungs_KC_luminex_df_PBSvsBt$Cytokine_pg_per_mL)
print(CF_lungs_KC_luminex_df_PBSvsBt_normalcy_result) #data is normal, but we have only 20 entries so we will still normalize it


#Run Stats on KC cytokine after abx to see if treatment or sex group has an affect on KC / setting Cage as random affect
CF_lungs_KC_luminex_df_PBSvsBt_model <- lm(log2(Cytokine_pg_per_mL) ~ Treatment + Sex + Experiment:Cage, data = CF_lungs_KC_luminex_df_PBSvsBt)
summary(CF_lungs_KC_luminex_df_PBSvsBt_model)



```

#Run a loop to find all cytokines that are differentially expressed in the serum between PBS and Bacteroides for nonCF mice, across both experiments
```{r}

#Extract just PBS and Bacteroides
nonCF_serum_luminex_PBSandBt_df <- luminex_df %>%
  filter(Genotype == "C57BL_6J",
         SampleType == "Serum",
         Treatment %in% c("PBS", "Bacteroides"))

nonCF_serum_luminex_PBSandBt_df <- nonCF_serum_luminex_PBSandBt_df %>%
  mutate(Cytokine_pg_per_mL = Cytokine_pg_per_mL + 1)

# Check levels of Treatment, Sex, and Experiment:Cage for each Cytokine
cytokine_level_check <- nonCF_serum_luminex_PBSandBt_df %>%
  group_by(Cytokine) %>%
  summarize(
    Treatment_levels = n_distinct(Treatment),
    Sex_levels = n_distinct(Sex),
    Experiment_Cage_levels = n_distinct(interaction(Experiment, Cage))
  )


# Ensure the 'linear_models_serum' directory exists
dir.create("linear_models_nonCF_serum", showWarnings = FALSE)

# Initialize the summary dataframe
linear_models_nonCFserum_summary <- data.frame(
  Cytokine = character(),
  P_Value_Treatment = numeric(),
  P_Value_Sex = numeric(),
  stringsAsFactors = FALSE
)

# Function to safely extract p-value
safe_extract_pvalue <- function(model_summary, term) {
  coef_table <- as.data.frame(model_summary$coefficients)
  if (term %in% rownames(coef_table)) {
    return(coef_table[term, "Pr(>|t|)"])
  }
  return(NA)
}

# Loop over each unique cytokine
for(cytokine in unique(nonCF_serum_luminex_PBSandBt_df$Cytokine)) {
  # Subset data for the current cytokine
  subset_data <- subset(nonCF_serum_luminex_PBSandBt_df, Cytokine == cytokine)

  # Fit the mixed linear model
  model <- lm(log2(Cytokine_pg_per_mL) ~ Treatment + Sex + Experiment:Cage, data = subset_data)

  # Obtain the summary of the model
  model_summary <- summary(model)

  # Extract the p-values for Genotype and Sex
  p_treatment <- safe_extract_pvalue(model_summary, "TreatmentBacteroides")
  p_sex <- safe_extract_pvalue(model_summary, "SexM")

  # Add to the summary dataframe
  linear_models_nonCFserum_summary <- rbind(linear_models_nonCFserum_summary, data.frame(
    Cytokine = cytokine,
    P_Value_Treatment = p_treatment,
    P_Value_Sex = p_sex
  ))

  # Save the summary of each model to a text file
  file_name <- paste0("linear_models_nonCF_serum/model_", cytokine, ".txt")
  capture.output(model_summary, file = file_name)
}

# View the summary dataframe
print(linear_models_nonCFserum_summary)

# Function to categorize significance
categorize_significance <- function(p_value) {
  if(p_value < 0.001) {
    return("***")
  } else if(p_value < 0.01) {
    return("**")
  } else if(p_value < 0.05) {
    return("*")
  } else {
    return("NS")
  }
}

# Apply the function to the dataframe
linear_models_nonCFserum_summary$Significance_Treatment <- sapply(linear_models_nonCFserum_summary$P_Value_Treatment, categorize_significance)
linear_models_nonCFserum_summary$Significance_Sex <- sapply(linear_models_nonCFserum_summary$P_Value_Sex, categorize_significance)

# View the updated dataframe
linear_models_nonCFserum_summary

# Apply Bonferroni correction to the P_Value_Genotype column
linear_models_nonCFserum_summary$P_Value_Treatment_Bonferroni <- p.adjust(linear_models_nonCFserum_summary$P_Value_Treatment, method = "bonferroni")
linear_models_nonCFserum_summary$P_Value_Sex_Bonferroni <- p.adjust(linear_models_nonCFserum_summary$P_Value_Sex, method = "bonferroni")



# Save the dataframe as a CSV file
write.csv(linear_models_nonCFserum_summary, "linear_models_nonCFserum_summary.csv", row.names = FALSE)



```


#Run a loop to find all cytokines that are differentially expressed in the lungs between PBS and Bacteroides for nonCF mice, across both experiments

```{r}

#Extract just PBS and Bacteroides
nonCF_lungs_luminex_PBSandBt_df <- luminex_df %>%
  filter(Genotype == "C57BL_6J",
         SampleType == "Lungs",
         Treatment %in% c("PBS", "Bacteroides"))

nonCF_lungs_luminex_PBSandBt_df <- nonCF_lungs_luminex_PBSandBt_df %>%
  mutate(Cytokine_pg_per_mL = Cytokine_pg_per_mL + 1)

# Check levels of Treatment, Sex, and Experiment:Cage for each Cytokine
cytokine_level_check2 <- nonCF_lungs_luminex_PBSandBt_df %>%
  group_by(Cytokine) %>%
  summarize(
    Treatment_levels = n_distinct(Treatment),
    Sex_levels = n_distinct(Sex),
    Experiment_Cage_levels = n_distinct(interaction(Experiment, Cage))
  )


# Ensure the 'linear_models_serum' directory exists
dir.create("linear_models_nonCF_lungs", showWarnings = FALSE)

# Initialize the summary dataframe
linear_models_nonCFlungs_summary <- data.frame(
  Cytokine = character(),
  P_Value_Treatment = numeric(),
  P_Value_Sex = numeric(),
  stringsAsFactors = FALSE
)

# Function to safely extract p-value
safe_extract_pvalue <- function(model_summary2, term) {
  coef_table <- as.data.frame(model_summary2$coefficients)
  if (term %in% rownames(coef_table)) {
    return(coef_table[term, "Pr(>|t|)"])
  }
  return(NA)
}

# Loop over each unique cytokine
for(cytokine in unique(nonCF_lungs_luminex_PBSandBt_df$Cytokine)) {
  # Subset data for the current cytokine
  subset_data2 <- subset(nonCF_lungs_luminex_PBSandBt_df, Cytokine == cytokine)

  # Fit the mixed linear model
  model2 <- lm(log2(Cytokine_pg_per_mL) ~ Treatment + Sex + Experiment:Cage, data = subset_data2)

  # Obtain the summary of the model
  model_summary2 <- summary(model2)

  # Extract the p-values for Genotype and Sex
  p_treatment <- safe_extract_pvalue(model_summary2, "TreatmentBacteroides")
  p_sex <- safe_extract_pvalue(model_summary2, "SexM")

  # Add to the summary dataframe
  linear_models_nonCFlungs_summary <- rbind(linear_models_nonCFlungs_summary, data.frame(
    Cytokine = cytokine,
    P_Value_Treatment = p_treatment,
    P_Value_Sex = p_sex
  ))

  # Save the summary of each model to a text file
  file_name <- paste0("linear_models_nonCF_lungs/model_", cytokine, ".txt")
  capture.output(model_summary2, file = file_name)
}

# View the summary dataframe
print(linear_models_nonCFlungs_summary)

# Function to categorize significance
categorize_significance <- function(p_value) {
  if(p_value < 0.001) {
    return("***")
  } else if(p_value < 0.01) {
    return("**")
  } else if(p_value < 0.05) {
    return("*")
  } else {
    return("NS")
  }
}

# Apply the function to the dataframe
linear_models_nonCFlungs_summary$Significance_Treatment <- sapply(linear_models_nonCFlungs_summary$P_Value_Treatment, categorize_significance)
linear_models_nonCFlungs_summary$Significance_Sex <- sapply(linear_models_nonCFlungs_summary$P_Value_Sex, categorize_significance)

# View the updated dataframe
linear_models_nonCFlungs_summary

# Apply Bonferroni correction to the P_Value_Genotype column
linear_models_nonCFlungs_summary$P_Value_Treatment_Bonferroni <- p.adjust(linear_models_nonCFlungs_summary$P_Value_Treatment, method = "bonferroni")



# Save the dataframe as a CSV file
write.csv(linear_models_nonCFlungs_summary, "linear_models_nonCFlungs_summary.csv", row.names = FALSE)




```

#Make a volcano plot showing fold change on x axis and p valye on y axis for serum cytokines
```{r}

# Filter data for nonCF and CF
nonCF_PBS_data_serum <- nonCF_serum_luminex_PBSandBt_df[nonCF_serum_luminex_PBSandBt_df$Treatment == 'PBS', ]
nonCF_Bt_data_serum <- nonCF_serum_luminex_PBSandBt_df[nonCF_serum_luminex_PBSandBt_df$Treatment == 'Bacteroides', ]

# Calculate median values for each cytokine in both nonCF and CF groups
median_nonCF_PBS_serum <- aggregate(Cytokine_pg_per_mL ~ Cytokine, nonCF_PBS_data_serum, median)
median_nonCF_Bt_serum <- aggregate(Cytokine_pg_per_mL ~ Cytokine, nonCF_Bt_data_serum, median)


# Merge the two dataframes on cytokine names
foldchange_nonCF_serum <- merge(median_nonCF_PBS_serum, median_nonCF_Bt_serum, by = "Cytokine")

# Calculate fold changes
foldchange_nonCF_serum$FoldChange <- foldchange_nonCF_serum$Cytokine_pg_per_mL.y / foldchange_nonCF_serum$Cytokine_pg_per_mL.x

# Rename columns for clarity
names(foldchange_nonCF_serum) <- c("Cytokine", "Median_PBS", "Median_Bacteroides", "FoldChange")

# Display the dataframe
print(foldchange_nonCF_serum)

# Merge the 'P_Value_Genotype' column from 'linear_models_serum_summary' into 'foldchange_serum'
foldchange_nonCF_serum <- merge(foldchange_nonCF_serum, linear_models_nonCFserum_summary[, c("Cytokine", "P_Value_Treatment")], by = "Cytokine", all.x = TRUE)



```

#Draw a volcano plot on serums
```{r}
# Transform the data
foldchange_nonCF_serum$log2FoldChange <- log2(foldchange_nonCF_serum$FoldChange)
foldchange_nonCF_serum$negLog10PValue <- -log10(foldchange_nonCF_serum$P_Value_Treatment)

# Add a new column "DiffExpression" to the foldchange_serum DataFrame
foldchange_nonCF_serum$DiffExpression <- ifelse(foldchange_nonCF_serum$P_Value_Treatment < 0.05, 
                                          ifelse(foldchange_nonCF_serum$FoldChange > 1, "Up in Bacteroides", 
                                                 ifelse(foldchange_nonCF_serum$FoldChange < 1, "Down in Bacteroides", "Not Significant")),
                                          "Not Significant")

# Biostatsquid theme
theme_set(theme_classic(base_size = 20) +
            theme(
              axis.title.y = element_text(face = "bold", margin = margin(0,20,0,0), size = rel(0.5), color = 'black'),
              axis.title.x = element_text(hjust = 0.5, face = "bold", margin = margin(20,0,0,0), size = rel(0.5), color = 'black'),
              plot.title = element_text(hjust = 0.5)
            ))


# Create a basic volcano plot
Volcanoplot_PBSvsBacteroides_nonCFserum<- ggplot(data = foldchange_nonCF_serum, aes(x = log2FoldChange, y = negLog10PValue, color = DiffExpression)) +
         geom_point()+
  geom_vline(xintercept = c(-0.6, 0.6), col = "gray", linetype = 'dashed') +
  geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed') +
  scale_color_manual(values = c("Up in Bacteroides" = "#00AFBB", "Down in Bacteroides" = "#bb0c00", "Not Significant" = "grey"))+
  geom_text(aes(label=Cytokine), vjust = 2, hjust = 0.5, size = 2, check_overlap = TRUE)+
  ggtitle("Cytokines Differentially Expressed Between PBS and Bacteroides gavaged nonCF Mice in Serum")+
  theme(plot.title = element_text(size = 12)) # Reduced font size for title


# Save the plot as a PDF
ggsave("Volcanoplot_PBSvsBacteroides_nonCFserum.pdf", plot = Volcanoplot_PBSvsBacteroides_nonCFserum, width = 10, height = 4, device = "pdf")


```
---
title: "08062024_LuminexAnalysis"
author: "Rebecca Valls"
date: "2024-08-06"
output: html_document
---
'
#Install libraries
```{r}
# Install and load necessary packages if not already installed
# install.packages("dplyr")
# install.packages("readr")
# install.packages("broom")
# install.packages("purrr")
# install.packages("hrbrthemes")
# install.packages("geomtextpath")
# install.packages("lmerTest")
# install.packages("ggplot2")
library(readr)
library(dplyr)
library(broom)
library(purrr)
library(ggplot2)
library(lme4)
library(lmerTest)
library(hrbrthemes)
library(geomtextpath)
library(dplyr)


```

#Load data
```{r}
BacteroidesEstablishment_SerumandLungs_2nonCFand2CF_MasterLuminex <- read_csv("~/Dartmouth College Dropbox/Rebecca Valls/Projects/Project_MalevsFemale_nonCFmice_2023/3_Experiment/3_DataAnalysis/Luminex_Aug2024/BacteroidesEstablishment_SerumandLungs__2nonCFand2CF_MasterLuminex.csv")

luminex_df <- as.data.frame(BacteroidesEstablishment_SerumandLungs_2nonCFand2CF_MasterLuminex)

# Factor the "Treatment" column
luminex_df <- luminex_df %>%
  mutate(Treatment = factor(Treatment, levels = c("Baseline", "PBS", "Bacteroides")))

# Factor the "Sex" column
luminex_df <- luminex_df %>%
  mutate(Sex = factor(Sex, levels = c("M", "F")))


luminex_df$Cage <- as.factor(luminex_df$Cage)




#Subset data into each experiment
#nonCF Exp1
nonCF_Exp1_luminex_df<- luminex_df %>%
  filter(Experiment == "nonCF_Exp1")

#nonCF Exp2
nonCF_Exp2_luminex_df<- luminex_df %>%
  filter(Experiment == "nonCF_Exp2")


#CF Exp1
CF_Exp1_luminex_df<- luminex_df %>%
  filter(Experiment %in% c("CF_Baseline", "CF_Exp1"))

#CF Exp2
CF_Exp2_luminex_df<- luminex_df %>%
  filter(Experiment == "CF_Exp2")


```


#Extract serum and lungs KC and plot for nonCF Experiment 1
```{r}
nonCF_Exp1_serum_KC_luminex_df <- nonCF_Exp1_luminex_df %>%
  filter(SampleType == "Serum",
         Cytokine == "KC")


nonCF_Exp1_serum_KC_dotplot <- ggplot(nonCF_Exp1_serum_KC_luminex_df, aes(x = Treatment, y = Cytokine_pg_per_mL, color = Treatment, shape = Sex)) +
  geom_jitter(width = 0.2, size = 3) +  # Adjust width to avoid overlap
    stat_summary(fun = median, geom = "crossbar", aes(group = Treatment), width = 0.5, fatten = 2) +  # Add median crossbars
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate labels at 45 degrees
  labs(y = "KC in Serum (pg/mL)", color = "Treatment Group")

nonCF_Exp1_serum_KC_dotplot

ggsave("nonCF_Exp1_serum_KC_dotplot.pdf", plot = nonCF_Exp1_serum_KC_dotplot, device = "pdf", width = 4, height = 3)


nonCF_Exp1_lungs_KC_luminex_df <- nonCF_Exp1_luminex_df %>%
  filter(SampleType == "Lungs",
         Cytokine == "KC")


nonCF_Exp1_lungs_KC_dotplot <- ggplot(nonCF_Exp1_lungs_KC_luminex_df, aes(x = Treatment, y = Cytokine_pg_per_mL, color = Treatment, shape = Sex)) +
  geom_jitter(width = 0.2, size = 3) +  # Adjust width to avoid overlap
    stat_summary(fun = median, geom = "crossbar", aes(group = Treatment), width = 0.5, fatten = 2) +  # Add median crossbars
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate labels at 45 degrees
  labs(y = "KC in Lungs (pg/mL)", color = "Treatment Group")

nonCF_Exp1_lungs_KC_dotplot

ggsave("nonCF_Exp1_lungs_KC_dotplot.pdf", plot = nonCF_Exp1_lungs_KC_dotplot, device = "pdf", width = 4, height = 3)




```

#Extract serum and lungs KC and plot for Experiment 2
```{r}
nonCF_Exp2_serum_KC_luminex_df <- nonCF_Exp2_luminex_df %>%
  filter(SampleType == "Serum",
         Cytokine == "KC")


nonCF_Exp2_serum_KC_dotplot <- ggplot(nonCF_Exp2_serum_KC_luminex_df, aes(x = Treatment, y = Cytokine_pg_per_mL, color = Treatment, shape = Sex)) +
  geom_jitter(width = 0.2, size = 3) +  # Adjust width to avoid overlap
    stat_summary(fun = median, geom = "crossbar", aes(group = Treatment), width = 0.5, fatten = 2) +  # Add median crossbars
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate labels at 45 degrees
  labs(y = "KC in Serum (pg/mL)", color = "Treatment Group")

nonCF_Exp2_serum_KC_dotplot

ggsave("nonCF_Exp2_serum_KC_dotplot.pdf", plot = nonCF_Exp2_serum_KC_dotplot, device = "pdf", width = 4, height = 3)


nonCF_Exp2_lungs_KC_luminex_df <- nonCF_Exp2_luminex_df %>%
  filter(SampleType == "Lungs",
         Cytokine == "KC")


nonCF_Exp2_lungs_KC_dotplot <- ggplot(nonCF_Exp2_lungs_KC_luminex_df, aes(x = Treatment, y = Cytokine_pg_per_mL, color = Treatment, shape = Sex)) +
  geom_jitter(width = 0.2, size = 3) +  # Adjust width to avoid overlap
    stat_summary(fun = median, geom = "crossbar", aes(group = Treatment), width = 0.5, fatten = 2) +  # Add median crossbars
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate labels at 45 degrees
  labs(y = "KC in Lungs (pg/mL)", color = "Treatment Group")

nonCF_Exp2_lungs_KC_dotplot
ggsave("nonCF_Exp2_lungs_KC_dotplot.pdf", plot = nonCF_Exp2_lungs_KC_dotplot, device = "pdf", width = 4, height = 3)


```
#Extract serum and lungs KC and plot for CF Experiment 1
```{r}

CF_Exp1_serum_KC_luminex_df <- CF_Exp1_luminex_df %>%
  filter(SampleType == "Serum",
         Cytokine == "KC")


CF_Exp1_serum_KC_dotplot <- ggplot(CF_Exp1_serum_KC_luminex_df, aes(x = Treatment, y = Cytokine_pg_per_mL, color = Treatment, shape = Sex)) +
  geom_jitter(width = 0.2, size = 3) +  # Adjust width to avoid overlap
    stat_summary(fun = median, geom = "crossbar", aes(group = Treatment), width = 0.5, fatten = 2) +  # Add median crossbars
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate labels at 45 degrees
  labs(y = "KC in Serum (pg/mL)", color = "Treatment Group")

CF_Exp1_serum_KC_dotplot

ggsave("CF_Exp1_serum_KC_dotplot.pdf", plot = CF_Exp1_serum_KC_dotplot, device = "pdf", width = 4, height = 3)


CF_Exp1_Lungs_KC_luminex_df <- CF_Exp1_luminex_df %>%
  filter(SampleType == "Lungs",
         Cytokine == "KC")


CF_Exp1_lungs_KC_dotplot <- ggplot(CF_Exp1_Lungs_KC_luminex_df, aes(x = Treatment, y = Cytokine_pg_per_mL, color = Treatment, shape = Sex)) +
  geom_jitter(width = 0.2, size = 3) +  # Adjust width to avoid overlap
    stat_summary(fun = median, geom = "crossbar", aes(group = Treatment), width = 0.5, fatten = 2) +  # Add median crossbars
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate labels at 45 degrees
  labs(y = "KC in Lungs (pg/mL)", color = "Treatment Group")

CF_Exp1_lungs_KC_dotplot

ggsave("CF_Exp1_lungs_KC_dotplot.pdf", plot = CF_Exp1_lungs_KC_dotplot, device = "pdf", width = 4, height = 3)



```

#Extract serum and lungs KC and plot for CF Experiment 2
```{r}

CF_Exp2_serum_KC_luminex_df <- CF_Exp2_luminex_df %>%
  filter(SampleType == "Serum",
         Cytokine == "KC")


CF_Exp2_serum_KC_dotplot <- ggplot(CF_Exp2_serum_KC_luminex_df, aes(x = Treatment, y = Cytokine_pg_per_mL, color = Treatment, shape = Sex)) +
  geom_jitter(width = 0.2, size = 3) +  # Adjust width to avoid overlap
    stat_summary(fun = median, geom = "crossbar", aes(group = Treatment), width = 0.5, fatten = 2) +  # Add median crossbars
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate labels at 45 degrees
  labs(y = "KC in Serum (pg/mL)", color = "Treatment Group")

CF_Exp2_serum_KC_dotplot
ggsave("CF_Exp2_serum_KC_dotplot.pdf", plot = CF_Exp2_serum_KC_dotplot, device = "pdf", width = 4, height = 3)


CF_Exp2_Lungs_KC_luminex_df <- CF_Exp2_luminex_df %>%
  filter(SampleType == "Lungs",
         Cytokine == "KC")


CF_Exp2_lungs_KC_dotplot <- ggplot(CF_Exp2_Lungs_KC_luminex_df, aes(x = Treatment, y = Cytokine_pg_per_mL, color = Treatment, shape = Sex)) +
  geom_jitter(width = 0.2, size = 3) +  # Adjust width to avoid overlap
    stat_summary(fun = median, geom = "crossbar", aes(group = Treatment), width = 0.5, fatten = 2) +  # Add median crossbars
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate labels at 45 degrees
  labs(y = "KC in Lungs (pg/mL)", color = "Treatment Group")

CF_Exp2_lungs_KC_dotplot
ggsave("CF_Exp2_lungs_KC_dotplot.pdf", plot = CF_Exp2_lungs_KC_dotplot, device = "pdf", width = 4, height = 3)


```


#Plot KC in serum and lungs from both nonCF experiments
```{r}
nonCF_serum_KC_luminex_df <- luminex_df %>%
  filter(Genotype == "C57BL_6J",
         SampleType == "Serum",
         Cytokine == "KC")

nonCF_serum_KC_dotplot <- ggplot(nonCF_serum_KC_luminex_df, aes(x = Treatment, y = Cytokine_pg_per_mL, color = Treatment, shape = Sex)) +
  geom_jitter(width = 0.2, size = 1.5) +  # Adjust width to avoid overlap
    stat_summary(fun = median, geom = "crossbar", aes(group = Treatment), width = 0.8, fatten = 5) +  # Add median crossbars
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  scale_shape_manual(values = c("F" = 16, "M" = 17)) +  # Set female to circles and male to triangles
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate labels at 45 degrees
  labs(y = "KC in Serum (pg/mL)", color = "Treatment Group")

nonCF_serum_KC_dotplot
ggsave("nonCF_serum_KC_dotplot.pdf", plot = nonCF_serum_KC_dotplot, device = "pdf", width = 6, height = 3)

#help from claude to outline the male triangles with black line: 
print(colnames(nonCF_serum_KC_luminex_df))

nonCF_serum_KC_dotplot_claude <- ggplot(nonCF_serum_KC_luminex_df, 
                                  aes(x = Treatment, y = Cytokine_pg_per_mL_serum, fill = Treatment)) +
  geom_jitter(data = subset(nonCF_serum_KC_luminex_df, Sex == "F"),
              aes(shape = Sex, fill = Treatment), color = "grey50", width = 0.2, size = 1.5, stroke = 0) +
  geom_jitter(data = subset(nonCF_serum_KC_luminex_df, Sex == "M"),
              aes(shape = Sex, fill = Treatment), color = "black", width = 0.2, size = 1.5, stroke = 0.5) +
  stat_summary(aes(group = Treatment, color = Treatment), fun = median, geom = "crossbar", width = 0.8, fatten = 2) +
  scale_shape_manual(values = c("F" = 21, "M" = 24)) +
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange"), guide = "none") +
  scale_fill_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange"), name = "Treatment") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(y = "KC in Serum (pg/mL)")

nonCF_serum_KC_dotplot_claude
ggsave("nonCF_serum_KC_dotplot_PR.pdf", plot = nonCF_serum_KC_dotplot_claude, device = "pdf", width = 6, height = 4)


#separate by sex
nonCF_serum_KC_dotplot_sexseparated <- ggplot(nonCF_serum_KC_luminex_df, aes(x = Treatment, y = Cytokine_pg_per_mL, color = Treatment, shape = Sex)) +
  facet_wrap(~Sex)+
  geom_jitter(width = 0.2, size = 1.5) +  # Adjust width to avoid overlap
    stat_summary(fun = median, geom = "crossbar", aes(group = Treatment), width = 0.8, fatten = 5) +  # Add median crossbars
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  scale_shape_manual(values = c("F" = 16, "M" = 17)) +  # Set female to circles and male to triangles
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate labels at 45 degrees
  labs(y = "KC in Serum (pg/mL)", color = "Treatment Group")

nonCF_serum_KC_dotplot_sexseparated
ggsave("nonCF_serum_KC_dotplot_sexseparated.pdf", plot = nonCF_serum_KC_dotplot_sexseparated, device = "pdf", width = 6, height = 3)

nonCF_lungs_KC_luminex_df <- luminex_df %>%
  filter(Genotype == "C57BL_6J",
         SampleType == "Lungs",
         Cytokine == "KC")

nonCF_lungs_KC_dotplot <- ggplot(nonCF_lungs_KC_luminex_df, aes(x = Treatment, y = Cytokine_pg_per_mL, color = Treatment, shape = Sex)) +
  geom_jitter(width = 0.2, size = 1.5) +  # Adjust width to avoid overlap
    stat_summary(fun = median, geom = "crossbar", aes(group = Treatment), width = 0.8, fatten = 2) +  # Add median crossbars
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  scale_shape_manual(values = c("F" = 16, "M" = 17)) +  # Set female to circles and male to triangles
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate labels at 45 degrees
  labs(y = "KC in Lungs (pg/mL)", color = "Treatment Group")


nonCF_lungs_KC_dotplot

ggsave("nonCF_lungs_KC_dotplot.pdf", plot = nonCF_lungs_KC_dotplot, device = "pdf", width = 6, height = 3)

#help from claude to outline the male triangles with black line: 
nonCF_lungs_KC_dotplot_claude <- ggplot(nonCF_lungs_KC_luminex_df, 
                                  aes(x = Treatment, y = Cytokine_pg_per_mL_lungs, fill = Treatment)) +
  geom_jitter(data = subset(nonCF_lungs_KC_luminex_df, Sex == "F"),
              aes(shape = Sex, fill = Treatment), color = "grey50", width = 0.2, size = 1.5, stroke = 0) +
  geom_jitter(data = subset(nonCF_lungs_KC_luminex_df, Sex == "M"),
              aes(shape = Sex, fill = Treatment), color = "black", width = 0.2, size = 1.5, stroke = 0.5) +
  stat_summary(aes(group = Treatment, color = Treatment), fun = median, geom = "crossbar", width = 0.8, fatten = 2) +
  scale_shape_manual(values = c("F" = 21, "M" = 24)) +
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange"), guide = "none") +
  scale_fill_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange"), name = "Treatment") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(y = "KC in Lungs (pg/mL)")

nonCF_lungs_KC_dotplot_claude
ggsave("nonCF_lungs_KC_dotplot_PR.pdf", plot = nonCF_lungs_KC_dotplot_claude, device = "pdf", width = 6, height = 4)


nonCF_lungs_KC_dotplot_sexseparated <- ggplot(nonCF_lungs_KC_luminex_df, aes(x = Treatment, y = Cytokine_pg_per_mL, color = Treatment, shape = Sex)) +
  facet_wrap(~Sex)+
  geom_jitter(width = 0.2, size = 1.5) +  # Adjust width to avoid overlap
    stat_summary(fun = median, geom = "crossbar", aes(group = Treatment), width = 0.8, fatten = 5) +  # Add median crossbars
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  scale_shape_manual(values = c("F" = 16, "M" = 17)) +  # Set female to circles and male to triangles
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate labels at 45 degrees
  labs(y = "KC in Lungs (pg/mL)", color = "Treatment Group")


nonCF_lungs_KC_dotplot_sexseparated

ggsave("nonCF_lungs_KC_dotplot_sexseparated.pdf", plot = nonCF_lungs_KC_dotplot_sexseparated, device = "pdf", width = 6, height = 3)

```

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
# Compute KC mean differences (PBS - Bacteroides) for non-CF mice
kc_diff_plot <- luminex_df %>%
  filter(Genotype == "C57BL_6J", Cytokine == "KC", SampleType %in% c("Lungs", "Serum"), Treatment %in% c("PBS", "Bacteroides")) %>%
  group_by(SampleType, Sex, Treatment) %>%
  summarise(median_KC = median(Cytokine_pg_per_mL), .groups = "drop") %>%
  pivot_wider(names_from = Treatment, values_from = median_KC) %>%
  mutate(Diff = PBS - Bacteroides) %>%
  ggplot(aes(x = interaction(SampleType, Sex), y = -Diff, fill = SampleType)) +
  geom_bar(stat = "identity", width = 0.6) +
  labs(
    x = "Sample Type and Sex",
    y = "Median Change in KC (PBS - Bacteroides)",
    title = "Median Change in KC by Sample Type and Sex (non-CF)"
  ) +
  scale_fill_manual(values = c("Serum" = "skyblue", "Lungs" = "tomato")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Print and save plot
kc_diff_plot
ggsave("nonCF_KC_median_diff_by_sampletype_and_sex.pdf", plot = kc_diff_plot, width = 5, height = 4)

```






#Plot KC in serum and lungs from both experiments
```{r}
CF_serum_KC_luminex_df <- luminex_df %>%
  filter(Genotype == "CftrF508del",
         SampleType == "Serum",
         Cytokine == "KC")

CF_serum_KC_dotplot <- ggplot(CF_serum_KC_luminex_df, aes(x = Treatment, y = Cytokine_pg_per_mL, color = Treatment, shape = Sex)) +
  geom_jitter(width = 0.2, size = 3) +  # Adjust width to avoid overlap
    stat_summary(fun = median, geom = "crossbar", aes(group = Treatment), width = 0.5, fatten = 2) +  # Add median crossbars
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate labels at 45 degrees
  labs(y = "KC in Serum (pg/mL)", color = "Treatment Group")

CF_serum_KC_dotplot
ggsave("CF_serum_KC_dotplot.pdf", plot = CF_serum_KC_dotplot, device = "pdf", width = 4, height = 3)

CF_lungs_KC_luminex_df <- luminex_df %>%
  filter(Genotype == "CftrF508del",
         SampleType == "Lungs",
         Cytokine == "KC")

CF_lungs_KC_dotplot <- ggplot(CF_lungs_KC_luminex_df, aes(x = Treatment, y = Cytokine_pg_per_mL, color = Treatment, shape = Sex)) +
  geom_jitter(width = 0.2, size = 3) +  # Adjust width to avoid overlap
    stat_summary(fun = median, geom = "crossbar", aes(group = Treatment), width = 0.5, fatten = 2) +  # Add median crossbars
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate labels at 45 degrees
  labs(y = "KC in Lungs (pg/mL)", color = "Treatment Group")

CF_lungs_KC_dotplot
ggsave("CF_lungs_KC_dotplot.pdf", plot = CF_lungs_KC_dotplot, device = "pdf", width = 4, height = 3)

```



Run stats for just KC of nonCF Exp1 and 2
```{r}
# Factor the "Sex" column
nonCF_serum_KC_luminex_df <- nonCF_serum_KC_luminex_df %>%
  mutate(Sex = factor(Sex, levels = c("M", "F")))

# Factor the "Treatment" column 
nonCF_serum_KC_luminex_df <- nonCF_serum_KC_luminex_df %>%
  mutate(Treatment = factor(Treatment, levels = c("Baseline", "PBS", "Bacteroides")))

#serum

#Subset just PBS and Bacteroides treatment groups to run stats (change between the two to set Baseline or PBS as the reference)
nonCF_serum_KC_luminex_df_PBSvsBt <- nonCF_serum_KC_luminex_df[nonCF_serum_KC_luminex_df$Treatment %in% c("Baseline", "PBS"), ]

#check if data is normal
nonCF_serum_KC_luminex_df_PBSvsBt_normalcy_result <- shapiro.test(nonCF_serum_KC_luminex_df_PBSvsBt$Cytokine_pg_per_mL)
print(nonCF_serum_KC_luminex_df_PBSvsBt_normalcy_result) #data is not normal


#Run Stats on KC cytokine after abx to see if treatment or sex group has an affect on KC 


nonCF_serum_KC_luminex_df_PBSvsBt_model2 <- lm(log2(Cytokine_pg_per_mL) ~ Treatment + Sex + Sex*Treatment + Experiment:Cage, data = nonCF_serum_KC_luminex_df_PBSvsBt)
summary(nonCF_serum_KC_luminex_df_PBSvsBt_model2) #this is the one that is used to discuss paper


#lungs
#Subset just PBS and Bacteroides treatment groups to run stats
nonCF_lungs_KC_luminex_df_PBSvsBt <- nonCF_lungs_KC_luminex_df[nonCF_lungs_KC_luminex_df$Treatment %in% c("PBS", "Bacteroides"), ]

# Factor the "Sex" column
nonCF_lungs_KC_luminex_df_PBSvsBt <- nonCF_lungs_KC_luminex_df_PBSvsBt %>%
  mutate(Sex = factor(Sex, levels = c("F", "M")))

#check if data is normal
nonCF_lungs_KC_luminex_df_PBSvsBt_normalcy_result <- shapiro.test(nonCF_lungs_KC_luminex_df_PBSvsBt$Cytokine_pg_per_mL)
print(nonCF_lungs_KC_luminex_df_PBSvsBt_normalcy_result) #data is not normal

#Run Stats on KC cytokine after abx to see if treatment or sex group has an affect on KC in lungs
nonCF_lungs_KC_luminex_df_PBSvsBt_model <- lm(log2(Cytokine_pg_per_mL) ~ Treatment + Sex + Sex*Treatment + Experiment:Cage, data = nonCF_lungs_KC_luminex_df_PBSvsBt)
summary(nonCF_lungs_KC_luminex_df_PBSvsBt_model)

#Compare baseline to Bacteroides
nonCF_lungs_KC_luminex_df_BaselinevsBt <- nonCF_lungs_KC_luminex_df[nonCF_lungs_KC_luminex_df$Treatment %in% c("Baseline", "Bacteroides"), ]

# Factor the "Sex" column
nonCF_lungs_KC_luminex_df_BaselinevsBt <- nonCF_lungs_KC_luminex_df_BaselinevsBt %>%
  mutate(Sex = factor(Sex, levels = c("F", "M")))

#check if data is normal
nonCF_lungs_KC_luminex_df_BaselinevsBt_normalcy_result <- shapiro.test(nonCF_lungs_KC_luminex_df_BaselinevsBt$Cytokine_pg_per_mL)

print(nonCF_lungs_KC_luminex_df_BaselinevsBt_normalcy_result) #data is not normal

#Run Stats on KC cytokine after abx to see if treatment or sex group has an affect on KC in lungs
nonCF_lungs_KC_luminex_df_BaselinevsBt_model <- lm(log2(Cytokine_pg_per_mL) ~ Treatment + Sex + Sex*Treatment + Experiment:Cage, data = nonCF_lungs_KC_luminex_df_BaselinevsBt)
summary(nonCF_lungs_KC_luminex_df_BaselinevsBt_model)


#Compare baseline to PBS
nonCF_lungs_KC_luminex_df_BaselinevsPBS <- nonCF_lungs_KC_luminex_df[nonCF_lungs_KC_luminex_df$Treatment %in% c("Baseline", "PBS"), ]

# Factor the "Sex" column
nonCF_lungs_KC_luminex_df_BaselinevsPBS <- nonCF_lungs_KC_luminex_df_BaselinevsPBS %>%
  mutate(Sex = factor(Sex, levels = c("F", "M")))

#check if data is normal
nonCF_lungs_KC_luminex_df_BaselinevsPBS_normalcy_result <- shapiro.test(nonCF_lungs_KC_luminex_df_BaselinevsPBS$Cytokine_pg_per_mL)

print(nonCF_lungs_KC_luminex_df_BaselinevsPBS_normalcy_result) #data is not normal

#Run Stats on KC cytokine after abx to see if treatment or sex group has an affect on KC in lungs
nonCF_lungs_KC_luminex_df_BaselinevsPBS_model <- lm(log2(Cytokine_pg_per_mL) ~ Treatment + Sex + Sex*Treatment + Experiment:Cage, data = nonCF_lungs_KC_luminex_df_BaselinevsPBS)
summary(nonCF_lungs_KC_luminex_df_BaselinevsPBS_model)


```

Run stats for just KC of CF Exp1 and 2
```{r}
# serum
#Subset just PBS and Bacteroides treatment groups to run stats
CF_serum_KC_luminex_df_PBSvsBt <- CF_serum_KC_luminex_df[CF_serum_KC_luminex_df$Treatment %in% c("PBS", "Bacteroides"), ]

#check if data is normal
CF_serum_KC_luminex_df_PBSvsBt_normalcy_result <- shapiro.test(CF_serum_KC_luminex_df_PBSvsBt$Cytokine_pg_per_mL)
print(CF_serum_KC_luminex_df_PBSvsBt_normalcy_result) #data is normal, but we have only 20 entries so we will still normalize it


#Run Stats on KC cytokine after abx to see if treatment or sex group has an affect on KC / setting Cage as random affect
CF_serum_KC_luminex_df_PBSvsBt_model <- lm(log2(Cytokine_pg_per_mL) ~ Treatment + Sex + Experiment:Cage, data = CF_serum_KC_luminex_df_PBSvsBt)
summary(CF_serum_KC_luminex_df_PBSvsBt_model)

CF_serum_KC_luminex_df_PBSvsBt_model2 <- lmer(log2(Cytokine_pg_per_mL) ~ Treatment + Sex + (1|Cage), data = CF_serum_KC_luminex_df_PBSvsBt)
summary(CF_serum_KC_luminex_df_PBSvsBt_model)

#lungs

#Subset just PBS and Bacteroides treatment groups to run stats
CF_lungs_KC_luminex_df_PBSvsBt <- CF_lungs_KC_luminex_df[CF_lungs_KC_luminex_df$Treatment %in% c("PBS", "Bacteroides"), ]

#check if data is normal
CF_lungs_KC_luminex_df_PBSvsBt_normalcy_result <- shapiro.test(CF_lungs_KC_luminex_df_PBSvsBt$Cytokine_pg_per_mL)
print(CF_lungs_KC_luminex_df_PBSvsBt_normalcy_result) #data is normal, but we have only 20 entries so we will still normalize it


#Run Stats on KC cytokine after abx to see if treatment or sex group has an affect on KC / setting Cage as random affect
CF_lungs_KC_luminex_df_PBSvsBt_model <- lm(log2(Cytokine_pg_per_mL) ~ Treatment + Sex + Experiment:Cage, data = CF_lungs_KC_luminex_df_PBSvsBt)
summary(CF_lungs_KC_luminex_df_PBSvsBt_model)



```

#Run a loop to find all cytokines that are differentially expressed in the serum between PBS and Bacteroides for nonCF mice, across both experiments
```{r}

#Extract just PBS and Bacteroides
nonCF_serum_luminex_PBSandBt_df <- luminex_df %>%
  filter(Genotype == "C57BL_6J",
         SampleType == "Serum",
         Treatment %in% c("PBS", "Bacteroides"))

nonCF_serum_luminex_PBSandBt_df <- nonCF_serum_luminex_PBSandBt_df %>%
  mutate(Cytokine_pg_per_mL = Cytokine_pg_per_mL + 1)

# Check levels of Treatment, Sex, and Experiment:Cage for each Cytokine
cytokine_level_check <- nonCF_serum_luminex_PBSandBt_df %>%
  group_by(Cytokine) %>%
  summarize(
    Treatment_levels = n_distinct(Treatment),
    Sex_levels = n_distinct(Sex),
    Experiment_Cage_levels = n_distinct(interaction(Experiment, Cage))
  )


# Ensure the 'linear_models_serum' directory exists
dir.create("linear_models_nonCF_serum", showWarnings = FALSE)

# Initialize the summary dataframe
linear_models_nonCFserum_summary <- data.frame(
  Cytokine = character(),
  P_Value_Treatment = numeric(),
  P_Value_Sex = numeric(),
  P_Value_Treatment_Sex = numeric(),
  stringsAsFactors = FALSE
)

# Function to safely extract p-value
safe_extract_pvalue <- function(model_summary, term) {
  coef_table <- as.data.frame(model_summary$coefficients)
  if (term %in% rownames(coef_table)) {
    return(coef_table[term, "Pr(>|t|)"])
  }
  return(NA)
}

# Loop over each unique cytokine
for(cytokine in unique(nonCF_serum_luminex_PBSandBt_df$Cytokine)) {
  # Subset data for the current cytokine
  subset_data <- subset(nonCF_serum_luminex_PBSandBt_df, Cytokine == cytokine)

  # Fit the mixed linear model
  model <- lm(log2(Cytokine_pg_per_mL) ~ Treatment + Sex + Sex*Treatment + Experiment:Cage, data = subset_data)

  # Obtain the summary of the model
  model_summary <- summary(model)

  # Extract the p-values for Genotype and Sex
  p_treatment <- safe_extract_pvalue(model_summary, "TreatmentBacteroides")
  p_sex <- safe_extract_pvalue(model_summary, "SexF")
  p_treatment_sex<- safe_extract_pvalue(model_summary, "TreatmentBacteroides:SexF")

  # Add to the summary dataframe
  linear_models_nonCFserum_summary <- rbind(linear_models_nonCFserum_summary, data.frame(
    Cytokine = cytokine,
    P_Value_Treatment = p_treatment,
    P_Value_Sex = p_sex,
    P_Value_Treatment_Sex = p_treatment_sex
  ))

  # Save the summary of each model to a text file
  file_name <- paste0("linear_models_nonCF_serum/model_", cytokine, ".txt")
  capture.output(model_summary, file = file_name)
}

# View the summary dataframe
print(linear_models_nonCFserum_summary)

# Function to categorize significance
categorize_significance <- function(p_value) {
  if(p_value < 0.001) {
    return("***")
  } else if(p_value < 0.01) {
    return("**")
  } else if(p_value < 0.05) {
    return("*")
  } else {
    return("NS")
  }
}

# Apply the function to the dataframe
linear_models_nonCFserum_summary$Significance_Treatment <- sapply(linear_models_nonCFserum_summary$P_Value_Treatment, categorize_significance)
linear_models_nonCFserum_summary$Significance_Sex <- sapply(linear_models_nonCFserum_summary$P_Value_Sex, categorize_significance)

# View the updated dataframe
linear_models_nonCFserum_summary

# Apply Bonferroni correction to the P_Value_Genotype column
linear_models_nonCFserum_summary$P_Value_Treatment_Bonferroni <- p.adjust(linear_models_nonCFserum_summary$P_Value_Treatment, method = "bonferroni")
linear_models_nonCFserum_summary$P_Value_Sex_Bonferroni <- p.adjust(linear_models_nonCFserum_summary$P_Value_Sex, method = "bonferroni")



# Save the dataframe as a CSV file
write.csv(linear_models_nonCFserum_summary, "linear_models_nonCFserum_summary.csv", row.names = FALSE)



```


#Run a loop to find all cytokines that are differentially expressed in the lungs between PBS and Bacteroides for nonCF mice, across both experiments

```{r}

#Extract just PBS and Bacteroides
nonCF_lungs_luminex_PBSandBt_df <- luminex_df %>%
  filter(Genotype == "C57BL_6J",
         SampleType == "Lungs",
         Treatment %in% c("PBS", "Bacteroides"))

nonCF_lungs_luminex_PBSandBt_df <- nonCF_lungs_luminex_PBSandBt_df %>%
  mutate(Cytokine_pg_per_mL = Cytokine_pg_per_mL + 1)

#factor females first

# Check levels of Treatment, Sex, and Experiment:Cage for each Cytokine
cytokine_level_check2 <- nonCF_lungs_luminex_PBSandBt_df %>%
  group_by(Cytokine) %>%
  summarize(
    Treatment_levels = n_distinct(Treatment),
    Sex_levels = n_distinct(Sex),
    Experiment_Cage_levels = n_distinct(interaction(Experiment, Cage))
  )


# Ensure the 'linear_models_serum' directory exists
dir.create("linear_models_nonCF_lungs", showWarnings = FALSE)

# Initialize the summary dataframe
linear_models_nonCFlungs_summary <- data.frame(
  Cytokine = character(),
  P_Value_Treatment = numeric(),
  P_Value_Sex = numeric(),
  P_Value_Sex_Treatment = numeric(),
  stringsAsFactors = FALSE
)

# Function to safely extract p-value
safe_extract_pvalue <- function(model_summary2, term) {
  coef_table <- as.data.frame(model_summary2$coefficients)
  if (term %in% rownames(coef_table)) {
    return(coef_table[term, "Pr(>|t|)"])
  }
  return(NA)
}


nonCF_lungs_luminex_PBSandBt_df$Sex <- factor(nonCF_lungs_luminex_PBSandBt_df$Sex, levels = c("F", "M"))

# Loop over each unique cytokine
for(cytokine in unique(nonCF_lungs_luminex_PBSandBt_df$Cytokine)) {
  # Subset data for the current cytokine
  subset_data2 <- subset(nonCF_lungs_luminex_PBSandBt_df, Cytokine == cytokine)

  # Fit the mixed linear model
  model2 <- lm(log2(Cytokine_pg_per_mL) ~ Treatment + Sex + Sex*Treatment+Experiment:Cage, data = subset_data2)

  # Obtain the summary of the model
  model_summary2 <- summary(model2)

  # Extract the p-values for Genotype and Sex
  p_treatment <- safe_extract_pvalue(model_summary2, "TreatmentBacteroides")
  p_sex <- safe_extract_pvalue(model_summary2, "SexM")
  p_sex_treatment <- safe_extract_pvalue(model_summary2, "TreatmentBacteroides:SexM")

  # Add to the summary dataframe
  linear_models_nonCFlungs_summary <- rbind(linear_models_nonCFlungs_summary, data.frame(
    Cytokine = cytokine,
    P_Value_Treatment = p_treatment,
    P_Value_Sex = p_sex,
    P_Value_Sex_Treatment=p_sex_treatment
  ))

  # Save the summary of each model to a text file
  file_name <- paste0("linear_models_nonCF_lungs/model_", cytokine, ".txt")
  capture.output(model_summary2, file = file_name)
}

# View the summary dataframe
print(linear_models_nonCFlungs_summary)

# Function to categorize significance
categorize_significance <- function(p_value) {
  if(p_value < 0.001) {
    return("***")
  } else if(p_value < 0.01) {
    return("**")
  } else if(p_value < 0.05) {
    return("*")
  } else {
    return("NS")
  }
}

# Apply the function to the dataframe
linear_models_nonCFlungs_summary$Significance_Treatment <- sapply(linear_models_nonCFlungs_summary$P_Value_Treatment, categorize_significance)
linear_models_nonCFlungs_summary$Significance_Sex <- sapply(linear_models_nonCFlungs_summary$P_Value_Sex, categorize_significance)
linear_models_nonCFlungs_summary$Significance_Sex_Treatment <- sapply(linear_models_nonCFlungs_summary$P_Value_Sex_Treatment, categorize_significance)

# View the updated dataframe
linear_models_nonCFlungs_summary

# Apply Bonferroni correction to the P_Value_Genotype column
linear_models_nonCFlungs_summary$P_Value_Treatment_Bonferroni <- p.adjust(linear_models_nonCFlungs_summary$P_Value_Treatment, method = "bonferroni")
linear_models_nonCFlungs_summary$P_Value_Sex_Bonferroni <- p.adjust(linear_models_nonCFlungs_summary$P_Value_Sex, method = "bonferroni")
linear_models_nonCFlungs_summary$P_Value_Sex_Treatment_Bonferroni <- p.adjust(linear_models_nonCFlungs_summary$P_Value_Sex_Treatment, method = "bonferroni")



# Save the dataframe as a CSV file
write.csv(linear_models_nonCFlungs_summary, "linear_models_nonCFlungs_summary.csv", row.names = FALSE)

```

#Make a volcano plot showing fold change on x axis and p valye on y axis for lungs cytokines
```{r}

# Filter data for nonCF and CF
nonCF_PBS_data_lungs <- nonCF_lungs_luminex_PBSandBt_df[nonCF_lungs_luminex_PBSandBt_df$Treatment == 'PBS', ]
nonCF_Bt_data_lungs <- nonCF_lungs_luminex_PBSandBt_df[nonCF_lungs_luminex_PBSandBt_df$Treatment == 'Bacteroides', ]

# Calculate median values for each cytokine in both nonCF and CF groups
median_nonCF_PBS_lungs <- aggregate(Cytokine_pg_per_mL ~ Cytokine, nonCF_PBS_data_lungs, median)
median_nonCF_Bt_lungs <- aggregate(Cytokine_pg_per_mL ~ Cytokine, nonCF_Bt_data_lungs, median)


# Merge the two dataframes on cytokine names
foldchange_nonCF_lungs <- merge(median_nonCF_PBS_lungs, median_nonCF_Bt_lungs, by = "Cytokine")

# Calculate fold changes
foldchange_nonCF_lungs$FoldChange <- foldchange_nonCF_lungs$Cytokine_pg_per_mL.y / foldchange_nonCF_lungs$Cytokine_pg_per_mL.x

# Rename columns for clarity
names(foldchange_nonCF_lungs) <- c("Cytokine", "Median_PBS", "Median_Bacteroides", "FoldChange")

# Display the dataframe
print(foldchange_nonCF_lungs)

# Merge the 'P_Value_Genotype' column from 'linear_models_lungs_summary' into 'foldchange_lungs'
foldchange_nonCF_lungs <- merge(foldchange_nonCF_lungs, linear_models_nonCFlungs_summary[, c("Cytokine", "P_Value_Treatment")], by = "Cytokine", all.x = TRUE)



```

#Draw a volcano plot on lungs
```{r}
# Transform the data
foldchange_nonCF_lungs$log2FoldChange <- log2(foldchange_nonCF_lungs$FoldChange)
foldchange_nonCF_lungs$negLog10PValue <- -log10(foldchange_nonCF_lungs$P_Value_Treatment)

# Add a new column "DiffExpression" to the foldchange_lungs DataFrame
foldchange_nonCF_lungs$DiffExpression <- ifelse(foldchange_nonCF_lungs$P_Value_Treatment < 0.05, 
                                          ifelse(foldchange_nonCF_lungs$FoldChange > 1, "Up in Bacteroides", 
                                                 ifelse(foldchange_nonCF_lungs$FoldChange < 1, "Down in Bacteroides", "Not Significant")),
                                          "Not Significant")

# Biostatsquid theme
theme_set(theme_classic(base_size = 20) +
            theme(
              axis.title.y = element_text(face = "bold", margin = margin(0,20,0,0), size = rel(0.5), color = 'black'),
              axis.title.x = element_text(hjust = 0.5, face = "bold", margin = margin(20,0,0,0), size = rel(0.5), color = 'black'),
              plot.title = element_text(hjust = 0.5)
            ))


# Create a basic volcano plot
Volcanoplot_PBSvsBacteroides_nonCFlungs<- ggplot(data = foldchange_nonCF_lungs, aes(x = log2FoldChange, y = negLog10PValue, color = DiffExpression)) +
         geom_point()+
  geom_vline(xintercept = c(-0.6, 0.6), col = "gray", linetype = 'dashed') +
  geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed') +
  scale_color_manual(values = c("Up in Bacteroides" = "#00AFBB", "Down in Bacteroides" = "#bb0c00", "Not Significant" = "grey"))+
  geom_text(aes(label=Cytokine), vjust = 2, hjust = 0.5, size = 2, check_overlap = TRUE)+
  ggtitle("Cytokines Differentially Expressed Between PBS and Bacteroides gavaged nonCF Mice in lungs")+
  theme(plot.title = element_text(size = 12)) # Reduced font size for title


# Save the plot as a PDF
ggsave("Volcanoplot_PBSvsBacteroides_nonCFlungs.pdf", plot = Volcanoplot_PBSvsBacteroides_nonCFlungs, width = 10, height = 4, device = "pdf")


```

#Run correlation between KC in the lungs vs serum
```{r}
library(ggpmisc)



#Create combined ID column in both dataframes
nonCF_lungs_KC_luminex_df$Mouse_ID <- paste(nonCF_lungs_KC_luminex_df$Experiment,
                                             nonCF_lungs_KC_luminex_df$Cage,
                                             nonCF_lungs_KC_luminex_df$Mouse,
                                             sep = "_") 
nonCF_serum_KC_luminex_df$Mouse_ID <- paste(nonCF_serum_KC_luminex_df$Experiment,
                                            nonCF_serum_KC_luminex_df$Cage,
                                            nonCF_serum_KC_luminex_df$Mouse,
                                            sep = "_")

#Rename the Cytokine column in both dataframes
names(nonCF_lungs_KC_luminex_df)[names(nonCF_lungs_KC_luminex_df) == "Cytokine_pg_per_mL"] <- "Cytokine_pg_per_mL_lungs"
names(nonCF_serum_KC_luminex_df)[names(nonCF_serum_KC_luminex_df) == "Cytokine_pg_per_mL"] <- "Cytokine_pg_per_mL_serum"

#Select only Mouse_ID and serum cytokine column from the serum dataframe
serum_subset <- nonCF_serum_KC_luminex_df[, c("Mouse_ID", "Cytokine_pg_per_mL_serum")]

#Merge serum data into lungs dataframe by Mouse_ID
nonCF_combined_KC_luminex_df <- merge(nonCF_lungs_KC_luminex_df, serum_subset, by = "Mouse_ID", all.x = TRUE)




serum_lungs_KCcorrelation <- ggplot(nonCF_combined_KC_luminex_df, 
                                    aes(x = log2(Cytokine_pg_per_mL_serum), 
                                        y = log2(Cytokine_pg_per_mL_lungs), 
                                        color = Treatment)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linewidth = 1) +
  stat_poly_eq(aes(label = after_stat(eq.label)), 
               formula = y ~ x,
               parse = TRUE, 
               size = 3,
               label.x = "right", 
               label.y = "top") +
  scale_color_manual(values = c("Baseline" = "lightgreen", 
                                "PBS" = "lightblue", 
                                "Bacteroides" = "orange")) +
  labs(x = "KC in Serum (log2 pg/mL)", 
       y = "KC in Lungs (log2 pg/mL)") +
  theme_bw() +
  guides(color = guide_legend(title = "Treatment Group")) +
  theme(legend.position = "right")

serum_lungs_KCcorrelation

ggsave("serum_lungs_KCcorrelation.pdf", plot = serum_lungs_KCcorrelation, device = "pdf", width = 5, height = 3)


#get R2 and p value for each treatment group
#Subset just Bacteroides 
bacteroides_df <- subset(nonCF_combined_KC_luminex_df, Treatment == "Bacteroides")
lm_model_bt <- lm(log2(Cytokine_pg_per_mL_lungs) ~ log2(Cytokine_pg_per_mL_serum), data = bacteroides_df)
summary(lm_model_bt)

#Subset just PBS 
PBS_df <- subset(nonCF_combined_KC_luminex_df, Treatment == "PBS")
lm_model_pbs <- lm(log2(Cytokine_pg_per_mL_lungs) ~ log2(Cytokine_pg_per_mL_serum), data = PBS_df)
summary(lm_model_pbs)

#Subset just Baseline 
Baseline_df <- subset(nonCF_combined_KC_luminex_df, Treatment == "Baseline")
lm_model_baseline <- lm(log2(Cytokine_pg_per_mL_lungs) ~ log2(Cytokine_pg_per_mL_serum), data = Baseline_df)
summary(lm_model_baseline)


library(ggplot2)
library(ggpmisc)

serum_lungs_KCcorrelation <- ggplot(nonCF_combined_KC_luminex_df, 
                                    aes(x = log2(Cytokine_pg_per_mL_lungs), 
                                        y = log2(Cytokine_pg_per_mL_serum), 
                                        color = Treatment)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linewidth = 1) +
  stat_poly_eq(
    aes(label = after_stat(paste(eq.label, rr.label, p.value.label, sep = "~~~"))),
    formula = y ~ x,
    parse = TRUE,
    size = 4.5,
    label.x = "left",
    label.y = "top"
  ) +
  scale_color_manual(values = c("Baseline" = "lightgreen", 
                                "PBS" = "lightblue", 
                                "Bacteroides" = "orange")) +
  labs(x = "KC in Lungs (log2 pg/mL)", 
       y = "KC in Serum (log2 pg/mL)") +
  theme_bw() +
  guides(color = guide_legend(title = "Treatment Group")) +
  theme(legend.position = "right")

serum_lungs_KCcorrelation

ggsave("serum_lungs_KCcorrelation.pdf", plot = serum_lungs_KCcorrelation, device = "pdf", width = 7, height = 5)


```
