---
title: "March2024_BtEstablishment_WT_Mice"
author: "Rebecca Valls"
date: "2024-03-13"
output: html_document
---

## Load the necessary libraries
```{r, message = FALSE}
#install.packages("tidyverse")
library(tidyverse) #R package used for data manipulation and visualization built to work with dataframes. Includes R packages dplyr, ggplot2, tidyr and others.

#if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
#BiocManager::install("phyloseq")
library(phyloseq) #R package that allows for the analysis of microbiome data.  

#BiocManager::install("ggforce") #R package that extends the functionality of ggplot2,
library(ggforce)

#install.packages("htmltools")
library(htmltools)

#install.packages("vegan") #an R package that provides tools for performing ecological and environmental analysis.
library(vegan)

#install.packages("viridis") #n R package that provides a set of color palettes for data visualization. The palettes are designed to be perceptually uniform and colorblind-friendly, making them well-suited for visualizing data in scientific and data-driven contexts. 
library(viridis)

#BiocManager::install("ggpubr") #provides a convenient set of tools for creating and customizing 'ggplot2' plots, with functions designed to simplify common tasks such as adding statistical tests, arranging multiple plots, and enhancing plot appearance.
library(ggpubr)


#BiocManager::install("decontam") #helps identify and remove contaminant sequences from microbiome data, improving the accuracy of the analysis. It applies statistical methods to distinguish real biological sequences from contaminants, thereby improving the accuracy and reliability of microbiome studies.
library(decontam)

#install.packages("broom") #simplifies the extraction and organization of statistical model outputs into tidy data frames, making it easier to analyze, manipulate, and visualize model results
library(broom)

#install.packages("lme4") #used for fitting linear and generalized linear mixed-effects models
library(lmerTest)

#install.packages("devtools")
#devtools::install_github("pmartinezarbizu/pairwiseAdonis/pairwiseAdonis")

library("pairwiseAdonis")

#BiocManager::install("randomForest") #version 4.7.1 #used for fitting random forest models for classication and regression
library(randomForest) 

#BiocManager::install("caTools") #used for moving (rolling, running) window statistic functions, read/write for GIF and ENVI binary files, fast calculation of AUC, LogitBoost classifier, base64 encoder/decoder, round-off error free sum and cumsum, etc.
library(caTools)

#BiocManager::install("caret") #used for classification and regression training
library(caret)

#BiocManager::install("Maaslin2") #MaAsLin2 is comprehensive R package for efficiently determining multivariable association between clinical metadata and microbial meta'omic features. MaAsLin2 relies on general linear models to accommodate most modern epidemiological study designs, including cross-sectional and longitudinal, and offers a variety of data exploration, normalization, and transformation methods.
library(Maaslin2)


library(pairwiseAdonis)
```

#### Bring in data for making your phyloseq object

### Import the DADA2 output ASV table as a data.frame
```{r include=FALSE}
#Bring in the counts
count_tab <- read.table("Jan2025_WTMouseBtEstablishment_compiled_counts.tsv", header=T, row.names=1, check.names=F, sep="\t")
head(count_tab)

#remove samples that were resequenced//some had original lower quality samples that were included in DADA2 and i want to remove these so that we don't have duplicate sequencing
# Define the column names to remove
columns_to_remove <- c(
  "5516_106_S102",
  "5516_111_S107",
  "5516_113_S109",
  "5516_145_S141",
  "5516_147_S143",
  "5516_153_S149",
  "4683_35_S195"
)

# Remove the specified columns
count_tab_filt <- count_tab[, !(colnames(count_tab) %in% columns_to_remove)]


#remove part of column names that say "_S**" or "Reseq_S*", and then remove the 0 before numbers for 5516 sequencing

colnames(count_tab_filt) <- gsub("_?Reseq_S\\d+|_S\\d+", "", colnames(count_tab_filt))
colnames(count_tab_filt) <- gsub("5516_0+", "5516_", colnames(count_tab_filt))


```

### Import the DADA2 output taxa table as a matrix
```{r}
#bring in the taxonomy classification for each ASV,
tax_tab <- as.matrix(read.table("Jan2025_WTMouseBtEstablishment_compiled_taxa.tsv", header=T, row.names=1, check.names=F, sep="\t"))
head(tax_tab)

```

### Import the sample information and place them in desired order
```{r}
#Bring in the mouse and treatment information
SeqCoast_Sample_metadata_nonCF_Exp1n2 <- read_csv("~/Dartmouth College Dropbox/Rebecca Valls/Projects/Project_MalevsFemale_nonCFmice_2023/3_Experiment/3_DataAnalysis/16s_analysis/SeqCoast_Sample_metadata_nonCF_Exp1n2.csv")
metadata_all <- SeqCoast_Sample_metadata_nonCF_Exp1n2
View(metadata_all)

#Add a column with the Biogeography and the Microniche
# Assuming 'metadata_all' is your dataframe
metadata_all <- metadata_all %>%
  # Create the 'Biogeography' column
  mutate(Biogeography = case_when(
    grepl("SI", TubeLabel) ~ "SmallIntestine",
    grepl("Cec", TubeLabel) ~ "Cecum",
    grepl("Pro", TubeLabel) ~ "ProximalColon",
    grepl("Dis", TubeLabel) ~ "DistalColon",
    TRUE ~ NA_character_ # default case
  )) %>%
  # Create the 'Microniche' column
  mutate(Microniche = case_when(
    grepl("LC", TubeLabel) ~ "LuminalContents",
    grepl("MS", TubeLabel) ~ "MucosalSurface",
    TRUE ~ NA_character_ # default case
  ))

# add sample_or_control column to identify samples for decontam
metadata_all <- metadata_all %>%
  mutate(sample_or_control = ifelse(TubeLabel == "CollectionControl", "Control", "Sample"))


#Make the sample names the row names for clinical information
metadata_all <- metadata_all %>% remove_rownames %>% column_to_rownames(var="Sample_ID")
metadata_all <- as.data.frame(metadata_all)
sample_info_tab <- as.data.frame(metadata_all)

##remove unannotated samples from count_tab and metadata. This removes 0 samples!
count_tab_filtered_w_metadata <- count_tab_filt[,colnames(count_tab_filt) %in% rownames(sample_info_tab)]
sample_info_tab <- sample_info_tab[rownames(sample_info_tab) %in% colnames(count_tab_filt),]
                                   

#Factor time point to be in correct order
print(sample_info_tab$Compiled_Timepoint)
sample_info_tab$Compiled_Timepoint <- factor(sample_info_tab$Compiled_Timepoint, levels=c("Baseline", "Late_Baseline", "Antibiotics", "Early_Gavage", "Late_Gavage", "Before_LPS", "LPS", "Sacrifice"))

sample_info_tab$Treatment <- factor(sample_info_tab$Treatment, levels=c("Baseline", "PBS", "Bacteroides"))

sample_info_tab$Biogeography <- factor(sample_info_tab$Biogeography, levels=c("SmallIntestine", "Cecum", "ProximalColon", "DistalColon"))

sample_info_tab$Sex <- factor(sample_info_tab$Sex, levels = c("F", "M"))



```


### Set the theme for plots
Sets up a global theme for all plots created during this session

```{r, message = FALSE}
theme_set(theme_bw()) # set global theme to bw
theme_replace(axis.title.x=element_text(size=14), # set individual theme options for plotting

              axis.title.y=element_text(size=14, angle = 90),
              axis.text.x = element_text(size = 12,
                                         angle = 45,
                                         hjust = 1,
                                         vjust = 1),
              axis.text.y=element_text(size=12),
              legend.title=element_text(size=14),
              legend.text=element_text(size=12))
```

### Instructions for Phyloseq
#### Make phyloseq object
```{r}

#transform data (?) into phyloseq object
#for some reason, if you don't do this, phyloseq will not recognize "Month" header
sample_info_tab_phy <- sample_data(sample_info_tab)
count_tab_phy <- otu_table(count_tab_filtered_w_metadata, taxa_are_rows=T)
tax_tab_phy <- tax_table(tax_tab)

#make phyloseq object
btEstablishmentWTmice_phyloseq <- phyloseq(count_tab_phy, tax_tab_phy, sample_info_tab_phy)
btEstablishmentWTmice_phyloseq
```


# Let's decontaminate the samples
The decontam package in R identifies and removes contaminant sequences from marker-gene and metagenomics data. By using statistical methods based on sequence prevalence or frequency, it distinguishes true biological sequences from contaminants. This improves data accuracy in microbiome research, ensuring that results reflect the actual microbial community composition.

```{r}
df <- as.data.frame(sample_data(btEstablishmentWTmice_phyloseq)) # put sample_data into ggplot friendly data frame

df$librarysize <- sample_sums(btEstablishmentWTmice_phyloseq) #generate librarysize column - by summing read counts
```

### Plot library size
Visualize which samples have small or large library sizes

```{r, message = FALSE}
ggplot(data = df, # input dataframe
       aes(x = reorder(TubeLabel,librarysize), # reorder sample by ascending library size
           y = librarysize, # set y axis
           color = sample_or_control)) + # set color of points
  geom_point(shape = 1, # set point shape
             alpha = 0.5) + # change point size
  ylab("Library size") + # relabel y axis
  xlab("Sample") + # relabel x axis
  scale_color_discrete(labels = c("Control","Sample")) + # relabel legend categories
  theme(axis.text.x = element_blank(), # change vertical justification 
        axis.ticks.x = element_blank(), # remove x axis ticks
        panel.grid.minor = element_blank(), # remove minor gridlines
        panel.grid.major = element_blank(), # remove major gridlines
        legend.title = element_blank(), # remove legend title
        legend.position = c(0.1,0.9), # move legend
        legend.background = element_rect(color = "black")) # set outer boundary of legend box
```
This generates a warning that there are four samples with 0 reads - we should remove these from our further analysis using *phyloseq*.

```{r, message = FALSE}
prune_samples(sample_sums(btEstablishmentWTmice_phyloseq)!=0, btEstablishmentWTmice_phyloseq) -> btEstablishmentWTmice_phyloseq_NZ # prune samples with 0 reads
# rename as filtered phyloseq object so that we do not mix up the two lots of data
```


### Identify contaminants through prevalence
*decontam* also provides a method for identifying contaminants through their presence and absence in known samples and controls. This method does not take into account the DNA concentrations of samples, just whether or not they are identified as sample or control in the metadata.

```{r, message = FALSE}
sample_data(btEstablishmentWTmice_phyloseq_NZ)$is.neg <- sample_data(btEstablishmentWTmice_phyloseq_NZ)$sample_or_control == "Control" # create new variable in phyloseq object that identifies negative controls based on sample_or_control column

contamdfPrev <- isContaminant(btEstablishmentWTmice_phyloseq_NZ, # input phyloseq object
                               method = "prevalence", # define method for contaminant identification
                               neg = "is.neg") # define column for identifying negative controls

table(contamdfPrev$contaminant)
```
### Activity - what happens if we change the threshold for contaminant identification in the "prevalence" method?
Note: at threshold 0.5, *decontam* will call all ASVs that are more prevalent in controls than samples as contaminants.

```{r, message = FALSE}
contamdfPrevThr <- isContaminant(btEstablishmentWTmice_phyloseq_NZ, # input phyloseq object
                                 method = "prevalence", # define method for contaminant identification
                                 neg = "is.neg", # define column for identifying negative controls
                                 threshold = 0.5) # define threshold for contaminant identification

table(contamdfPrevThr$contaminant)

print(contamdfPrevThr[contamdfPrevThr$contaminant == TRUE, ])
print(tax_tab["ASV_26006", ])


```
### Plot ASV prevalence in samples vs. controls to check if this method is appropriate

```{r, message = FALSE}
psPrev <- transform_sample_counts(btEstablishmentWTmice_phyloseq_NZ, 
                                  function(abund) 1*(abund>0)) # transform sample counts

psPrev.neg <- prune_samples(sample_data(psPrev)$sample_or_control == "Control", 
                            psPrev) # make list of controls

psPrev.pos <- prune_samples(sample_data(psPrev)$sample_or_control == "Sample", 
                            psPrev) # make list of samples

dfPrev <- data.frame(prevPos = taxa_sums(psPrev.pos), # build dataframe
                     prevNeg = taxa_sums(psPrev.neg),
                     contaminant = contamdfPrev$contaminant)

ggplot(data = dfPrev, # input dataframe
       aes(x = prevNeg, # x axis prevalence in negative controls
           y = prevPos, # y axis prevalence in samples
           color = contaminant)) + 
  geom_point(shape = 1) +
  xlab("Prevalence (controls)") + 
  ylab("Prevalence (samples)")
```


```{r}
btEstablishmentWTmice_phyloseq_NZ # check dimensions of current phyloseq object
```


"contaminant" is a Boolean variable so using an exclamation point tells phyloseq not to select ASVs which have TRUE in the contaminant column for taxa pruning.

```{r}
Phyloseq_object_NonContam <- prune_taxa(!contamdfPrevThr$contaminant,
                          btEstablishmentWTmice_phyloseq_NZ)



Phyloseq_object_NonContam # check dimensions of new phyloseq object #removed 1 contaminant
```

#Plot the alpha diversity across experiment, showing each treatment group
```{r}

# Define custom colors for Treatment groups
custom_colors <- c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")

#Filter data to only include subset groups

#Data from samples collected across experiment
btphylo_nosacrifice <- subset_samples(Phyloseq_object_NonContam, Compiled_Timepoint != "Sacrifice")
btphylo_nosacrifice <- subset_samples(btphylo_nosacrifice, sample_or_control != "C")


#Data from samples collected across gut at sacrifice
btphylo_atsacrifice <- subset_samples(Phyloseq_object_NonContam, Compiled_Timepoint == "Sacrifice")


colnames(sample_data(btphylo_nosacrifice))

#Plot Chao1 alpha diversity for time point. Here you should use unnormalized data.
alpha_diversity_chao1 <- plot_richness(btphylo_nosacrifice, x="Experiment_Day", color = "Treatment", measures=c("Chao1"))+
  stat_summary(fun.y = mean, size=1, geom = "line", aes(group = Treatment)) +
  geom_point(size=1.5)+
  theme_bw()+
  geom_vline(xintercept = c(0, -25, 12), linetype = "dashed", color = "blue", size = 0.5) +
  geom_text(aes(x = -25, y = Inf, label = "ABX"), vjust = 2, color = "red", hjust = 0, size = 3) +
  geom_text(aes(x = 0, y = Inf, label = "Gavage"), vjust = 2, color = "red", hjust = 0, size = 3) +
  geom_text(aes(x = 12, y = Inf, label = "LPS"), vjust = 2, color = "red", hjust = 0, size = 3) +
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  geom_line(aes(group = interaction("SeqCoastOrder","Cage..","Mouse..")), size = 0.25)  # Add this line to connect points for each MouseID

alpha_diversity_chao1

ggsave("alpha_diversity_chao1_acrossexperiment.pdf", plot = alpha_diversity_chao1, device = "pdf", width = 10, height = 4)





```



```{r}
alpha_diversity_shannon <- plot_richness(btphylo_nosacrifice, x="Experiment_Day", color = "Treatment", measures=c("Shannon"))+
  stat_summary(fun.y = mean, size=2, geom = "line", aes(group = Treatment)) +
  geom_point(size=1)+
  theme_bw()+
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  geom_line(aes(group = interaction("SeqCoastOrder","Cage #","Mouse #")), size = 0.25)  # Add this line to connect points for each MouseID

alpha_diversity_shannon

#plot alpha diversity across expreimemnt with boxplot
alpha_diversity_shannon_v2 <- plot_richness(btphylo_nosacrifice, x="Compiled_Timepoint", color = "Treatment", measures=c("Shannon"))+
  geom_boxplot()+
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange"))+
  labs(y = "Alpha Diversity - Shannon", x = "Experiment Day", color = "Treatment Group")+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=0.5))
  

alpha_diversity_shannon_v2
?plot_richness

alpha_diversity_simpson <- plot_richness(btphylo_nosacrifice, x="Compiled_Timepoint", color = "Treatment", measures=c("Simpson"))+
  geom_boxplot()+
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange"))+
  labs(y = "Alpha Diversity - Simpson", x = "Experiment Day", color = "Treatment Group")+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=0.5))
  

alpha_diversity_simpson

alpha_diversity_Chao1 <- plot_richness(btphylo_nosacrifice, x="Compiled_Timepoint", color = "Treatment", measures=c("Chao1"))+
  geom_boxplot()+
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange"))+
  labs(y = "Alpha Diversity - Chao1", x = "Experiment Day", color = "Treatment Group")+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=0.5))
  

alpha_diversity_Chao1



#Plot Fisher alpha diversity for time point. Here you should use unnormalized data.
alpha_diversity_fisher <- plot_richness(btphylo_nosacrifice, x="Experiment_Day", color = "Treatment", measures=c("Fisher"))+
  geom_point(size=1.5)+
  theme_bw()+
  geom_vline(xintercept = c(0, -25, 12), linetype = "dashed", color = "blue", size = 0.5) +
  geom_text(aes(x = -25, y = Inf, label = "ABX"), vjust = 2, color = "red", hjust = 0, size = 3) +
  geom_text(aes(x = 0, y = Inf, label = "Gavage"), vjust = 2, color = "red", hjust = 0, size = 3) +
  geom_text(aes(x = 12, y = Inf, label = "LPS"), vjust = 2, color = "red", hjust = 0, size = 3) +
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  geom_line(aes(group = interaction("SeqCoastOrder","Cage..","Mouse..")), size = 0.25)  # Add this line to connect points for each MouseID

alpha_diversity_fisher

ggsave("alpha_diversity_fisher_boxplot.pdf", plot = alpha_diversity_fisher_boxplot, device = "pdf", width = 10, height = 4)

#pull alpha diversity data and plot across experiment
# Step 1: Estimate Fisher alpha diversity from the phyloseq object
fisher_diversity_nosacrifice <- estimate_richness(btphylo_nosacrifice, measures = "Fisher")

# Step 2: Combine Fisher diversity with sample data
# Convert the Fisher diversity to a data frame
fisher_nosacrifice_df <- as.data.frame(fisher_diversity_nosacrifice)

# Ensure the row names (samples) are included in the data frame
fisher_nosacrifice_df$Sample <- rownames(fisher_nosacrifice_df)
metadata_all$Sample <- rownames(metadata_all)


# Merge with your metadata
combined_nosacrifice_df <- fisher_nosacrifice_df %>%
  left_join(metadata_all, by ="Sample")  # Adjust "SampleID" according to your metadata

# Set factor levels for Treatment to ensure "PBS" comes before "Bacteroides"
combined_nosacrifice_df$Treatment <- factor(combined_nosacrifice_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))


combined_nosacrifice_filtered <- combined_nosacrifice_df %>%
  filter(Compiled_Timepoint != "Baseline")  # Keep only rows where Compiled_Timepoint is not "Baseline"

combined_nosacrifice_filtered$Compiled_Timepoint <- factor(
  combined_nosacrifice_filtered$Compiled_Timepoint, 
  levels = c("Late_Baseline", "Antibiotics", "Early_Gavage", "Late_Gavage", "Before_LPS", "LPS")
)

# Step 3: Create the alpha diversity plot for Fisher's measure across gut
alpha_fisher_across_experiment <- ggplot(combined_nosacrifice_filtered, 
                                   aes(x = Compiled_Timepoint, y = Fisher, fill = Treatment)) +
  geom_boxplot(color = "black", alpha = 0.6) +  # Boxplot with borders and transparency
  geom_point(aes(color = Treatment, shape = Sex),  # Points mapped to Treatment and shape
             position = position_jitterdodge(dodge.width = 0.75),  # Jitter points for visibility
             alpha = 0.8, size = 2) +  # Points with defined aesthetics
   scale_fill_manual(values = c("Baseline" = "lightgreen", 
                                "PBS" = "lightblue", 
                                "Bacteroides" = "orange")) +  # Fill colors for treatments
  scale_color_manual(values = c("Baseline" = "lightgreen", 
                                 "PBS" = "lightblue", 
                                 "Bacteroides" = "orange")) +  # Color mapping for points
  theme_minimal()+
  labs(x = "Experiment Day",
       y = expression("Alpha Diversity - Fisher"),  # Y-axis label with expression
       fill = "Treatment") +  # Updated fill label
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
        legend.position = "right")  # Position the legend

# Display the plot
print(alpha_fisher_across_experiment)

ggsave("alpha_fisher_across_experiment_boxplot.pdf", plot = alpha_fisher_across_experiment, device = "pdf", width = 10, height = 4)

# Subset the data for Late_Baseline

Late_Baseline_fisher_df <- combined_nosacrifice_df %>%
  filter(Compiled_Timepoint %in% c("Late_Baseline"))  # Keep rows for both time points

Late_Baseline_fisher_df$Treatment <- factor(Late_Baseline_fisher_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))

Late_Baseline_fisher_df$Treatment <- factor(Late_Baseline_fisher_df$Treatment, 
                                         levels = c("PBS", "Bacteroides", "Baseline"))

# Run the linear model on the subset
shapiro_latebaseline_fisher <- shapiro.test(Late_Baseline_fisher_df$Fisher)
shapiro_latebaseline_shapiro

fisher_alpha_lm_Late_Baseline <- lm(Fisher ~ Treatment + Sex + Treatment*Sex+Cage, data = Late_Baseline_fisher_df)
summary(fisher_alpha_lm_Late_Baseline)

# Subset the data for Antibiotics

Antibiotics_fisher_df <- combined_nosacrifice_df %>%
  filter(Compiled_Timepoint %in% c("Antibiotics"))  # Keep rows for both time points

Antibiotics_fisher_df$Treatment <- factor(Antibiotics_fisher_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))


# Run the linear model on the subset
shapiro_antibiotics_fisher <- shapiro.test(Antibiotics_fisher_df$Fisher)
shapiro_antibiotics_fisher

fisher_alpha_lm_Antibiotics <- lm(log2(Fisher) ~ Treatment + Sex + Treatment*Sex+Cage, data = Antibiotics_fisher_df)
summary(fisher_alpha_lm_Antibiotics)

# Subset the data for Early_Gavage

Early_Gavage_fisher_df <- combined_nosacrifice_df %>%
  filter(Compiled_Timepoint %in% c("Early_Gavage"))  # Keep rows for both time points

Early_Gavage_fisher_df$Treatment <- factor(Early_Gavage_fisher_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))



# Run the linear model on the subset
shapiro_early_gavage_fisher <- shapiro.test(Early_Gavage_fisher_df$Fisher)
shapiro_early_gavage_fisher


fisher_alpha_lm_Early_Gavage <- lm(Fisher ~ Treatment + Sex + Treatment*Sex+Cage, data = Early_Gavage_fisher_df)
summary(fisher_alpha_lm_Early_Gavage)

# Subset the data for Late_Gavage

Late_Gavage_fisher_df <- combined_nosacrifice_df %>%
  filter(Compiled_Timepoint %in% c("Late_Gavage"))  # Keep rows for both time points

Late_Gavage_fisher_df$Treatment <- factor(Late_Gavage_fisher_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))



# Run the linear model on the subset
shapiro_late_gavage_fisher <- shapiro.test(Late_Gavage_fisher_df$Fisher)
shapiro_late_gavage_fisher

fisher_alpha_lm_Late_Gavage <- lm(log2(Fisher) ~ Treatment + Sex + Treatment*Sex+Cage, data = Late_Gavage_fisher_df)
summary(fisher_alpha_lm_Late_Gavage)

# Subset the data for Before_LPS

Before_LPS_fisher_df <- combined_nosacrifice_df %>%
  filter(Compiled_Timepoint %in% c("Before_LPS"))  # Keep rows for both time points

Before_LPS_fisher_df$Treatment <- factor(Before_LPS_fisher_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))



# Run the linear model on the subset
shapiro_before_lps_fisher <- shapiro.test(Before_LPS_fisher_df$Fisher)
shapiro_before_lps_fisher

fisher_alpha_lm_Before_LPS <- lm(log2(Fisher) ~ Treatment + Sex + Treatment*Sex+Cage, data = Before_LPS_fisher_df)
summary(fisher_alpha_lm_Before_LPS)

# Subset the data for LPS

LPS_fisher_df <- combined_nosacrifice_df %>%
  filter(Compiled_Timepoint %in% c("LPS"))  # Keep rows for both time points

LPS_fisher_df$Treatment <- factor(LPS_fisher_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))



# Run the linear model on the subset
shapiro_lps_fisher <- shapiro.test(LPS_fisher_df$Fisher)
shapiro_before_lps_fisher

fisher_alpha_lm_LPS <- lm(log2(Fisher) ~ Treatment + Sex + Treatment*Sex+Cage, data = LPS_fisher_df)
summary(fisher_alpha_lm_LPS)
```

Plot the Pielou's Evenness and Observed Richness
```{r}
# Step 1: Estimate Shannon alpha diversity from the phyloseq object
Shannon_diversity_nosacrifice <- estimate_richness(btphylo_nosacrifice, measures = "Shannon")
observed_diversity_nosacrifice <- estimate_richness(btphylo_nosacrifice, measures = "Observed")


# Step 2: Combine Shannon diversity with sample data
# Convert the Shannon diversity to a data frame
Shannon_nosacrifice_df <- as.data.frame(Shannon_diversity_nosacrifice)
observed_nosacrifice_df <- as.data.frame(observed_diversity_nosacrifice)


# Ensure the row names (samples) are included in the data frame
Shannon_nosacrifice_df$Sample <- rownames(Shannon_nosacrifice_df)

observed_nosacrifice_df$Sample <- rownames(observed_nosacrifice_df)
observed_nosacrifice_df$Sample <- gsub("X", "", observed_nosacrifice_df$Sample) #remove the x that is coming up for some reason


metadata_all$Sample <- rownames(metadata_all)

# Merge with your metadata
combined_nosacrifice_df <- Shannon_nosacrifice_df %>%
  left_join(metadata_all, by ="Sample")  # Adjust "SampleID" according to your metadata

# Merge with your metadata
combined_nosacrifice_df <- combined_nosacrifice_df %>%
  left_join(observed_nosacrifice_df, by ="Sample")  # Adjust "SampleID" according to your metadata

combined_nosacrifice_df$Pielou <- combined_nosacrifice_df$Shannon / log(combined_nosacrifice_df$Observed)

# Set factor levels for Treatment to ensure "PBS" comes before "Bacteroides"
combined_nosacrifice_df$Treatment <- factor(combined_nosacrifice_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))


combined_nosacrifice_filtered <- combined_nosacrifice_df %>%
  filter(Compiled_Timepoint != "Baseline")  # Keep only rows where Compiled_Timepoint is not "Baseline"

combined_nosacrifice_filtered$Compiled_Timepoint <- factor(
  combined_nosacrifice_filtered$Compiled_Timepoint, 
  levels = c("Late_Baseline", "Antibiotics", "Early_Gavage", "Late_Gavage", "Before_LPS", "LPS")
)


alpha_Observed_across_experiment <- ggplot(combined_nosacrifice_filtered, 
                                   aes(x = Compiled_Timepoint, y = Observed, fill = Treatment)) +
  geom_boxplot(color = "black", alpha = 0.6) +  # Boxplot with borders and transparency
  geom_point(aes(color = Treatment, shape = Sex),  # Points mapped to Treatment and shape
             position = position_jitterdodge(dodge.width = 0.75),  # Jitter points for visibility
             alpha = 0.8, size = 2) +  # Points with defined aesthetics
   scale_fill_manual(values = c("Baseline" = "lightgreen", 
                                "PBS" = "lightblue", 
                                "Bacteroides" = "orange")) +  # Fill colors for treatments
  scale_color_manual(values = c("Baseline" = "lightgreen", 
                                 "PBS" = "lightblue", 
                                 "Bacteroides" = "orange")) +  # Color mapping for points
  theme_minimal()+
  labs(x = "Experiment Day",
       y = expression("Alpha Diversity - Observed"),  # Y-axis label with expression
       fill = "Treatment") +  # Updated fill label
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
        legend.position = "right")  # Position the legend

# Display the plot
print(alpha_Observed_across_experiment)


# Step 3: Create the alpha diversity plot for Shannon's measure across gut
alpha_Pielou_across_experiment <- ggplot(combined_nosacrifice_filtered, 
                                   aes(x = Compiled_Timepoint, y = Pielou, fill = Treatment)) +
  geom_boxplot(color = "black", alpha = 0.6) +  # Boxplot with borders and transparency
  geom_point(aes(color = Treatment, shape = Sex),  # Points mapped to Treatment and shape
             position = position_jitterdodge(dodge.width = 0.75),  # Jitter points for visibility
             alpha = 0.8, size = 2) +  # Points with defined aesthetics
   scale_fill_manual(values = c("Baseline" = "lightgreen", 
                                "PBS" = "lightblue", 
                                "Bacteroides" = "orange")) +  # Fill colors for treatments
  scale_color_manual(values = c("Baseline" = "lightgreen", 
                                 "PBS" = "lightblue", 
                                 "Bacteroides" = "orange")) +  # Color mapping for points
  theme_minimal()+
  labs(x = "Experiment Day",
       y = expression("Alpha Diversity - Pielou"),  # Y-axis label with expression
       fill = "Treatment") +  # Updated fill label
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
        legend.position = "right")  # Position the legend

# Display the plot
print(alpha_Pielou_across_experiment)

ggsave("alpha_Pielou_across_experiment_boxplot.pdf", plot = alpha_Pielou_across_experiment, device = "pdf", width = 6, height = 4)

ggsave("alpha_Observed_across_experiment_boxplot.pdf", plot = alpha_Observed_across_experiment, device = "pdf", width = 6, height = 4)

```

#Run stats for observed across experiment
```{r}


# Subset the data for Late_Baseline

Late_Baseline_observed_df <- combined_nosacrifice_filtered %>%
  filter(Compiled_Timepoint %in% c("Late_Baseline"))  # Keep rows for both time points

Late_Baseline_observed_df$Treatment <- factor(Late_Baseline_observed_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))

Late_Baseline_observed_df$Treatment <- factor(Late_Baseline_observed_df$Treatment, 
                                         levels = c("PBS", "Bacteroides", "Baseline"))

# Run the linear model on the subset
shapiro_latebaseline_shapiro <- shapiro.test(Late_Baseline_observed_df$Observed)
shapiro_latebaseline_shapiro

observed_alpha_lm_Late_Baseline <- lm(Observed ~ Treatment + Sex + Treatment*Sex+Cage, data = Late_Baseline_observed_df)
summary(observed_alpha_lm_Late_Baseline)

# Subset the data for antibiotics

Antibiotics_observed_df <- combined_nosacrifice_filtered %>%
  filter(Compiled_Timepoint %in% c("Antibiotics"))  # Keep rows for both time points

Antibiotics_observed_df$Treatment <- factor(Antibiotics_observed_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))


# Run the linear model on the subset
shapiro_antibiotics_shapiro <- shapiro.test(Antibiotics_observed_df$Observed)
shapiro_antibiotics_shapiro

observed_alpha_lm_antibiotics <- lm(Observed ~ Treatment + Sex + Treatment*Sex+Cage, data = Antibiotics_observed_df)
summary(observed_alpha_lm_antibiotics)

# Subset the data for early gavage

early_gavage_observed_df <- combined_nosacrifice_filtered %>%
  filter(Compiled_Timepoint %in% c("Early_Gavage"))  # Keep rows for both time points

early_gavage_observed_df$Treatment <- factor(early_gavage_observed_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))


# Run the linear model on the subset
shapiro_early_gavage_shapiro <- shapiro.test(early_gavage_observed_df$Observed)
shapiro_early_gavage_shapiro

observed_alpha_lm_early_gavage <- lm(Observed ~ Treatment + Sex + Treatment*Sex+Cage, data = early_gavage_observed_df)
summary(observed_alpha_lm_early_gavage)

# Subset the data for late gavage

late_gavage_observed_df <- combined_nosacrifice_filtered %>%
  filter(Compiled_Timepoint %in% c("Late_Gavage"))  # Keep rows for both time points

late_gavage_observed_df$Treatment <- factor(late_gavage_observed_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))

late_gavage_observed_df$Treatment <- factor(late_gavage_observed_df$Treatment, 
                                         levels = c("PBS", "Bacteroides", "Baseline"))

# Run the linear model on the subset
shapiro_late_gavage_shapiro <- shapiro.test(late_gavage_observed_df$Observed)
shapiro_late_gavage_shapiro

observed_alpha_lm_late_gavage <- lm(log2(Observed) ~ Treatment + Sex + Treatment*Sex+Cage, data = late_gavage_observed_df)
summary(observed_alpha_lm_late_gavage)

# Subset the data for before_LPS

before_LPS_observed_df <- combined_nosacrifice_filtered %>%
  filter(Compiled_Timepoint %in% c("Before_LPS"))  # Keep rows for both time points

before_LPS_observed_df$Treatment <- factor(before_LPS_observed_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))


# Run the linear model on the subset
shapiro_before_LPS_shapiro <- shapiro.test(before_LPS_observed_df$Observed)
shapiro_before_LPS_shapiro

observed_alpha_lm_before_LPS <- lm(log2(Observed) ~ Treatment + Sex + Treatment*Sex+Cage, data = before_LPS_observed_df)
summary(observed_alpha_lm_before_LPS)

# Subset the data for LPS

LPS_observed_df <- combined_nosacrifice_filtered %>%
  filter(Compiled_Timepoint %in% c("LPS"))  # Keep rows for both time points

LPS_observed_df$Treatment <- factor(LPS_observed_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))


# Run the linear model on the subset
shapiro_LPS_shapiro <- shapiro.test(LPS_observed_df$Observed)
shapiro_LPS_shapiro

observed_alpha_lm_LPS <- lm(Observed ~ Treatment + Sex + Treatment*Sex+Cage, data = LPS_observed_df)
summary(observed_alpha_lm_LPS)
```
#Run stats for Pielous across experiment
```{r}


# Subset the data for Late_Baseline

Late_Baseline_Pielou_df <- combined_nosacrifice_filtered %>%
  filter(Compiled_Timepoint %in% c("Late_Baseline"))  # Keep rows for both time points

Late_Baseline_Pielou_df$Treatment <- factor(Late_Baseline_Pielou_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))

Late_Baseline_Pielou_df$Treatment <- factor(Late_Baseline_Pielou_df$Treatment, 
                                         levels = c("PBS", "Bacteroides", "Baseline"))

# Run the linear model on the subset
shapiro_latebaseline_shapiro <- shapiro.test(Late_Baseline_Pielou_df$Pielou)
shapiro_latebaseline_shapiro

Pielou_alpha_lm_Late_Baseline <- lm(Pielou ~ Treatment + Sex + Treatment*Sex+Cage, data = Late_Baseline_Pielou_df)
summary(Pielou_alpha_lm_Late_Baseline)

# Subset the data for antibiotics

Antibiotics_Pielou_df <- combined_nosacrifice_filtered %>%
  filter(Compiled_Timepoint %in% c("Antibiotics"))  # Keep rows for both time points

Antibiotics_Pielou_df$Treatment <- factor(Antibiotics_Pielou_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))


# Run the linear model on the subset
shapiro_antibiotics_Pielou <- shapiro.test(Antibiotics_Pielou_df$Pielou)
shapiro_antibiotics_Pielou

Pielou_alpha_lm_antibiotics <- lm(Pielou ~ Treatment + Sex + Treatment*Sex+Cage, data = Antibiotics_Pielou_df)
summary(Pielou_alpha_lm_antibiotics)

# Subset the data for early gavage

early_gavage_Pielou_df <- combined_nosacrifice_filtered %>%
  filter(Compiled_Timepoint %in% c("Early_Gavage"))  # Keep rows for both time points

early_gavage_Pielou_df$Treatment <- factor(early_gavage_Pielou_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))


# Run the linear model on the subset
shapiro_early_gavage_Pielou <- shapiro.test(early_gavage_Pielou_df$Pielou)
shapiro_early_gavage_Pielou

Pielou_alpha_lm_early_gavage <- lm(Pielou ~ Treatment + Sex + Treatment*Sex+Cage, data = early_gavage_Pielou_df)
summary(Pielou_alpha_lm_early_gavage)

# Subset the data for late gavage

late_gavage_Pielou_df <- combined_nosacrifice_filtered %>%
  filter(Compiled_Timepoint %in% c("Late_Gavage"))  # Keep rows for both time points

late_gavage_Pielou_df$Treatment <- factor(late_gavage_Pielou_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))


# Run the linear model on the subset
shapiro_late_gavage_Pielou <- shapiro.test(late_gavage_Pielou_df$Pielou)
shapiro_late_gavage_Pielou

Pielou_alpha_lm_late_gavage <- lm(log2(Pielou) ~ Treatment + Sex + Treatment*Sex+Cage, data = late_gavage_Pielou_df)
summary(Pielou_alpha_lm_late_gavage)

# Subset the data for before_LPS

before_LPS_Pielou_df <- combined_nosacrifice_filtered %>%
  filter(Compiled_Timepoint %in% c("Before_LPS"))  # Keep rows for both time points

before_LPS_Pielou_df$Treatment <- factor(before_LPS_Pielou_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))


# Run the linear model on the subset
shapiro_before_LPS_Pielou <- shapiro.test(before_LPS_Pielou_df$Pielou)
shapiro_before_LPS_Pielou

Pielou_alpha_lm_before_LPS <- lm(log2(Pielou) ~ Treatment + Sex + Treatment*Sex+Cage, data = before_LPS_Pielou_df)
summary(Pielou_alpha_lm_before_LPS)

# Subset the data for LPS

LPS_Pielou_df <- combined_nosacrifice_filtered %>%
  filter(Compiled_Timepoint %in% c("LPS"))  # Keep rows for both time points

LPS_Pielou_df$Treatment <- factor(LPS_Pielou_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))


# Run the linear model on the subset
shapiro_LPS_Pielou <- shapiro.test(LPS_Pielou_df$Pielou)
shapiro_LPS_Pielou

Pielou_alpha_lm_LPS <- lm(Pielou ~ Treatment + Sex + Treatment*Sex+Cage, data = LPS_Pielou_df)
summary(Pielou_alpha_lm_LPS)
```

### Prepare data for relative abundance plots
```{r}
#Remove ASVs that do not show appear more than 1 time in more than 10% the samples
filteredASVs = genefilter_sample(btphylo_nosacrifice, #genefilter_sample() function is used to filter samples based on certain criteria
                                 filterfun_sample(function(x) x > 1),  #filterfun_sample() function is applied with a condition that keeps only those samples (ASVs) where the abundance is greater than 1. This will remove ASVs with low abundance across all samples.
                                 A=0.1*nsamples(Phyloseq_object_NonContam)) #A argument specifies the minimum proportion of samples in which an ASV must exceed the filtering threshold (1 in this case). An ASV must exceed the threshold (abundance > 1) in at least 10% of the samples to be retained.

# Remove samples that have zero total counts
Phyloseq_object_NonContam_pruned <- prune_samples(sample_sums(btphylo_nosacrifice) > 0, btphylo_nosacrifice)

# Prune taxa based on filtered ASVs
Phyloseq_subset_prunedst = prune_taxa(filteredASVs, Phyloseq_object_NonContam_pruned)

#Explanation: This is a common approach to reduce the noise in the dataset and improve the accuracy of downstream analysis.

#Transform to even sampling depth.
Phyloseq_subset_pruned_transformed = transform_sample_counts(Phyloseq_subset_prunedst, function(x) x/sum(x)*100) 

#Explanation: The function transform_sample_counts is being used to divide each count in a sample by the sum of all counts in that sample, and then multiply the result by 100. This effectively converts the counts to percentages, with each value representing the relative abundance of a particular ASV within its respective sample. The resulting object is called Phyloseq_object_pruned_transformed.
```


#Plot relative abundance of just Baseline 
```{r}


#Plot relative abundance with top 20 ASVs
rank_names(Phyloseq_subset_pruned_transformed)
table(tax_table(Phyloseq_subset_pruned_transformed)[,"Genus"] == "" | is.na(tax_table(Phyloseq_subset_pruned_transformed)[,"Genus"]))
#Merge all species (or in this case ASVs) that are in the same genus
dat.aglo = tax_glom(Phyloseq_subset_prunedst, taxrank = "Genus")
dat_baseline <- subset_samples(dat.aglo, Compiled_Timepoint == "Baseline")


#create list of top 20 ASVs in the phyloseq object that has merged ASVs based on Genus
top10 <- names(sort(taxa_sums(dat_baseline), decreasing = TRUE)[1:10])
top10


#This will prune out all taxa except those top 20
prune.top10 = prune_taxa(top10, dat.aglo)

ntaxa(prune.top10)

prune.top10 = prune_taxa(top10, dat.aglo)

ntaxa(prune.top10)

#Re normalize the top 20 taxa to be out of 100
dat.trans10 = transform_sample_counts(prune.top10, function(x) x/sum(x)*100)


#Create a dataframe that has melted or merged the normalized relative abundances of the top 20 taxa so that we can plot using ggplot2
dat.dataframe10 = psmelt(dat.trans10)
dat_baseline <- dat.dataframe10 %>%
  filter(Compiled_Timepoint == "Baseline")

#Average out the relative abundances based on Treatment Short Label and create new, simpler data frame
dat.agr10 = aggregate(Abundance~Timepoint+Sample+Genus, data=dat_baseline, FUN=mean)


#Plot relative abundance of top 20 genera in the gut, normalized for each sample
relative_abundance_treatment <- ggplot(dat.agr10, aes(x=Sample, y=Abundance, fill=Genus)) + geom_bar(stat="identity")+labs(title = "Relative Abundance across Experiment", x="Treatment", y="Relative Abundance %")+theme(text = element_text(size = 15))+
  theme(axis.text.x = element_text(angle = 45, hjust=1))

relative_abundance_treatment
```
### Plot relative abundance of just Bacteroides
```{r}
#Use your pruned and transformed phyloseq object
Phyloseq_subset_pruned_transformed

ps_bacteroides <- subset_taxa(Phyloseq_subset_pruned_transformed, Genus == "Bacteroides")

otu_df <- as.data.frame(otu_table(ps_bacteroides))
bacteroides_df <- data.frame(Bacteroides_ra = colSums(otu_df[, sapply(otu_df, is.numeric)]))

bacteroides_df$SampleID <- rownames(bacteroides_df)
metadata_all$SampleID <- rownames(metadata_all)

# Merge the two dataframes on the "SampleID" column
metadata_bt_df <- merge(bacteroides_df, metadata_all, by = "SampleID")


#Plot relative abundance across experiment
ra_Bacteroides_across_timeline_smooth <- ggplot(metadata_bt_df, aes(x = Experiment_Day, y = Bacteroides_ra, color = Treatment, shape = Sex)) +
  geom_point() +
  stat_summary(fun.y = median, size=2, geom = "line", aes(group = Treatment)) +
  geom_point(size=1.5)+
  theme_bw()+
  geom_vline(xintercept = c(0, -25, 12), linetype = "dashed", color = "blue", size = 0.5) +
  geom_text(aes(x = -25, y = Inf, label = "ABX"), vjust = 2, color = "red", hjust = 0, size = 3) +
  geom_text(aes(x = 0, y = Inf, label = "Gavage"), vjust = 2, color = "red", hjust = 0, size = 3) +
  geom_text(aes(x = 12, y = Inf, label = "LPS"), vjust = 2, color = "red", hjust = 0, size = 3) +
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  geom_line(aes(group = interaction(SeqCoastOrder,Cage,Mouse)), size = 0.25, alpha=0.3)  # Add this line to connect points for each MouseID


ra_Bacteroides_across_timeline_smooth


ggsave("Fig1B_ra_Bacteroides_across_timeline_smooth.pdf", plot = ra_Bacteroides_across_timeline_smooth, device = "pdf", width = 6, height = 3)


#getting help with coding from claude.ai
ra_Bacteroides_across_timeline_smooth_claude <- ggplot(metadata_bt_df, 
    aes(x = Experiment_Day, y = Bacteroides_ra, color = Treatment, fill = Treatment)) +
  geom_point(aes(shape = Sex), size = 1.5, stroke = 0.5) +
  geom_point(data = subset(metadata_bt_df, Sex == "M"),  # black outline for males only
             aes(shape = Sex, fill = Treatment),
             size = 1.5, color = "black", stroke = 0.5) +
  stat_summary(fun.y = median, size = 2, geom = "line", aes(group = Treatment)) +
  theme_bw() +
  geom_vline(xintercept = c(0, -25, 12), linetype = "dashed", color = "blue", size = 0.5) +
  geom_text(aes(x = -25, y = Inf, label = "ABX"), vjust = 1, color = "red", hjust = 0, size = 3) +
  geom_text(aes(x = 0, y = Inf, label = "Gavage"), vjust = 1, color = "red", hjust = 0, size = 3) +
  geom_text(aes(x = 12, y = Inf, label = "LPS"), vjust = 1, color = "red", hjust = 0, size = 3) +
  scale_shape_manual(values = c("F" = 21, "M" = 24)) +
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  scale_fill_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  geom_line(aes(group = interaction(SeqCoastOrder, Cage, Mouse)), size = 0.25, alpha = 0.3)+
    labs(
    x = "Experiment Day",
    y = expression(paste("Relative Abundance of", "\n", italic("Bacteroides")))
  )+  
  theme_bw() +
  theme(
    axis.title.x = element_text(size = 14),   # X‑axis title
    axis.title.y = element_text(size = 14),   # Y‑axis title
    axis.text   = element_text(size = 12)    # tick labels
  )


ra_Bacteroides_across_timeline_smooth_claude

ggsave("Fig1B_ra_Bacteroides_across_timeline_smooth_PR.pdf", plot = ra_Bacteroides_across_timeline_smooth_claude, device = "pdf", width = 6, height = 3)


#Plot relative abundance across experiment, separated by sex
ra_Bacteroides_across_timeline_smooth <- ggplot(metadata_bt_df, aes(x = Experiment_Day, y = Bacteroides_ra, color = Treatment, shape = Sex)) +geom_point() +
  stat_summary(fun.y = median, size=2, geom = "line", aes(group = Treatment)) +
  geom_point(size=1.5)+
  theme_bw()+
  facet_wrap(~Sex)+
  geom_vline(xintercept = c(0, -25, 12), linetype = "dashed", color = "blue", size = 0.5) +
  geom_text(aes(x = -25, y = Inf, label = "ABX"), vjust = 2, color = "red", hjust = 0, size = 3) +
  geom_text(aes(x = 0, y = Inf, label = "Gavage"), vjust = 2, color = "red", hjust = 0, size = 3) +
  geom_text(aes(x = 12, y = Inf, label = "LPS"), vjust = 2, color = "red", hjust = 0, size = 3) +
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  geom_line(aes(group = interaction(SeqCoastOrder,Cage,Mouse)), size = 0.25) + # Add this line to connect points for each MouseID


ra_Bacteroides_across_timeline_smooth



#linear model on relative abundance of Bacteroides at the end of experiment
# Subset metadata_bt_df to only include rows with Timepoint == "LPS"
metadata_bt_lps <- subset(metadata_bt_df, Compiled_Timepoint == "Before_LPS")

#Relevel the Treatment factor to make "PBS" the reference level
metadata_bt_lps$Treatment <- factor(metadata_bt_lps$Treatment, levels = c("PBS", "Bacteroides"))


# Run a linear model
lm_model <- lm(Bacteroides_ra ~ Treatment+Sex+Treatment*Sex+SeqCoastOrder:Cage, data = metadata_bt_lps)

# Display the summary of the linear model
summary(lm_model)

bacteroides_sex_plot <- ggplot(metadata_bt_lps, aes(x = Sex, y = Bacteroides_ra)) +
  geom_boxplot(outlier.shape = NA, fill = "lightblue", alpha = 0.5) +  # Boxplot without outlier points
  geom_jitter(width = 0.2, size = 2, alpha = 0.6, aes(color = Sex)) +  # Jittered points for individual values
  facet_wrap(~Treatment)+
  theme_bw() +
  labs(
    x = "Sex",
    y = "Bacteroides Relative Abundance",
    title = "Bacteroides Relative Abundance by Sex"
  ) +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.position = "none"
  )

# Display the plot
bacteroides_sex_plot

ggsave("bacteroides_sex_plot.pdf", plot = bacteroides_sex_plot, device = "pdf", width = 6, height = 3)

```

#Plot Escherichia-Shigella across the experiment
```{r}
#Transform to even sampling depth.
btphylo_nosacrifice_transformed = transform_sample_counts(btphylo_nosacrifice, function(x) x/sum(x)*100) 

ps_escherichia <- subset_taxa(btphylo_nosacrifice_transformed, Genus == "Escherichia-Shigella")


otu_df_es <- as.data.frame(otu_table(ps_escherichia))
escherichia_df <- data.frame(Escherichia_Shigella_ra = colSums(otu_df_es[, sapply(otu_df_es, is.numeric)]))

escherichia_df$SampleID <- rownames(escherichia_df)
metadata_all$SampleID <- rownames(metadata_all)

# Merge the two dataframes on the "SampleID" column
metadata_es_df <- merge(escherichia_df, metadata_all, by = "SampleID")


#Plot relative abundance across experiment
ra_Escherichia_Shigella_across_timeline_smooth <- ggplot(metadata_es_df, aes(x = Experiment_Day, y = Escherichia_Shigella_ra, color = Treatment, shape = Sex)) +geom_point() +
  stat_summary(fun.y = median, size=2, geom = "line", aes(group = Treatment)) +
  geom_point(size=1.5)+
  theme_bw()+
  geom_vline(xintercept = c(0, -25, 12), linetype = "dashed", color = "blue", size = 0.5) +
  geom_text(aes(x = -25, y = Inf, label = "ABX"), vjust = 2, color = "red", hjust = 0, size = 3) +
  geom_text(aes(x = 0, y = Inf, label = "Gavage"), vjust = 2, color = "red", hjust = 0, size = 3) +
  geom_text(aes(x = 12, y = Inf, label = "LPS"), vjust = 2, color = "red", hjust = 0, size = 3) +
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  geom_line(aes(group = interaction(SeqCoastOrder,Cage,Mouse)), size = 0.25)  # Add this line to connect points for each MouseID



ra_Escherichia_Shigella_across_timeline_smooth

ggsave("FigS2B_ra_Escherichia_across_timeline_smooth.pdf", plot = ra_Escherichia_Shigella_across_timeline_smooth, device = "pdf", width = 6, height = 3)


unique(metadata_es_LPS_df$Compiled_Timepoint)

metadata_es_LPS_df <- metadata_es_df %>% filter(Treatment %in% c("PBS", "Bacteroides"))
metadata_es_LPS_df_s <- metadata_es_LPS_df %>%  filter(Compiled_Timepoint %in% c("Early_Gavage", "Late_Gavage", "Before_LPS", "LPS"))

unique(metadata_es_LPS_df$Compiled_Timepoint)


ecoli_across_experiment_lm <- lm(Escherichia_Shigella_ra ~ Treatment + Sex + Sex*Treatment+SeqCoastOrder:Cage, data = metadata_es_LPS_df_s)

summary(ecoli_across_experiment_lm)

```

``
#plot other top taxa across experiment
```{r}

#Remove ASVs that do not show appear more than 1 time in more than 10% the samples
filteredASVs_all = genefilter_sample(btphylo_nosacrifice, #genefilter_sample() function is used to filter samples based on certain criteria
                                 filterfun_sample(function(x) x > 1))  #filterfun_sample() function is applied with a condition that keeps only those samples (ASVs) where the abundance is greater than 1. This will remove ASVs with low abundance across all samples.
                                 #A=0.1*nsamples(Phyloseq_object_NonContam)) #A argument specifies the minimum proportion of samples in which an ASV must exceed the filtering threshold (1 in this case). An ASV must exceed the threshold (abundance > 1) in at least 10% of the samples to be retained.

# Remove samples that have zero total counts
btphylo_nosacrifice_pruned <- prune_samples(sample_sums(btphylo_nosacrifice) > 0, btphylo_nosacrifice)

# Prune taxa based on filtered ASVs
btphylo_nosacrifice_prunedst = prune_taxa(filteredASVs_all, btphylo_nosacrifice_pruned)

#Explanation: This is a common approach to reduce the noise in the dataset and improve the accuracy of downstream analysis.

#Transform to even sampling depth.
btphylo_nosacrifice_prunedst_transformed = transform_sample_counts(btphylo_nosacrifice_prunedst, function(x) x/sum(x)*100) 

#Explanation: The function transform_sample_counts is being used to divide each count in a sample by the sum of all counts in that sample, and then multiply the result by 100. This effectively converts the counts to percentages, with each value representing the relative abundance of a particular ASV within its respective sample. The resulting object is called Phyloseq_object_pruned_transformed.

genus_rel_df <- psmelt(btphylo_nosacrifice_prunedst_transformed)
genus_rel_df_notransformed <- psmelt(btphylo_nosacrifice_prunedst)


# Summarize the abundance for each unique genus within each sample
collapsed_df <- genus_rel_df_notransformed %>%
  group_by(Sample, Genus) %>%
  summarize(Abundance = sum(Abundance, na.rm = TRUE)) %>%
  ungroup()

# Filter out genera that do not have an abundance > 0 in any sample
collapsed_df_filtered <- collapsed_df %>%
  group_by(Genus) %>%
  filter(any(Abundance > 0)) %>%
  ungroup()

# Merge the filtered collapsed dataframe with metadata_all
merged_df <- collapsed_df_filtered %>%
  left_join(metadata_all, by = c("Sample" = "SampleID"))

# Calculate relative abundance for each genus within each sample as a percentage


# Identify the top 10 genera based on total abundance across all samples
top_10_genera <- merged_df %>%
  group_by(Genus) %>%
  summarize(Total_Abundance = sum(Abundance, na.rm = TRUE)) %>%
  arrange(desc(Total_Abundance)) %>%
  slice(1:10) %>%
  pull(Genus)

print(top_10_genera)

# Relabel the "Genus" column to keep only top 10 genera, others as "Other"
merged_df <- merged_df %>%
  mutate(Genus = ifelse(Genus %in% top_10_genera, Genus, "Other"))

# Define a custom color palette for the genera, ensuring "Bacteroides" is orange
genus_colors <- c(
  "Bacteroides" = "orange",  # Set Bacteroides to orange
  "Burkholderia-Caballeronia-Paraburkholderia" = "#56B4E9",
  "Turicibacter" = "#D55E50",
  "Enterococcus" = "#9467bd",
  "Dubosiella" = "#CC79A7",
  "Ralstonia" = "#009E73",
  "NA" = "#7f7f7f",
  "Lactobacillus" = "#17becf",
  "Acinetobacter" = "#FFDB58",
  "Clostridium sensu stricto 1" = "#8B4513",
  "Other" = "lightgray"
)

# Reorder the Treatment factor to ensure desired facet order
merged_df$Treatment <- factor(merged_df$Treatment, levels = c("Baseline", "PBS", "Bacteroides"))

# Filter out the Baseline treatment group
merged_df2 <- merged_df %>%
  filter(Treatment != "Baseline")

# Plot relative abundance across the timeline
ra_all_genera_across_timeline <- ggplot(merged_df2, aes(x = Experiment_Day, y = Abundance, color = Genus)) +
  stat_summary(fun = mean, size = 2, geom = "line", aes(group = Genus)) +
  theme_bw() +
  facet_grid(Treatment~., scales = "free_y") +
  geom_vline(xintercept = c(0, -25, 12), linetype = "dashed", color = "blue", size = 0.5) +
  geom_text(aes(x = -25, y = Inf, label = "ABX"), vjust = 2, color = "red", hjust = 0, size = 3) +
  geom_text(aes(x = 0, y = Inf, label = "Gavage"), vjust = 2, color = "red", hjust = 0, size = 3) +
  geom_text(aes(x = 12, y = Inf, label = "LPS"), vjust = 2, color = "red", hjust = 0, size = 3) +
  labs(
    x = "Experiment Day",
    y = "Relative Abundance (%)",
    title = "Relative Abundance of Genera Across Timeline",
    color = "Genus"
  ) +
  scale_color_manual(values = genus_colors) +  # Apply custom colors
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

# Display the plot
ra_all_genera_across_timeline

ggsave("ra_all_genera_across_timeline.pdf", plot = ra_all_genera_across_timeline, device = "pdf", width = 12, height = 6)


```



###Make plots to show where in the gut Bacteroides is establishing
```{r}
#Relative abundance of Bacteroides across the gut

#filter phyloseq object so that only the samples collected at sacrifice are included
btphylo_atsacrifice <- subset_samples(Phyloseq_object_NonContam, Compiled_Timepoint == "Sacrifice")

# Remove ASVs that do not appear more than once in at least 10% of the samples
filteredASVs_atsac = genefilter_sample(btphylo_atsacrifice,  
                                       filterfun_sample(function(x) x > 1),  
                                       A = 0.1 * nsamples(btphylo_atsacrifice))  

# Remove samples that have zero total counts  
btphylo_atsac_pruned <- prune_samples(sample_sums(btphylo_atsacrifice) > 0, btphylo_atsacrifice)  

# Prune taxa based on filtered ASVs  
btphylo_atsac_pruned_subset = prune_taxa(filteredASVs_atsac, btphylo_atsac_pruned)  

# Transform to even sampling depth (relative abundance percentage)  
btphylo_atsac_transformed = transform_sample_counts(btphylo_atsac_pruned_subset, function(x) x / sum(x) * 100)  

# Use pruned and transformed phyloseq object  
ps_bacteroides_atsac <- subset_taxa(btphylo_atsac_transformed, Genus == "Bacteroides")  


otu_df_atsac <- as.data.frame(otu_table(ps_bacteroides_atsac))  
bacteroides_df_atsac <- data.frame(Bacteroides_ra = colSums(otu_df_atsac[, sapply(otu_df_atsac, is.numeric)]))  

bacteroides_df_atsac$SampleID <- rownames(bacteroides_df_atsac)  
metadata_all$SampleID <- rownames(metadata_all)  

# Merge the two dataframes on the "SampleID" column  
metadata_bt_df_atsac <- merge(bacteroides_df_atsac, metadata_all, by = "SampleID")  

# Set factor levels for Treatment to ensure "PBS" comes before "Bacteroides"
metadata_bt_df_atsac$Treatment <- factor(metadata_bt_df_atsac$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))

# Set factor levels for Biogeography in the desired order
metadata_bt_df_atsac$Biogeography <- factor(metadata_bt_df_atsac$Biogeography, 
                                            levels = c("SmallIntestine", "Cecum", "ProximalColon", "DistalColon"))


#Plot relative abundance across biogeography
ra_Bacteroides_across_gut <- ggplot(metadata_bt_df_atsac, aes(x = Biogeography, y = Bacteroides_ra, fill = Treatment)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +  # Boxplot with transparent fill
  geom_point(aes(color = Treatment, shape = Sex),  # Map shape to Sex within geom_point
             position = position_jitterdodge(dodge.width = 0.5), 
             alpha = 0.8, size = 2) +  # Points with color and shape mapped
  scale_fill_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +  # Corrected scale_fill_manual
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +  # Corrected scale_color_manual
  theme_minimal() +
  labs(x = "Biogeography",
       y = expression("Relative Abundance of " * italic("Bacteroides")),  # Apply italics to "Bacteroides"
       fill = "Treatment", 
       color = "Treatment",  # Add a label for color
       shape = "Sex") +  # Add a label for shape
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels if needed

ra_Bacteroides_across_gut

ggsave("ra_Bacteroides_across_gut.pdf", plot = ra_Bacteroides_across_gut, device = "pdf", width = 6, height = 4)

#outline the males with help from claude
ra_Bacteroides_across_gut_claude <- ggplot(metadata_bt_df_atsac, aes(x = Biogeography, y = Bacteroides_ra, fill = Treatment)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_point(data = subset(metadata_bt_df_atsac, Sex == "F"),
             aes(fill = Treatment, shape = Sex),
             position = position_jitterdodge(dodge.width = 0.5),
             color = "grey50", size = 2, alpha = 0.8, stroke = 0.3) +
  geom_point(data = subset(metadata_bt_df_atsac, Sex == "M"),
             aes(fill = Treatment, shape = Sex),
             position = position_jitterdodge(dodge.width = 0.5),
             color = "black", size = 2, alpha = 0.8, stroke = 0.8) +
  scale_shape_manual(values = c("F" = 21, "M" = 24)) +
  scale_fill_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange"), guide = "none") +
  theme_minimal() +
  labs(x = "Biogeography",
       y = expression("Relative Abundance of " * italic("Bacteroides")),
       fill = "Treatment",
       shape = "Sex") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ra_Bacteroides_across_gut_claude
ggsave("ra_Bacteroides_across_gut_PR.pdf", plot = ra_Bacteroides_across_gut_claude, device = "pdf", width = 6, height = 4)

# Subset the data to include only Bacteroides treatment
metadata_bt_df_atsac_bt <- subset(metadata_bt_df_atsac, Treatment == "Bacteroides")


# Run the linear model
metadata_bt_df_atsac_bt_lm <- lm(Bacteroides_ra ~ Biogeography+Sex+SeqCoastOrder:Cage, data = metadata_bt_df_atsac_bt)

# Display the summary of the linear model
summary(metadata_bt_df_atsac_bt_lm)

# Subset the data to include only Bacteroides treatment
metadata_bt_df_atsac_PBS <- subset(metadata_bt_df_atsac, Treatment == "Bacteroides")

# Run the linear model
metadata_bt_df_atsac_PBS_lm <- lm(Bacteroides_ra ~ Biogeography+Sex+SeqCoastOrder:Cage, data = metadata_bt_df_atsac_PBS)

# Display the summary of the linear model
summary(metadata_bt_df_atsac_PBS_lm)

#Subset samples for microniche data
metadata_bt_df_atsac_mironiche <- metadata_bt_df_atsac %>%
  filter(!is.na(Microniche) & Microniche != "")

#Plot relative abundance across microniche
ra_Bacteroides_across_microniche <- ggplot(metadata_bt_df_atsac_mironiche, aes(x = Microniche, y = Bacteroides_ra, fill = Treatment, shape=Sex)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +  # Boxplot with transparent fill
  geom_point(aes(color = Treatment), position = position_jitterdodge(dodge.width = 0.5), alpha = 0.8, size = 2) +  # Points with color mapped to Treatment
  scale_fill_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +  # Corrected scale_fill_manual
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +  # Corrected scale_color_manual
  theme_minimal() +
  labs(x = "Microniche", 
       y = expression("Relative Abundance of " * italic("Bacteroides")),  # Apply italics to "Bacteroides"
       fill = "Treatment") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels if needed

ra_Bacteroides_across_microniche

ggsave("ra_Bacteroides_across_microniche.pdf", plot = ra_Bacteroides_across_microniche, device = "pdf", width = 6, height = 4)

#Plot relative abundance across microniche
ra_Bacteroides_across_microniche_bysex <- ggplot(metadata_bt_df_atsac_mironiche, aes(x = Microniche, y = Bacteroides_ra, fill = Treatment)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +  # Boxplot with transparent fill
  geom_point(aes(color = Treatment, shape = Sex),  # Map shape to Sex within geom_point
             position = position_jitterdodge(dodge.width = 0.5), 
             alpha = 0.8, size = 2) +  # Points with color and shape mapped
  scale_fill_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +  # Corrected scale_fill_manual
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +  # Corrected scale_color_manual
  theme_minimal() +
  labs(x = "Microniche",
       y = expression("Relative Abundance of " * italic("Bacteroides")),  # Apply italics to "Bacteroides"
       fill = "Treatment", 
       color = "Treatment",  # Add a label for color
       shape = "Sex") +  # Add a label for shape
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels if needed

ra_Bacteroides_across_microniche_bysex

ggsave("ra_Bacteroides_across_microniche_bysex.pdf", plot = ra_Bacteroides_across_microniche_bysex, device = "pdf", width = 6, height = 4)


# Subset the data to include only Bacteroides treatment

# Set factor levels for Biogeography in the desired order
metadata_bt_df_atsac_mironiche$Biogeography <- factor(metadata_bt_df_atsac_mironiche$Biogeography, 
                                            levels = c("SmallIntestine", "Cecum", "ProximalColon", "DistalColon"))


metadata_bt_df_atsac_microniche_bt <- subset(metadata_bt_df_atsac_mironiche, Treatment == "Bacteroides")

# Run the linear model
metadata_bt_df_atsac_microniche_Bt_lm <- lm(Bacteroides_ra ~ Microniche+Biogeography+Sex, data = metadata_bt_df_atsac_microniche_bt)

# Display the summary of the linear model
summary(metadata_bt_df_atsac_microniche_Bt_lm)


#outline the male triangles in black
ra_Bacteroides_across_microniche_claude <- ggplot(metadata_bt_df_atsac_mironiche, aes(x = Microniche, y = Bacteroides_ra, fill = Treatment)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_point(data = subset(metadata_bt_df_atsac_mironiche, Sex == "F"),
             aes(fill = Treatment, shape = Sex),
             position = position_jitterdodge(dodge.width = 0.5),
             color = "grey50", size = 2, alpha = 0.8, stroke = 0.3) +
  geom_point(data = subset(metadata_bt_df_atsac_mironiche, Sex == "M"),
             aes(fill = Treatment, shape = Sex),
             position = position_jitterdodge(dodge.width = 0.5),
             color = "black", size = 2, alpha = 0.8, stroke = 0.8) +
  scale_shape_manual(values = c("F" = 21, "M" = 24)) +
  scale_fill_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange"), guide = "none") +
  theme_minimal() +
  labs(x = "Microniche",
       y = expression("Relative Abundance of " * italic("Bacteroides")),
       fill = "Treatment",
       shape = "Sex") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ra_Bacteroides_across_microniche_claude

ggsave("ra_Bacteroides_across_microniche_PR.pdf", plot = ra_Bacteroides_across_microniche_claude, device = "pdf", width = 6, height = 4)
```

#plot relative abundance of Escherichia across the gut
```{r}

ps_Escherichia_Shigella_atsac <- subset_taxa(btphylo_atsac_transformed, Genus == "Escherichia-Shigella")  

otu_df_atsac_es <- as.data.frame(otu_table(ps_Escherichia_Shigella_atsac))  
Escherichia_Shigella_df_atsac <- data.frame(Escherichia_Shigella_ra = colSums(otu_df_atsac_es[, sapply(otu_df_atsac_es, is.numeric)]))  

Escherichia_Shigella_df_atsac$SampleID <- rownames(Escherichia_Shigella_df_atsac)  
metadata_all$SampleID <- rownames(metadata_all)  

# Merge the two dataframes on the "SampleID" column  
metadata_bt_es_atsac <- merge(Escherichia_Shigella_df_atsac, metadata_all, by = "SampleID")  

# Set factor levels for Treatment to ensure "PBS" comes before "Bacteroides"
metadata_bt_es_atsac$Treatment <- factor(metadata_bt_es_atsac$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))

# Set factor levels for Biogeography in the desired order
metadata_bt_es_atsac$Biogeography <- factor(metadata_bt_es_atsac$Biogeography, 
                                            levels = c("SmallIntestine", "Cecum", "ProximalColon", "DistalColon"))


#Plot relative abundance across biogeography
ra_escherichia_across_gut <- ggplot(metadata_bt_es_atsac, aes(x = Biogeography, y = Escherichia_Shigella_ra, fill = Treatment)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +  # Boxplot with transparent fill
  geom_point(aes(color = Treatment, shape = Sex),  # Map shape to Sex within geom_point
             position = position_jitterdodge(dodge.width = 0.5), 
             alpha = 0.8, size = 2) +  # Points with color and shape mapped
  scale_fill_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +  # Corrected scale_fill_manual
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +  # Corrected scale_color_manual
  theme_minimal() +
  labs(x = "Biogeography",
       y = expression("Relative Abundance of " * italic("Escherichia-Shigella")),  # Apply italics to "Escherichia-Shigella"
       fill = "Treatment", 
       color = "Treatment",  # Add a label for color
       shape = "Sex") +  # Add a label for shape
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels if needed

ra_escherichia_across_gut

ggsave("ra_escherichia_across_gut.pdf", plot = ra_escherichia_across_gut, device = "pdf", width = 6, height = 4)


#running stats to see if there's a significant difference between E coli relative abundance at any region of the intestines

metadata_bt_es_atsac_pvb <- metadata_bt_es_atsac %>%
  filter(Treatment %in% c("PBS", "Bacteroides"))

#subset small intestine

metadata_bt_es_atsac_pvb_SI <- subset(metadata_bt_es_atsac_pvb, Biogeography == "SmallIntestine")

metadata_bt_es_atsac_pvb_SI_lm <- lm(Escherichia_Shigella_ra ~Treatment+Sex+Sex*Treatment, data=metadata_bt_es_atsac_pvb_SI)

summary(metadata_bt_es_atsac_pvb_SI_lm)

#subset cecum

metadata_bt_es_atsac_pvb_cec <- subset(metadata_bt_es_atsac_pvb, Biogeography == "Cecum")

metadata_bt_es_atsac_pvb_cec_lm <- lm(Escherichia_Shigella_ra ~Treatment+Sex+Sex*Treatment, data=metadata_bt_es_atsac_pvb_cec)

summary(metadata_bt_es_atsac_pvb_cec_lm)

#subset proximal colon

metadata_bt_es_atsac_pvb_pro <- subset(metadata_bt_es_atsac_pvb, Biogeography == "ProximalColon")

metadata_bt_es_atsac_pvb_pro_lm <- lm(Escherichia_Shigella_ra ~Treatment+Sex+Sex*Treatment, data=metadata_bt_es_atsac_pvb_pro)

summary(metadata_bt_es_atsac_pvb_pro_lm)

#subset distal colon

metadata_bt_es_atsac_pvb_distal <- subset(metadata_bt_es_atsac_pvb, Biogeography == "DistalColon")

metadata_bt_es_atsac_pvb_dis_lm <- lm(Escherichia_Shigella_ra ~Treatment+Sex+Sex*Treatment, data=metadata_bt_es_atsac_pvb_distal)

summary(metadata_bt_es_atsac_pvb_dis_lm)

```


```{r}
# Subset the data to include only Bacteroides treatment
metadata_bt_es_atsac_bt <- subset(metadata_bt_es_atsac, Treatment == "Bacteroides")


# Run the linear model
metadata_bt_es_atsac_bt_lm <- lm(Escherichia_Shigella_ra ~ Biogeography+Sex+Microniche, data = metadata_bt_es_atsac_bt)

# Display the summary of the linear model
summary(metadata_bt_es_atsac_bt_lm)

# Subset the data to include only Bacteroides treatment
metadata_bt_es_atsac_PBS <- subset(metadata_bt_es_atsac, Treatment == "PBS")

# Run the linear model
metadata_bt_es_atsac_PBS_lm <- lm(Escherichia_Shigella_ra ~ Biogeography+Sex, data = metadata_bt_es_atsac_PBS)

# Display the summary of the linear model
summary(metadata_bt_es_atsac_PBS_lm)

#Subset samples for microniche data
metadata_bt_es_atsac_mironiche <- metadata_bt_es_atsac %>%
  filter(!is.na(Microniche) & Microniche != "")

#Plot relative abundance across microniche
ra_escherichia_across_microniche <- ggplot(metadata_bt_es_atsac_mironiche, aes(x = Microniche, y = Escherichia_Shigella_ra, fill = Treatment)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +  # Boxplot with transparent fill, outliers hidden
  geom_point(aes(color = Treatment, shape = Sex),  # Map shape to Sex and color to Treatment within geom_point
             position = position_jitterdodge(dodge.width = 0.5),
             alpha = 0.8, size = 2) +  # Points with color and shape mapped
  scale_fill_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +  # Treatment fill colors
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +  # Treatment color mapping
  theme_minimal() +
  labs(x = "Microniche",
       y = expression("Relative Abundance of " * italic("Escherichia-Shigella")),  # Apply italics to "Escherichia-Shigella"
       fill = "Treatment",
       color = "Treatment",  # Add a label for color mapping
       shape = "Sex") +  # Add a label for shape mapping
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels if needed

# Print the plot
print(ra_escherichia_across_microniche)

ggsave("ra_escherichia_across_microniche.pdf", plot = ra_escherichia_across_microniche, device = "pdf", width = 6, height = 4)



# Subset the data to include only Bacteroides treatment

# Set factor levels for Biogeography in the desired order
metadata_bt_es_atsac_mironiche$Biogeography <- factor(metadata_bt_es_atsac_mironiche$Biogeography, 
                                            levels = c("SmallIntestine", "Cecum", "ProximalColon", "DistalColon"))


metadata_bt_es_atsac_microniche_bt <- subset(metadata_bt_es_atsac_mironiche, Treatment == "Bacteroides")

# Run the linear model
metadata_bt_es_atsac_microniche_bt_lm <- lm(Escherichia_Shigella_ra ~ Microniche+Biogeography+Sex, data = metadata_bt_es_atsac_microniche_bt)

# Display the summary of the linear model
summary(metadata_bt_es_atsac_microniche_bt_lm)

#running stats to see if there's a significant difference between E coli relative abundance at any microniche of the intestines

metadata_bt_es_atsac_microniche_bt_pvb <- metadata_bt_es_atsac_mironiche %>%
  filter(Treatment %in% c("PBS", "Bacteroides"))

#subset small luminal contents

metadata_bt_es_atsac_microniche_bt_pvb_lumen <- subset(metadata_bt_es_atsac_microniche_bt_pvb, Microniche == "LuminalContents")

metadata_bt_es_atsac_microniche_bt_pvb_lumen_lm <- lm(Escherichia_Shigella_ra ~Treatment+Sex+Sex*Treatment, data=metadata_bt_es_atsac_microniche_bt_pvb_lumen)

summary(metadata_bt_es_atsac_microniche_bt_pvb_lumen_lm)

#subset small mucosalsurface

metadata_bt_es_atsac_microniche_bt_pvb_mucosa <- subset(metadata_bt_es_atsac_microniche_bt_pvb, Microniche == "MucosalSurface")

metadata_bt_es_atsac_microniche_bt_pvb_lumen_mucosa <- lm(Escherichia_Shigella_ra ~Treatment+Sex+Sex*Treatment, data=metadata_bt_es_atsac_microniche_bt_pvb_mucosa)

summary(metadata_bt_es_atsac_microniche_bt_pvb_lumen_mucosa)

```



#Plot beta diversity on communities between lumen and mucosal surface. We want to see how communities compare between regions for PBS, Bacteroides, and Baseline groups

#Subset data and label the non dissection samples as stool. 
# I want to see whether region collected drives differences in microbial communities
#First analyze just the PBS 
```{r}
#Merge a late baseline collection from the baseline treatment group to be able to plot stool from that group
#Data from samples collected across gut at sacrifice
btphylo_baseline_stool <- subset_samples(Phyloseq_object_NonContam, 
                                      Treatment == "Baseline" & Compiled_Timepoint == "Late_Baseline")

# Merge the two phyloseq objects
phylo_microniche_comparison <- merge_phyloseq(btphylo_baseline_stool, btphylo_atsacrifice)



# Extract sample data as a dataframe
sample_data_df <- as.data.frame(sample_data(phylo_microniche_comparison))

# Replace NA values in the Microniche column with "stool"
sample_data_df$Microniche[is.na(sample_data_df$Microniche)] <- "stool"

# Assign the modified sample data back to the phyloseq object
sample_data(phylo_microniche_comparison) <- sample_data(sample_data_df)


#Remove ASVs that do not show appear more than 1 time in more than 10% the samples
filteredASVs = genefilter_sample(phylo_microniche_comparison, 
                                 filterfun_sample(function(x) x > 1), 
                                 A=0.2*nsamples(phylo_microniche_comparison))


# Remove samples that have zero total counts
btphylo_atsacrifice_pruned <- prune_samples(sample_sums(phylo_microniche_comparison) > 0, phylo_microniche_comparison)


# Prune taxa based on filtered ASVs
btphylo_atsacrifice_pruned = prune_taxa(filteredASVs, btphylo_atsacrifice_pruned)

#Transform to even sampling depth.
btphylo_atsacrifice_pruned_transformed = transform_sample_counts(btphylo_atsacrifice_pruned, function(x) x/sum(x)*100) 

```


### Calculate distance and plot PCoA across the gut
```{r}
# Troubleshooting: Check for NA values in the OTU table
sum(is.na(otu_table(btphylo_atsacrifice_pruned_transformed)))

# Check for rows (samples) with only zeros
zero_rows <- rowSums(otu_table(btphylo_atsacrifice_pruned_transformed)) == 0
sum(zero_rows)  # Count of empty samples

# Replace NAs with 0s
otu_table(btphylo_atsacrifice_pruned_transformed)[is.na(otu_table(btphylo_atsacrifice_pruned_transformed))] <- 0

# Calculate Bray-Curtis dissimilarity distance matrix
distance_bray <- phyloseq::distance(btphylo_atsacrifice_pruned_transformed, method = "bray")

# Perform a principal coordinate analysis (PCoA) on the distance matrix
ordinate_PCoA_bray <- ordinate(btphylo_atsacrifice_pruned_transformed, method = "PCoA", distance = distance_bray)

# Define a pastel color palette for Microniche
pastel_colors <- c(
  "LuminalContents" = "#8FCACA",  # Pastel Green
  "MucosalSurface" = "#CCE2CB",   # Pastel Blue
  "stool" = "#CBAACB"             # Pastel Purple
)

# Plot the results with points colored by the Treatment group
PCoA_plot_all_biogeography <- plot_ordination(btphylo_atsacrifice_pruned_transformed, 
                                              ordinate_PCoA_bray, 
                                              color = "Treatment",,shape="Sex") +
  stat_ellipse(aes(group=Treatment)) +  # Adds an ellipse for visualization
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +  
  facet_wrap(~Biogeography, ncol = 5)  # Facet into 5 columns

# Print the plot
PCoA_plot_all_biogeography


# Save the PCoA plot as a PDF
ggsave("PCoA_plot_biogeography.pdf", plot = PCoA_plot_all_biogeography, width = 8, height = 3, dpi = 300)

PCoA_plot_all_microniche <- plot_ordination(btphylo_atsacrifice_pruned_transformed, 
                                              ordinate_PCoA_bray, 
                                              color = "Treatment",,shape="Sex") +
  stat_ellipse(aes(group=Treatment)) +  # Adds an ellipse for visualization
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +  
  facet_wrap(~Microniche, ncol = 5) + # Facet into 5 columns
  theme_bw()


# Print the plot
PCoA_plot_all_microniche

ggsave("PCoA_plot_microniche.pdf", plot = PCoA_plot_all_microniche, width = 8, height = 3, dpi = 300)


#help from claude
# Remove samples where Biogeography is NA
btphylo_atsacrifice_pruned_transformed_clean <- subset_samples(
  btphylo_atsacrifice_pruned_transformed, 
  !is.na(Biogeography)
)

# Then recalculate distance and ordination on the clean object
distance_bray <- phyloseq::distance(btphylo_atsacrifice_pruned_transformed_clean, method = "bray")
ordinate_PCoA_bray <- ordinate(btphylo_atsacrifice_pruned_transformed_clean, method = "PCoA", distance = distance_bray)

# Plot using the clean object
base_plot <- plot_ordination(btphylo_atsacrifice_pruned_transformed_clean, 
                             ordinate_PCoA_bray, 
                             color = "Treatment", shape = "Sex")

# Build final plot
PCoA_plot_all_biogeography_claude <- base_plot +
  geom_point(data = subset(base_plot$data, Sex == "M"),
             aes(fill = Treatment, shape = Sex),
             color = "black", size = 1.5, stroke = 0.8) +
  stat_ellipse(aes(group = Treatment)) +
  scale_shape_manual(values = c("F" = 21, "M" = 24)) +
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  scale_fill_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  facet_wrap(~Biogeography, ncol = 5) +
  theme_bw()

PCoA_plot_all_biogeography_claude

# Save the PCoA plot as a PDF
ggsave("PCoA_plot_biogeography_PR.pdf", plot = PCoA_plot_all_biogeography_claude, width = 8, height = 4, dpi = 300)

# Build final plot for microniche

base_plot_2 <- plot_ordination(btphylo_atsacrifice_pruned_transformed, 
                             ordinate_PCoA_bray, 
                             color = "Treatment", shape = "Sex")

PCoA_plot_all_microniche_claude <- base_plot_2 +
  geom_point(data = subset(base_plot_2$data, Sex == "F"),
             aes(x = Axis.1, y = Axis.2, fill = Treatment, shape = Sex),
             color = NA, size = 1.5) +
  geom_point(data = subset(base_plot_2$data, Sex == "M"),
             aes(x = Axis.1, y = Axis.2, fill = Treatment, shape = Sex),
             color = "black", size = 1.5, stroke = 0.8) +
  stat_ellipse(aes(group = Treatment)) +
  scale_shape_manual(values = c("F" = 21, "M" = 24)) +
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange"), guide = "none") +
  scale_fill_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange"), name = "Treatment") +
  facet_wrap(~Microniche, ncol = 5) +
  theme_bw()

PCoA_plot_all_microniche_claude

ggsave("PCoA_plot_microniche_PR.pdf", plot = PCoA_plot_all_microniche_claude, width = 8, height = 4, dpi = 300)



```
#Plot alpha diversity across the gut and microniches
```{r}
#Plot  alpha diversity for each region. Here you should use unnormalized data.
alpha_diversity_shannon_biogeography <- plot_richness(btphylo_atsacrifice, x="Biogeography", color = "Treatment", measures=c("Shannon"))+
  geom_boxplot()+
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange"))+
  labs(y = "Alpha Diversity - Shannon", x = "Biogeography", color = "Treatment Group")+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=0.5))
  

alpha_diversity_shannon_biogeography

alpha_diversity_simpson_biogeography <- plot_richness(btphylo_atsacrifice, x="Biogeography", color = "Treatment", measures=c("Simpson"))+
  geom_boxplot()+
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange"))+
  labs(y = "Alpha Diversity - Simpson", x = "Biogeography", color = "Treatment Group")+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=0.5))
  

alpha_diversity_simpson_biogeography


alpha_diversity_Chao1_biogeography <- plot_richness(btphylo_atsacrifice, x="Biogeography", color = "Treatment", measures=c("Chao1"))+
  geom_boxplot()+
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange"))+
  labs(y = "Alpha Diversity - Chao1", x = "Biogeography", color = "Treatment Group")+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=0.5))
  

alpha_diversity_Chao1_biogeography

  
# Step 1: Estimate Fisher alpha diversity from the phyloseq object
fisher_diversity <- estimate_richness(btphylo_atsacrifice, measures = "Fisher")

# Step 2: Combine Fisher diversity with sample data
# Convert the Fisher diversity to a data frame
fisher_df <- as.data.frame(fisher_diversity)

# Ensure the row names (samples) are included in the data frame
fisher_df$Sample <- rownames(fisher_df)

# Merge with your metadata
combined_df <- fisher_df %>%
  left_join(metadata_all, by = "Sample")  # Adjust "SampleID" according to your metadata

# Set factor levels for Treatment to ensure "PBS" comes before "Bacteroides"
combined_df$Treatment <- factor(combined_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))

# Set factor levels for Biogeography in the desired order
combined_df$Biogeography <- factor(combined_df$Biogeography, 
                                            levels = c("SmallIntestine", "Cecum", "ProximalColon", "DistalColon"))



# Step 3: Create the alpha diversity plot for Fisher's measure across gut
alpha_fisher_across_gut <- ggplot(combined_df, 
                                   aes(x = Biogeography, y = Fisher, fill = Treatment)) +
  geom_boxplot(color = "black", alpha = 0.6) +  # Boxplot with borders and transparency
  geom_point(aes(color = Treatment, shape = Sex),  # Points mapped to Treatment and shape
             position = position_jitterdodge(dodge.width = 0.75),  # Jitter points for visibility
             alpha = 0.8, size = 2) +  # Points with defined aesthetics
   scale_fill_manual(values = c("Baseline" = "lightgreen", 
                                "PBS" = "lightblue", 
                                "Bacteroides" = "orange")) +  # Fill colors for treatments
  scale_color_manual(values = c("Baseline" = "lightgreen", 
                                 "PBS" = "lightblue", 
                                 "Bacteroides" = "orange")) +  # Color mapping for points
  theme_minimal() +theme_minimal() +
  labs(x = "Biogeography",
       y = expression("Alpha Diversity - Fisher"),  # Y-axis label with expression
       fill = "Treatment") +  # Updated fill label
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
        legend.position = "right")  # Position the legend

# Display the plot
print(alpha_fisher_across_gut)


ggsave("alpha_diversity_fisher_boxplot_acrossgut.pdf", plot = alpha_fisher_across_gut, device = "pdf", width = 5, height = 4)

#Run stats to see if there's a region that has more alpha diversity at baseline
# Subset the data for Baseline
acrossgut_baseline_fisher_df <- combined_df %>%
  filter(Treatment == "Baseline")

shapiro_acrossgut_baseline <- shapiro.test(acrossgut_baseline_fisher_df$Fisher)
shapiro_acrossgut_baseline

fisher_alpha_lm_acrossgut_baseline <- lm(Fisher ~ Biogeography+Sex+Biogeography*Sex+Microniche, data = acrossgut_baseline_fisher_df)
summary(fisher_alpha_lm_acrossgut_baseline)


#Run stats for alpha diversity along the gut. Run a lm for each region to see if there's a difference between PBS, 

# Subset the data for SmallIntestine
small_intestine_fisher_df <- combined_df %>%
  filter(Biogeography == "SmallIntestine")


small_intestine_pvb_fisher_df <- small_intestine_fisher_df %>% filter(Treatment %in% c("PBS", "Bacteroides"))


# Run the linear model on the subset
shapiro_acrossgut_smallintestine_fisher <- shapiro.test(small_intestine_fisher_df$Fisher)
shapiro_acrossgut_smallintestine_fisher

fisher_alpha_lm_small_intestine <- lm(Fisher ~ Treatment + Sex + Microniche+Treatment*Sex, data = small_intestine_fisher_df)
summary(fisher_alpha_lm_small_intestine)

fisher_alpha_lm_small_intestine_pvb <- lm(Fisher ~ Treatment + Sex + Microniche+Treatment*Sex, data = small_intestine_pvb_fisher_df)
summary(fisher_alpha_lm_small_intestine_pvb)

# Subset the data for Cecum
Cecum_fisher_df <- combined_df %>%
  filter(Biogeography == "Cecum")



cecum_pvb_fisher_df <- Cecum_fisher_df %>% filter(Treatment %in% c("PBS", "Bacteroides"))


Cecum_fisher_df$Treatment <- factor(Cecum_fisher_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))

# Run the linear model on the subset
shapiro_acrossgut_cecum_fisher <- shapiro.test(Cecum_fisher_df$Fisher)
shapiro_acrossgut_cecum_fisher

fisher_alpha_lm_cecum <- lm(log2(Fisher) ~ Treatment + Sex + Microniche + Treatment*Sex, data = Cecum_fisher_df)
summary(fisher_alpha_lm_cecum)

shapiro_acrossgut_cecum_fisher_pvb <- shapiro.test(cecum_pvb_fisher_df$Fisher)
shapiro_acrossgut_cecum_fisher_pvb

fisher_alpha_lm_cecum_pvb <- lm(Fisher ~ Treatment + Sex + Microniche + Treatment*Sex, data = cecum_pvb_fisher_df)
summary(fisher_alpha_lm_cecum_pvb)

# Subset the data for ProximalColon

ProximalColon_fisher_df <- combined_df %>%
  filter(Biogeography == "ProximalColon")

proximalcolon_pvb_fisher_df <- ProximalColon_fisher_df %>% filter(Treatment %in% c("PBS", "Bacteroides"))


ProximalColon_fisher_df$Treatment <- factor(ProximalColon_fisher_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))

ProximalColon_fisher_df$Treatment <- factor(ProximalColon_fisher_df$Treatment, 
                                         levels = c("PBS", "Bacteroides", "Baseline"))


# Run the linear model on the subset
shapiro_acrossgut_proximalcolon_fisher <- shapiro.test(ProximalColon_fisher_df$Fisher)
shapiro_acrossgut_proximalcolon_fisher

fisher_alpha_lm_ProximalColon <- lm(log2(Fisher) ~ Treatment + Sex + Microniche+Treatment*Sex, data = ProximalColon_fisher_df)
summary(fisher_alpha_lm_ProximalColon)

shapiro_acrossgut_proximalcolon_fisher_pvb<- shapiro.test(proximalcolon_pvb_fisher_df$Fisher)
shapiro_acrossgut_proximalcolon_fisher_pvb

fisher_alpha_lm_ProximalColon_pvb <- lm(Fisher ~ Treatment + Sex + Microniche+Treatment*Sex, data = proximalcolon_pvb_fisher_df)
summary(fisher_alpha_lm_ProximalColon_pvb)


# Subset the data for DistalColon

DistalColon_fisher_df <- combined_df %>%
  filter(Biogeography == "DistalColon")

distalcolon_pvb_fisher_df <- DistalColon_fisher_df %>% filter(Treatment %in% c("PBS", "Bacteroides"))


DistalColon_fisher_df$Treatment <- factor(DistalColon_fisher_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))

DistalColon_fisher_df$Treatment <- factor(DistalColon_fisher_df$Treatment, 
                                         levels = c("PBS", "Bacteroides", "Baseline"))


# Run the linear model on the subset
shapiro_acrossgut_distalcolon_fisher <- shapiro.test(DistalColon_fisher_df$Fisher)
shapiro_acrossgut_distalcolon_fisher

fisher_alpha_lm_DistalColon <- lm(log2(Fisher) ~ Treatment + Sex + Microniche+Treatment*Sex, data = DistalColon_fisher_df)
summary(fisher_alpha_lm_DistalColon)

shapiro_acrossgut_distalcolon_fisher_pvb <- shapiro.test(distalcolon_pvb_fisher_df$Fisher)
shapiro_acrossgut_distalcolon_fisher_pvb

fisher_alpha_lm_DistalColon_pvb <- lm(Fisher ~ Treatment + Sex + Microniche+Treatment*Sex, data = distalcolon_pvb_fisher_df)
summary(fisher_alpha_lm_DistalColon_pvb)


#Plot alpha diverisity for microniche
btphylo_atsacrifice_microniche <- subset_samples(btphylo_atsacrifice, !is.na(Microniche) & Microniche != "")

alpha_diversity_shannon_Microniche <- plot_richness(btphylo_atsacrifice_microniche, x="Microniche", color = "Treatment", measures=c("Shannon"))+
  geom_boxplot()+
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange"))+
  labs(y = "Alpha Diversity - Shannon", x = "Microniche", color = "Treatment Group")+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=0.5))
  

alpha_diversity_shannon_Microniche

alpha_diversity_Chao1_Microniche <- plot_richness(btphylo_atsacrifice_microniche, x="Microniche", color = "Treatment", measures=c("Chao1"))+
  geom_boxplot()+
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange"))+
  labs(y = "Alpha Diversity - Chao-1", x = "Microniche", color = "Treatment Group")+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=0.5))
  

alpha_diversity_Chao1_Microniche

#Plot alpha diversity between microniches
combined_df_filtered <- combined_df %>%
  filter(!is.na(Microniche))  # Keep only rows where Microniche is not NA


alpha_fisher_across_microniche <- ggplot(combined_df_filtered, 
                                   aes(x = Microniche, y = Fisher, fill = Treatment)) +
  geom_boxplot(color = "black", alpha = 0.6) +  # Boxplot with borders and transparency
  geom_point(aes(color = Treatment, shape = Sex),  # Points mapped to Treatment and shape
             position = position_jitterdodge(dodge.width = 0.75),  # Jitter points for visibility
             alpha = 0.8, size = 2) +  # Points with defined aesthetics
   scale_fill_manual(values = c("Baseline" = "lightgreen", 
                                "PBS" = "lightblue", 
                                "Bacteroides" = "orange")) +  # Fill colors for treatments
  scale_color_manual(values = c("Baseline" = "lightgreen", 
                                 "PBS" = "lightblue", 
                                 "Bacteroides" = "orange")) +  # Color mapping for points
  theme_minimal() +theme_minimal() +
  labs(x = "Microniche",
       y = expression("Alpha Diversity - Fisher"),  # Y-axis label with expression
       fill = "Treatment") +  # Updated fill label
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
        legend.position = "right")  # Position the legend

# Display the plot
print(alpha_fisher_across_microniche)

ggsave("alpha_diversity_fisher_boxplot_microniche.pdf", plot = alpha_fisher_across_microniche, device = "pdf", width = 5, height = 4)

#Run stats to see if there's a microniche that has more alpha diversity at baseline
# Subset the data for Baseline
microniche_baseline_fisher_df <- combined_df_filtered %>%
  filter(Treatment == "Baseline")

shapiro_microniche_baseline_fisher <- shapiro.test(microniche_baseline_fisher_df$Fisher)
shapiro_microniche_baseline_fisher

fisher_alpha_lm_microniche_baseline <- lm(Fisher ~ Microniche+Microniche*Sex+Biogeography+Microniche*Biogeography, data = microniche_baseline_fisher_df)
summary(fisher_alpha_lm_microniche_baseline)




# Subset the data for LuminalContents

luminalcontents_fisher_df <- combined_df_filtered %>%
  filter(Microniche == "LuminalContents")

luminalcontents_pvb_fisher_df <- luminalcontents_fisher_df %>% filter(Treatment %in% c("PBS", "Bacteroides"))


luminalcontents_fisher_df$Treatment <- factor(luminalcontents_fisher_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))

luminalcontents_fisher_df$Treatment <- factor(luminalcontents_fisher_df$Treatment, 
                                         levels = c("PBS", "Bacteroides", "Baseline"))

# Run the linear model on the subset
shapiro_microniche_luminalcontents_fisher <- shapiro.test(luminalcontents_fisher_df$Fisher)
shapiro_microniche_luminalcontents_fisher

fisher_alpha_lm_luminalcontents <- lm(log2(Fisher) ~ Treatment + Sex +Treatment*Sex+Biogeography, data = luminalcontents_fisher_df)
summary(fisher_alpha_lm_luminalcontents)

shapiro_microniche_luminalcontents_fisher_pvb <- shapiro.test(luminalcontents_pvb_fisher_df$Fisher)
shapiro_microniche_luminalcontents_fisher_pvb

fisher_alpha_pvb_lm_luminalcontents <- lm(Fisher ~ Treatment + Sex +Treatment*Sex+Biogeography, data = luminalcontents_pvb_fisher_df)
summary(fisher_alpha_pvb_lm_luminalcontents)

# Subset the data for MucosalSurface

MucosalSurface_fisher_df <- combined_df_filtered %>%
  filter(Microniche == "MucosalSurface")

MucosalSurface_pvb_fisher_df <- MucosalSurface_fisher_df %>% filter(Treatment %in% c("PBS", "Bacteroides"))


MucosalSurface_fisher_df$Treatment <- factor(MucosalSurface_fisher_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))

MucosalSurface_fisher_df$Treatment <- factor(MucosalSurface_fisher_df$Treatment, 
                                         levels = c("PBS", "Bacteroides", "Baseline"))

# Run the linear model on the subset
shapiro_microniche_mucosalsurface_fisher <- shapiro.test(MucosalSurface_fisher_df$Fisher)
shapiro_microniche_mucosalsurface_fisher

fisher_alpha_lm_MucosalSurface <- lm(log2(Fisher) ~ Treatment + Sex +Treatment*Sex + Biogeography, data = MucosalSurface_fisher_df)
summary(fisher_alpha_lm_MucosalSurface)

shapiro_microniche_mucosalsurface_fisher_pvb <- shapiro.test(MucosalSurface_pvb_fisher_df$Fisher)
shapiro_microniche_mucosalsurface_fisher_pvb

fisher_alpha_lm_MucosalSurface_pvb <- lm(Fisher ~ Treatment + Sex +Treatment*Sex+Biogeography, data = MucosalSurface_pvb_fisher_df)
summary(fisher_alpha_lm_MucosalSurface_pvb)

```


Plot the Pielou's Evenness and Observed Richness across gut
```{r}
# Step 1: Estimate Shannon alpha diversity from the phyloseq object
Shannon_diversity_atsacrifice <- estimate_richness(btphylo_atsacrifice, measures = "Shannon")
observed_diversity_atsacrifice <- estimate_richness(btphylo_atsacrifice, measures = "Observed")


# Step 2: Combine Shannon diversity with sample data
# Convert the Shannon diversity to a data frame
Shannon_atsacrifice_df <- as.data.frame(Shannon_diversity_atsacrifice)
observed_atsacrifice_df <- as.data.frame(observed_diversity_atsacrifice)


# Ensure the row names (samples) are included in the data frame
Shannon_atsacrifice_df$Sample <- rownames(Shannon_atsacrifice_df)

observed_atsacrifice_df$Sample <- rownames(observed_atsacrifice_df)
observed_atsacrifice_df$Sample <- gsub("X", "", observed_atsacrifice_df$Sample) #remove the x that is coming up for some reason


metadata_all$Sample <- rownames(metadata_all)

# Merge with your metadata
combined_atsacrifice_df <- Shannon_atsacrifice_df %>%
  left_join(metadata_all, by ="Sample")  # Adjust "SampleID" according to your metadata

# Merge with your metadata
combined_atsacrifice_df <- combined_atsacrifice_df %>%
  left_join(observed_atsacrifice_df, by ="Sample")  # Adjust "SampleID" according to your metadata

combined_atsacrifice_df$Pielou <- combined_atsacrifice_df$Shannon / log(combined_atsacrifice_df$Observed)

# Set factor levels for Treatment to ensure "PBS" comes before "Bacteroides"
combined_atsacrifice_df$Treatment <- factor(combined_atsacrifice_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))

# Set factor levels for Treatment to ensure "PBS" comes before "Bacteroides"
combined_atsacrifice_df$Biogeography <- factor(combined_atsacrifice_df$Biogeography, 
                                         levels = c("SmallIntestine", "Cecum", "ProximalColon", "DistalColon"))


combined_atsacrifice_filtered <- combined_atsacrifice_df %>%
  filter(Compiled_Timepoint != "Baseline")  # Keep only rows where Compiled_Timepoint is not "Baseline"



alpha_Observed_across_gut <- ggplot(combined_atsacrifice_filtered, 
                                   aes(x = Biogeography, y = Observed, fill = Treatment)) +
  geom_boxplot(color = "black", alpha = 0.6) +  # Boxplot with borders and transparency
  geom_point(aes(color = Treatment, shape = Sex),  # Points mapped to Treatment and shape
             position = position_jitterdodge(dodge.width = 0.75),  # Jitter points for visibility
             alpha = 0.8, size = 2) +  # Points with defined aesthetics
   scale_fill_manual(values = c("Baseline" = "lightgreen", 
                                "PBS" = "lightblue", 
                                "Bacteroides" = "orange")) +  # Fill colors for treatments
  scale_color_manual(values = c("Baseline" = "lightgreen", 
                                 "PBS" = "lightblue", 
                                 "Bacteroides" = "orange")) +  # Color mapping for points
  theme_minimal()+
  labs(x = "Experiment Day",
       y = expression("Alpha Diversity - Observed"),  # Y-axis label with expression
       fill = "Treatment") +  # Updated fill label
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
        legend.position = "right")  # Position the legend

# Display the plot
print(alpha_Observed_across_gut)


# Step 3: Create the alpha diversity plot for Shannon's measure across gut
alpha_Pielou_across_gut <- ggplot(combined_atsacrifice_filtered, 
                                   aes(x = Biogeography, y = Pielou, fill = Treatment)) +
  geom_boxplot(color = "black", alpha = 0.6) +  # Boxplot with borders and transparency
  geom_point(aes(color = Treatment, shape = Sex),  # Points mapped to Treatment and shape
             position = position_jitterdodge(dodge.width = 0.75),  # Jitter points for visibility
             alpha = 0.8, size = 2) +  # Points with defined aesthetics
   scale_fill_manual(values = c("Baseline" = "lightgreen", 
                                "PBS" = "lightblue", 
                                "Bacteroides" = "orange")) +  # Fill colors for treatments
  scale_color_manual(values = c("Baseline" = "lightgreen", 
                                 "PBS" = "lightblue", 
                                 "Bacteroides" = "orange")) +  # Color mapping for points
  theme_minimal()+
  labs(x = "Biogeography",
       y = expression("Alpha Diversity - Pielou"),  # Y-axis label with expression
       fill = "Treatment") +  # Updated fill label
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
        legend.position = "right")  # Position the legend

# Display the plot
print(alpha_Pielou_across_gut)

ggsave("alpha_Pielou_across_gut_boxplot.pdf", plot = alpha_Pielou_across_gut, device = "pdf", width = 6, height = 4)

ggsave("alpha_Observed_across_gut_boxplot.pdf", plot = alpha_Observed_across_gut, device = "pdf", width = 6, height = 4)

```
#Run stats on observed alpha diversity across biogeography
```{r}
#Run stats to see if there's a region that has more alpha diversity at baseline
# Subset the data for Baseline
acrossgut_baseline_Observed_df <- combined_atsacrifice_filtered %>%
  filter(Treatment == "Baseline")

shapiro_acrossgut_baseline_observed <- shapiro.test(acrossgut_baseline_Observed_df$Observed)
shapiro_acrossgut_baseline_observed

Observed_alpha_lm_acrossgut_baseline <- lm(Observed ~ Biogeography+Sex+Biogeography*Sex+Microniche, data = acrossgut_baseline_Observed_df)
summary(Observed_alpha_lm_acrossgut_baseline)


#Run stats for alpha diversity along the gut. Run a lm for each region to see if there's a difference between PBS, 

# Subset the data for SmallIntestine
small_intestine_Observed_df <- combined_atsacrifice_filtered %>%
  filter(Biogeography == "SmallIntestine")


small_intestine_pvb_Observed_df <- small_intestine_Observed_df %>% filter(Treatment %in% c("PBS", "Bacteroides"))


# Run the linear model on the subset
shapiro_acrossgut_smallintestine_Observed <- shapiro.test(small_intestine_Observed_df$Observed)
shapiro_acrossgut_smallintestine_Observed

Observed_alpha_lm_small_intestine <- lm(Observed ~ Treatment + Sex + Microniche+Treatment*Sex, data = small_intestine_Observed_df)
summary(Observed_alpha_lm_small_intestine)

shapiro_acrossgut_smallintestine_Observed_pvb <- shapiro.test(small_intestine_pvb_Observed_df$Observed)
shapiro_acrossgut_smallintestine_Observed_pvb

Observed_alpha_lm_small_intestine_pvb <- lm(Observed ~ Treatment + Sex + Microniche+Treatment*Sex, data = small_intestine_pvb_Observed_df)
summary(Observed_alpha_lm_small_intestine_pvb)

# Subset the data for Cecum
Cecum_Observed_df <- combined_atsacrifice_filtered %>%
  filter(Biogeography == "Cecum")



cecum_pvb_Observed_df <- Cecum_Observed_df %>% filter(Treatment %in% c("PBS", "Bacteroides"))


Cecum_Observed_df$Treatment <- factor(Cecum_Observed_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))

# Run the linear model on the subset
shapiro_acrossgut_cecum_Observed <- shapiro.test(Cecum_Observed_df$Observed)
shapiro_acrossgut_cecum_Observed

Observed_alpha_lm_cecum <- lm(log2(Observed) ~ Treatment + Sex + Microniche + Treatment*Sex, data = Cecum_Observed_df)
summary(Observed_alpha_lm_cecum)

shapiro_acrossgut_cecum_Observed_pvb <- shapiro.test(cecum_pvb_Observed_df$Observed)
shapiro_acrossgut_cecum_Observed_pvb

Observed_alpha_lm_cecum_pvb <- lm(Observed ~ Treatment + Sex + Microniche + Treatment*Sex, data = cecum_pvb_Observed_df)
summary(Observed_alpha_lm_cecum_pvb)

# Subset the data for ProximalColon

ProximalColon_Observed_df <- combined_atsacrifice_filtered %>%
  filter(Biogeography == "ProximalColon")

proximalcolon_pvb_Observed_df <- ProximalColon_Observed_df %>% filter(Treatment %in% c("PBS", "Bacteroides"))


ProximalColon_Observed_df$Treatment <- factor(ProximalColon_Observed_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))

ProximalColon_Observed_df$Treatment <- factor(ProximalColon_Observed_df$Treatment, 
                                         levels = c("PBS", "Bacteroides", "Baseline"))


# Run the linear model on the subset
shapiro_acrossgut_proximalcolon_Observed <- shapiro.test(ProximalColon_Observed_df$Observed)
shapiro_acrossgut_proximalcolon_Observed

Observed_alpha_lm_ProximalColon <- lm(log2(Observed) ~ Treatment + Sex + Microniche+Treatment*Sex, data = ProximalColon_Observed_df)
summary(Observed_alpha_lm_ProximalColon)

shapiro_acrossgut_proximalcolon_Observed_pvb<- shapiro.test(proximalcolon_pvb_Observed_df$Observed)
shapiro_acrossgut_proximalcolon_Observed_pvb

Observed_alpha_lm_ProximalColon_pvb <- lm(log2(Observed) ~ Treatment + Sex + Microniche+Treatment*Sex, data = proximalcolon_pvb_Observed_df)
summary(Observed_alpha_lm_ProximalColon_pvb)


# Subset the data for DistalColon

DistalColon_Observed_df <- combined_atsacrifice_filtered %>%
  filter(Biogeography == "DistalColon")

distalcolon_pvb_Observed_df <- DistalColon_Observed_df %>% filter(Treatment %in% c("PBS", "Bacteroides"))


DistalColon_Observed_df$Treatment <- factor(DistalColon_Observed_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))

DistalColon_Observed_df$Treatment <- factor(DistalColon_Observed_df$Treatment, 
                                         levels = c("PBS", "Bacteroides", "Baseline"))


# Run the linear model on the subset
shapiro_acrossgut_distalcolon_Observed <- shapiro.test(DistalColon_Observed_df$Observed)
shapiro_acrossgut_distalcolon_Observed

Observed_alpha_lm_DistalColon <- lm(log2(Observed) ~ Treatment + Sex + Microniche+Treatment*Sex, data = DistalColon_Observed_df)
summary(Observed_alpha_lm_DistalColon)

shapiro_acrossgut_distalcolon_Observed_pvb <- shapiro.test(distalcolon_pvb_Observed_df$Observed)
shapiro_acrossgut_distalcolon_Observed_pvb

Observed_alpha_lm_DistalColon_pvb <- lm(log2(Observed) ~ Treatment + Sex + Microniche+Treatment*Sex, data = distalcolon_pvb_Observed_df)
summary(Observed_alpha_lm_DistalColon_pvb)

```

#Run stats on Pielou alpha diversity across biogeography
```{r}
#Run stats to see if there's a region that has more alpha diversity at baseline
# Subset the data for Baseline
acrossgut_baseline_Pielou_df <- combined_atsacrifice_filtered %>%
  filter(Treatment == "Baseline")

shapiro_acrossgut_baseline_Pielou <- shapiro.test(acrossgut_baseline_Pielou_df$Pielou)
shapiro_acrossgut_baseline_Pielou

Pielou_alpha_lm_acrossgut_baseline <- lm(log2(Pielou) ~ Biogeography+Sex+Biogeography*Sex+Microniche, data = acrossgut_baseline_Pielou_df)
summary(Pielou_alpha_lm_acrossgut_baseline)


#Run stats for alpha diversity along the gut. Run a lm for each region to see if there's a difference between PBS, 

# Subset the data for SmallIntestine
small_intestine_Pielou_df <- combined_atsacrifice_filtered %>%
  filter(Biogeography == "SmallIntestine")


small_intestine_pvb_Pielou_df <- small_intestine_Pielou_df %>% filter(Treatment %in% c("PBS", "Bacteroides"))


# Run the linear model on the subset
shapiro_acrossgut_smallintestine_Pielou <- shapiro.test(small_intestine_Pielou_df$Pielou)
shapiro_acrossgut_smallintestine_Pielou

Pielou_alpha_lm_small_intestine <- lm(log2(Pielou) ~ Treatment + Sex + Microniche+Treatment*Sex, data = small_intestine_Pielou_df)
summary(Pielou_alpha_lm_small_intestine)

shapiro_acrossgut_smallintestine_Pielou_pvb <- shapiro.test(small_intestine_pvb_Pielou_df$Pielou)
shapiro_acrossgut_smallintestine_Pielou_pvb

Pielou_alpha_lm_small_intestine_pvb <- lm(Pielou ~ Treatment + Sex + Microniche+Treatment*Sex, data = small_intestine_pvb_Pielou_df)
summary(Pielou_alpha_lm_small_intestine_pvb)

# Subset the data for Cecum
Cecum_Pielou_df <- combined_atsacrifice_filtered %>%
  filter(Biogeography == "Cecum")



cecum_pvb_Pielou_df <- Cecum_Pielou_df %>% filter(Treatment %in% c("PBS", "Bacteroides"))


Cecum_Pielou_df$Treatment <- factor(Cecum_Pielou_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))

# Run the linear model on the subset
shapiro_acrossgut_cecum_Pielou <- shapiro.test(Cecum_Pielou_df$Pielou)
shapiro_acrossgut_cecum_Pielou

Pielou_alpha_lm_cecum <- lm(Pielou ~ Treatment + Sex + Microniche + Treatment*Sex, data = Cecum_Pielou_df)
summary(Pielou_alpha_lm_cecum)

shapiro_acrossgut_cecum_Pielou_pvb <- shapiro.test(cecum_pvb_Pielou_df$Pielou)
shapiro_acrossgut_cecum_Pielou_pvb

Pielou_alpha_lm_cecum_pvb <- lm(Pielou ~ Treatment + Sex + Microniche + Treatment*Sex, data = cecum_pvb_Pielou_df)
summary(Pielou_alpha_lm_cecum_pvb)

# Subset the data for ProximalColon

ProximalColon_Pielou_df <- combined_atsacrifice_filtered %>%
  filter(Biogeography == "ProximalColon")

proximalcolon_pvb_Pielou_df <- ProximalColon_Pielou_df %>% filter(Treatment %in% c("PBS", "Bacteroides"))


ProximalColon_Pielou_df$Treatment <- factor(ProximalColon_Pielou_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))

ProximalColon_Pielou_df$Treatment <- factor(ProximalColon_Pielou_df$Treatment, 
                                         levels = c("PBS", "Bacteroides", "Baseline"))


# Run the linear model on the subset
shapiro_acrossgut_proximalcolon_Pielou <- shapiro.test(ProximalColon_Pielou_df$Pielou)
shapiro_acrossgut_proximalcolon_Pielou

Pielou_alpha_lm_ProximalColon <- lm(log2(Pielou) ~ Treatment + Sex + Microniche+Treatment*Sex, data = ProximalColon_Pielou_df)
summary(Pielou_alpha_lm_ProximalColon)

shapiro_acrossgut_proximalcolon_Pielou_pvb<- shapiro.test(proximalcolon_pvb_Pielou_df$Pielou)
shapiro_acrossgut_proximalcolon_Pielou_pvb

Pielou_alpha_lm_ProximalColon_pvb <- lm(log2(Pielou) ~ Treatment + Sex + Microniche+Treatment*Sex, data = proximalcolon_pvb_Pielou_df)
summary(Pielou_alpha_lm_ProximalColon_pvb)


# Subset the data for DistalColon

DistalColon_Pielou_df <- combined_atsacrifice_filtered %>%
  filter(Biogeography == "DistalColon")

distalcolon_pvb_Pielou_df <- DistalColon_Pielou_df %>% filter(Treatment %in% c("PBS", "Bacteroides"))


DistalColon_Pielou_df$Treatment <- factor(DistalColon_Pielou_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))

DistalColon_Pielou_df$Treatment <- factor(DistalColon_Pielou_df$Treatment, 
                                         levels = c("PBS", "Bacteroides", "Baseline"))


# Run the linear model on the subset
shapiro_acrossgut_distalcolon_Pielou <- shapiro.test(DistalColon_Pielou_df$Pielou)
shapiro_acrossgut_distalcolon_Pielou

Pielou_alpha_lm_DistalColon <- lm(Pielou ~ Treatment + Sex + Microniche+Treatment*Sex, data = DistalColon_Pielou_df)
summary(Pielou_alpha_lm_DistalColon)

shapiro_acrossgut_distalcolon_Pielou_pvb <- shapiro.test(distalcolon_pvb_Pielou_df$Pielou)
shapiro_acrossgut_distalcolon_Pielou_pvb

Pielou_alpha_lm_DistalColon_pvb <- lm(Pielou ~ Treatment + Sex + Microniche+Treatment*Sex, data = distalcolon_pvb_Pielou_df)
summary(Pielou_alpha_lm_DistalColon_pvb)

```

Plot the Pielou's Evenness and Observed Richness across microniche

```{r}

combined_atsacrifice_filtered_microniche <- combined_atsacrifice_filtered %>%
  filter(!is.na(Microniche))  # Keep only rows where Microniche is not NA


alpha_Observed_across_microniche <- ggplot(combined_atsacrifice_filtered_microniche, 
                                   aes(x = Microniche, y = Observed, fill = Treatment)) +
  geom_boxplot(color = "black", alpha = 0.6) +  # Boxplot with borders and transparency
  geom_point(aes(color = Treatment, shape = Sex),  # Points mapped to Treatment and shape
             position = position_jitterdodge(dodge.width = 0.75),  # Jitter points for visibility
             alpha = 0.8, size = 2) +  # Points with defined aesthetics
   scale_fill_manual(values = c("Baseline" = "lightgreen", 
                                "PBS" = "lightblue", 
                                "Bacteroides" = "orange")) +  # Fill colors for treatments
  scale_color_manual(values = c("Baseline" = "lightgreen", 
                                 "PBS" = "lightblue", 
                                 "Bacteroides" = "orange")) +  # Color mapping for points
  theme_minimal()+
  labs(x = "Microniche",
       y = expression("Alpha Diversity - Observed"),  # Y-axis label with expression
       fill = "Treatment") +  # Updated fill label
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
        legend.position = "right")  # Position the legend

alpha_Observed_across_microniche


alpha_Pielou_across_microniche <- ggplot(combined_atsacrifice_filtered_microniche, 
                                   aes(x = Microniche, y = Pielou, fill = Treatment)) +
  geom_boxplot(color = "black", alpha = 0.6) +  # Boxplot with borders and transparency
  geom_point(aes(color = Treatment, shape = Sex),  # Points mapped to Treatment and shape
             position = position_jitterdodge(dodge.width = 0.75),  # Jitter points for visibility
             alpha = 0.8, size = 2) +  # Points with defined aesthetics
   scale_fill_manual(values = c("Baseline" = "lightgreen", 
                                "PBS" = "lightblue", 
                                "Bacteroides" = "orange")) +  # Fill colors for treatments
  scale_color_manual(values = c("Baseline" = "lightgreen", 
                                 "PBS" = "lightblue", 
                                 "Bacteroides" = "orange")) +  # Color mapping for points
  theme_minimal()+
  labs(x = "Microniche",
       y = expression("Alpha Diversity - Pielou"),  # Y-axis label with expression
       fill = "Treatment") +  # Updated fill label
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
        legend.position = "right")  # Position the legend

alpha_Pielou_across_microniche

ggsave("alpha_Pielou_across__microniche_boxplot.pdf", plot = alpha_Pielou_across_microniche, device = "pdf", width = 6, height = 4)

ggsave("alpha_Observed_across_microniche_boxplot.pdf", plot = alpha_Observed_across_microniche, device = "pdf", width = 6, height = 4)

```

#Run stats on Observed alpha diversity across microniches
```{r}
#Run stats to see if there's a microniche that has more alpha diversity at baseline
# Subset the data for Baseline
microniche_baseline_Observed_df <- combined_atsacrifice_filtered_microniche %>%
  filter(Treatment == "Baseline")

shapiro_microniche_baseline_Observed <- shapiro.test(microniche_baseline_Observed_df$Observed)
shapiro_microniche_baseline_Observed

Observed_alpha_lm_microniche_baseline <- lm(log2(Observed) ~ Microniche+Microniche*Sex+Biogeography+Microniche*Biogeography, data = microniche_baseline_Observed_df)
summary(Observed_alpha_lm_microniche_baseline)




# Subset the data for LuminalContents

luminalcontents_Observed_df <- combined_atsacrifice_filtered_microniche %>%
  filter(Microniche == "LuminalContents")

luminalcontents_pvb_Observed_df <- luminalcontents_Observed_df %>% filter(Treatment %in% c("PBS", "Bacteroides"))


luminalcontents_Observed_df$Treatment <- factor(luminalcontents_Observed_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))

luminalcontents_Observed_df$Treatment <- factor(luminalcontents_Observed_df$Treatment, 
                                         levels = c("PBS", "Bacteroides", "Baseline"))

# Run the linear model on the subset
shapiro_microniche_luminalcontents_Observed <- shapiro.test(luminalcontents_Observed_df$Observed)
shapiro_microniche_luminalcontents_Observed

Observed_alpha_lm_luminalcontents <- lm(log2(Observed) ~ Treatment + Sex +Treatment*Sex+Biogeography, data = luminalcontents_Observed_df)
summary(Observed_alpha_lm_luminalcontents)

shapiro_microniche_luminalcontents_Observed_pvb <- shapiro.test(luminalcontents_pvb_Observed_df$Observed)
shapiro_microniche_luminalcontents_Observed_pvb

Observed_alpha_pvb_lm_luminalcontents <- lm(Observed ~ Treatment + Sex +Treatment*Sex+Biogeography, data = luminalcontents_pvb_Observed_df)
summary(Observed_alpha_pvb_lm_luminalcontents)

# Subset the data for MucosalSurface

MucosalSurface_Observed_df <- combined_atsacrifice_filtered_microniche %>%
  filter(Microniche == "MucosalSurface")

MucosalSurface_pvb_Observed_df <- MucosalSurface_Observed_df %>% filter(Treatment %in% c("PBS", "Bacteroides"))


MucosalSurface_Observed_df$Treatment <- factor(MucosalSurface_Observed_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))

MucosalSurface_Observed_df$Treatment <- factor(MucosalSurface_Observed_df$Treatment, 
                                         levels = c("PBS", "Bacteroides", "Baseline"))

# Run the linear model on the subset
shapiro_microniche_mucosalsurface_Observed <- shapiro.test(MucosalSurface_Observed_df$Observed)
shapiro_microniche_mucosalsurface_Observed

Observed_alpha_lm_MucosalSurface <- lm(log2(Observed) ~ Treatment + Sex +Treatment*Sex + Biogeography, data = MucosalSurface_Observed_df)
summary(Observed_alpha_lm_MucosalSurface)

shapiro_microniche_mucosalsurface_Observed_pvb <- shapiro.test(MucosalSurface_pvb_Observed_df$Observed)
shapiro_microniche_mucosalsurface_Observed_pvb

Observed_alpha_lm_MucosalSurface_pvb <- lm(Observed ~ Treatment + Sex +Treatment*Sex+Biogeography, data = MucosalSurface_pvb_Observed_df)
summary(Observed_alpha_lm_MucosalSurface_pvb)

```

#Run stats on Pielou alpha diversity across microniches
```{r}
#Run stats to see if there's a microniche that has more alpha diversity at baseline
# Subset the data for Baseline
microniche_baseline_Pielou_df <- combined_atsacrifice_filtered_microniche %>%
  filter(Treatment == "Baseline")

shapiro_microniche_baseline_Pielou <- shapiro.test(microniche_baseline_Pielou_df$Pielou)
shapiro_microniche_baseline_Pielou

Pielou_alpha_lm_microniche_baseline <- lm(log2(Pielou) ~ Microniche+Microniche*Sex+Biogeography+Microniche*Biogeography, data = microniche_baseline_Pielou_df)
summary(Pielou_alpha_lm_microniche_baseline)




# Subset the data for LuminalContents

luminalcontents_Pielou_df <- combined_atsacrifice_filtered_microniche %>%
  filter(Microniche == "LuminalContents")

luminalcontents_pvb_Pielou_df <- luminalcontents_Pielou_df %>% filter(Treatment %in% c("PBS", "Bacteroides"))


luminalcontents_Pielou_df$Treatment <- factor(luminalcontents_Pielou_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))

luminalcontents_Pielou_df$Treatment <- factor(luminalcontents_Pielou_df$Treatment, 
                                         levels = c("PBS", "Bacteroides", "Baseline"))

# Run the linear model on the subset
shapiro_microniche_luminalcontents_Pielou <- shapiro.test(luminalcontents_Pielou_df$Pielou)
shapiro_microniche_luminalcontents_Pielou

Pielou_alpha_lm_luminalcontents <- lm(log2(Pielou) ~ Treatment + Sex +Treatment*Sex+Biogeography, data = luminalcontents_Pielou_df)
summary(Pielou_alpha_lm_luminalcontents)

shapiro_microniche_luminalcontents_Pielou_pvb <- shapiro.test(luminalcontents_pvb_Pielou_df$Pielou)
shapiro_microniche_luminalcontents_Pielou_pvb

Pielou_alpha_pvb_lm_luminalcontents <- lm(Pielou ~ Treatment + Sex +Treatment*Sex+Biogeography, data = luminalcontents_pvb_Pielou_df)
summary(Pielou_alpha_pvb_lm_luminalcontents)

# Subset the data for MucosalSurface

MucosalSurface_Pielou_df <- combined_atsacrifice_filtered_microniche %>%
  filter(Microniche == "MucosalSurface")

MucosalSurface_pvb_Pielou_df <- MucosalSurface_Pielou_df %>% filter(Treatment %in% c("PBS", "Bacteroides"))


MucosalSurface_Pielou_df$Treatment <- factor(MucosalSurface_Pielou_df$Treatment, 
                                         levels = c("Baseline", "PBS", "Bacteroides"))

MucosalSurface_Pielou_df$Treatment <- factor(MucosalSurface_Pielou_df$Treatment, 
                                         levels = c("PBS", "Bacteroides", "Baseline"))

# Run the linear model on the subset
shapiro_microniche_mucosalsurface_Pielou <- shapiro.test(MucosalSurface_Pielou_df$Pielou)
shapiro_microniche_mucosalsurface_Pielou

Pielou_alpha_lm_MucosalSurface <- lm(Pielou ~ Treatment + Sex +Treatment*Sex + Biogeography, data = MucosalSurface_Pielou_df)
summary(Pielou_alpha_lm_MucosalSurface)

shapiro_microniche_mucosalsurface_Pielou_pvb <- shapiro.test(MucosalSurface_pvb_Pielou_df$Pielou)
shapiro_microniche_mucosalsurface_Pielou_pvb

Pielou_alpha_lm_MucosalSurface_pvb <- lm(Pielou ~ Treatment + Sex +Treatment*Sex+Biogeography, data = MucosalSurface_pvb_Pielou_df)
summary(Pielou_alpha_lm_MucosalSurface_pvb)

```



```

```{r}
#Plot the results with points colored by the Treatment group.
PCoA_plot_all <- plot_ordination(btphylo_atsacrifice_pruned_transformed, ordinate_PCoA_bray, color = "Treatment",shape="Sex")+
  stat_ellipse(aes(group=Treatment)) + #adds an ellipse to a scatter plot to help visualize the distribution of points in the plot. The ellipse is drawn using a confidence interval, which can be specified using the level argument. By default, stat_ellipse() uses a 95% confidence interval.
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +  # Corrected scale_color_manual
   facet_wrap(~Microniche, ncol=3)

PCoA_plot_all

# Save the PCoA plot as a PDF
ggsave("PCoA_plot_Microniches.pdf", plot = PCoA_plot_all, width = 8, height = 5, dpi = 300)



```


#Run stats to see if Treatment has a significant effect on beta diversity in each biogeography, you can subset the data to only include samples from the small intestine and then run the PERMANOVA analysis. Here is the code that will do this:
```{r}


#Subset phyloseq object to only have the data from Small Intestine
btphylo_atsacrifice_pruned_transformed_smallintestine <- subset_samples(btphylo_atsacrifice_pruned_transformed_clean, 
                                                                       Biogeography %in% "SmallIntestine")

#Calculate Bray-Curtis dissimilarity distance matrix
distance_bray_smallintestine = phyloseq::distance(btphylo_atsacrifice_pruned_transformed_smallintestine, method = "bray")

#Perform a principal coordinate analysis (PCoA) on the distance matrix
smallintestine_ordinate_PCoA_bray = ordinate(btphylo_atsacrifice_pruned_transformed_smallintestine, 
                                             method = "PCoA", 
                                             distance = distance_bray_smallintestine)


PCoA_plot_smallintestine <- plot_ordination(btphylo_atsacrifice_pruned_transformed_smallintestine, smallintestine_ordinate_PCoA_bray, color = "Treatment")+
  theme_bw()+
  stat_ellipse() #adds an ellipse to a scatter plot to help visualize the distribution of points in the plot. The ellipse is drawn using a confidence interval, which can be specified using the level argument. By default, stat_ellipse() uses a 95% confidence interval.
PCoA_plot_smallintestine+
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange"))

ord_smallintestine_df <- data.frame(sample_data(btphylo_atsacrifice_pruned_transformed_smallintestine))
ord_smallintestine_df$SeqCoastOrder_Cage_Mouse <- interaction(ord_smallintestine_df$SeqCoastOrder, ord_smallintestine_df$Cage, ord_smallintestine_df$Mouse, drop = TRUE)
ord_smallintestine_df$SeqCoastOrder_Cage <- interaction(ord_smallintestine_df$SeqCoastOrder, ord_smallintestine_df$Cage, drop = TRUE)

# Run PERMANOVA using adonis2
smallintestine_perm_results <- adonis2(distance_bray_smallintestine ~ Treatment + Sex + Treatment*Sex + Microniche + SeqCoastOrder_Cage_Mouse+SeqCoastOrder:Cage, 
                                       data = ord_smallintestine_df, 
                                       method="bray", 
                                       permutations = 999,
                                       by="terms")
 

smallintestine_perm_results



# Run pairwise Adonis
smallintestine_pairwise_results <- pairwise.adonis(
  x = distance_bray_smallintestine,                  # Bray-Curtis distance matrix
  factors = ord_smallintestine_df$Treatment,         # The grouping factor (Treatment)
  sim.function = "vegdist",                           # Use vegan distance function
  sim.method = "bray",                               # Bray-Curtis method
  p.adjust.m = "bonferroni",                         # Adjust p-values using Bonferroni correction
  perm = 999                                         # Number of permutations
)

# Print the results
print(smallintestine_pairwise_results)

```
#run stats on beta diversity in Cecum
```{r}
# Subset phyloseq object to only have the data from Cecum
btphylo_atsacrifice_pruned_transformed_cecum <- subset_samples(btphylo_atsacrifice_pruned_transformed, 
                                                              Biogeography %in% "Cecum")

# Calculate Bray-Curtis dissimilarity distance matrix for Cecum
distance_bray_cecum = phyloseq::distance(btphylo_atsacrifice_pruned_transformed_cecum, method = "bray")

# Perform a principal coordinate analysis (PCoA) on the distance matrix for Cecum
cecum_ordinate_PCoA_bray = ordinate(btphylo_atsacrifice_pruned_transformed_cecum, 
                                     method = "PCoA", 
                                     distance = distance_bray_cecum)

# Create PCoA plot for Cecum
PCoA_plot_cecum <- plot_ordination(btphylo_atsacrifice_pruned_transformed_cecum, cecum_ordinate_PCoA_bray, color = "Treatment") +
  theme_bw() +
  stat_ellipse() # adds an ellipse to the scatter plot for confidence intervals
PCoA_plot_cecum +
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange"))

# Convert sample data to a dataframe
ord_cecum_df <- data.frame(sample_data(btphylo_atsacrifice_pruned_transformed_cecum))
ord_cecum_df$SeqCoastOrder_Cage_Mouse <- interaction(ord_cecum_df$SeqCoastOrder, ord_cecum_df$Cage, ord_cecum_df$Mouse, drop = TRUE)
ord_cecum_df$SeqCoastOrder_Cage <- interaction(ord_cecum_df$SeqCoastOrder, ord_cecum_df$Cage, drop = TRUE)

# Run PERMANOVA using adonis2 for Cecum
cecum_perm_results <- adonis2(distance_bray_cecum ~ Treatment + Sex + Treatment*Sex+Microniche+SeqCoastOrder_Cage+SeqCoastOrder_Cage_Mouse, 
                              data = ord_cecum_df,
                              method="bray",
                              permutations = 999,
                              by="terms")
 

cecum_perm_results

# Run pairwise Adonis for Cecum
cecum_pairwise_results <- pairwise.adonis(
  x = distance_bray_cecum,                  # Bray-Curtis distance matrix
  factors = ord_cecum_df$Treatment,         # The grouping factor (Treatment)
  sim.function = "vegdist",                  # Use vegan distance function
  sim.method = "bray",                      # Bray-Curtis method
  p.adjust.m = "bonferroni",                # Adjust p-values using Bonferroni correction
  perm = 999                                # Number of permutations
)

# Print the results of pairwise Adonis for Cecum
print(cecum_pairwise_results)



```

#now run stats for beta diversity in proximal colon
```{r}
# Subset phyloseq object to only have the data from Proximal Colon
btphylo_atsacrifice_pruned_transformed_proximalcolon <- subset_samples(btphylo_atsacrifice_pruned_transformed, 
                                                                       Biogeography %in% "ProximalColon")

# Calculate Bray-Curtis dissimilarity distance matrix for Proximal Colon
distance_bray_proximalcolon = phyloseq::distance(btphylo_atsacrifice_pruned_transformed_proximalcolon, method = "bray")

# Perform a principal coordinate analysis (PCoA) on the distance matrix for Proximal Colon
proximalcolon_ordinate_PCoA_bray = ordinate(btphylo_atsacrifice_pruned_transformed_proximalcolon, 
                                            method = "PCoA", 
                                            distance = distance_bray_proximalcolon)

# Create PCoA plot for Proximal Colon
PCoA_plot_proximalcolon <- plot_ordination(btphylo_atsacrifice_pruned_transformed_proximalcolon, proximalcolon_ordinate_PCoA_bray, color = "Treatment") +
  theme_bw() +
  stat_ellipse() # adds an ellipse to the scatter plot for confidence intervals
PCoA_plot_proximalcolon +
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange"))

# Convert sample data to a dataframe
ord_proximalcolon_df <- data.frame(sample_data(btphylo_atsacrifice_pruned_transformed_proximalcolon))
ord_proximalcolon_df$SeqCoastOrder_Cage_Mouse <- interaction(ord_proximalcolon_df$SeqCoastOrder, ord_proximalcolon_df$Cage, ord_proximalcolon_df$Mouse, drop = TRUE)
ord_proximalcolon_df$SeqCoastOrder_Cage <- interaction(ord_proximalcolon_df$SeqCoastOrder, ord_proximalcolon_df$Cage, drop = TRUE)


# Run PERMANOVA using adonis2 for Proximal Colon
proximalcolon_perm_results <- adonis2(distance_bray_proximalcolon ~ Treatment + Sex + Treatment*Sex+Microniche+SeqCoastOrder_Cage+SeqCoastOrder_Cage_Mouse, 
                                      data = ord_proximalcolon_df, 
                                      permutations = 999,
                                      by="terms")
 

proximalcolon_perm_results

# Run pairwise Adonis for Proximal Colon
proximalcolon_pairwise_results <- pairwise.adonis(
  x = distance_bray_proximalcolon,                  # Bray-Curtis distance matrix
  factors = ord_proximalcolon_df$Treatment,         # The grouping factor (Treatment)
  sim.function = "vegdist",                         # Use vegan distance function
  sim.method = "bray",                              # Bray-Curtis method
  p.adjust.m = "bonferroni",                        # Adjust p-values using Bonferroni correction
  perm = 999                                        # Number of permutations
)

# Print the results of pairwise Adonis for Proximal Colon
print(proximalcolon_pairwise_results)



```
#Now stats for beta diversity in distal colon
```{r}
# Subset phyloseq object to only have the data from Distal Colon
btphylo_atsacrifice_pruned_transformed_distalcolon <- subset_samples(btphylo_atsacrifice_pruned_transformed, 
                                                                       Biogeography %in% "DistalColon")

# Calculate Bray-Curtis dissimilarity distance matrix for Distal Colon
distance_bray_distalcolon = phyloseq::distance(btphylo_atsacrifice_pruned_transformed_distalcolon, method = "bray")

# Perform a principal coordinate analysis (PCoA) on the distance matrix for Distal Colon
distalcolon_ordinate_PCoA_bray = ordinate(btphylo_atsacrifice_pruned_transformed_distalcolon, 
                                           method = "PCoA", 
                                           distance = distance_bray_distalcolon)

# Create PCoA plot for Distal Colon
PCoA_plot_distalcolon <- plot_ordination(btphylo_atsacrifice_pruned_transformed_distalcolon, distalcolon_ordinate_PCoA_bray, color = "Treatment") +
  theme_bw() +
  stat_ellipse() # adds an ellipse to the scatter plot for confidence intervals
PCoA_plot_distalcolon +
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange"))

# Convert sample data to a dataframe
ord_distalcolon_df <- data.frame(sample_data(btphylo_atsacrifice_pruned_transformed_distalcolon))
ord_distalcolon_df$SeqCoastOrder_Cage_Mouse <- interaction(ord_distalcolon_df$SeqCoastOrder, ord_distalcolon_df$Cage, ord_distalcolon_df$Mouse, drop = TRUE)
ord_distalcolon_df$SeqCoastOrder_Cage <- interaction(ord_distalcolon_df$SeqCoastOrder, ord_distalcolon_df$Cage, drop = TRUE)


# Run PERMANOVA using adonis2 for Distal Colon
distalcolon_perm_results <- adonis2(distance_bray_distalcolon ~ Treatment + Sex + Treatment*Sex+Microniche+SeqCoastOrder_Cage+SeqCoastOrder_Cage_Mouse, 
                                    data = ord_distalcolon_df, 
                                    permutations = 999,
                                      by="terms")
 

distalcolon_perm_results

# Run pairwise Adonis for Distal Colon
distalcolon_pairwise_results <- pairwise.adonis(
  x = distance_bray_distalcolon,                  # Bray-Curtis distance matrix
  factors = ord_distalcolon_df$Treatment,         # The grouping factor (Treatment)
  sim.function = "vegdist",                        # Use vegan distance function
  sim.method = "bray",                             # Bray-Curtis method
  p.adjust.m = "bonferroni",                       # Adjust p-values using Bonferroni correction
  perm = 999                                       # Number of permutations
)

# Print the results of pairwise Adonis for Distal Colon
print(distalcolon_pairwise_results)



```
#Mucusal Surface beta diversity pairwise adonis
```{r}
# Subset phyloseq object to only have the data from Mucosal Surface
btphylo_atsacrifice_pruned_transformed_mucosal <- subset_samples(btphylo_atsacrifice_pruned_transformed, 
                                                                Microniche %in% "MucosalSurface")

# Calculate Bray-Curtis dissimilarity distance matrix for Mucosal Surface
distance_bray_mucosal = phyloseq::distance(btphylo_atsacrifice_pruned_transformed_mucosal, method = "bray")

# Perform a principal coordinate analysis (PCoA) on the distance matrix for Mucosal Surface
mucosal_ordinate_PCoA_bray = ordinate(btphylo_atsacrifice_pruned_transformed_mucosal, 
                                       method = "PCoA", 
                                       distance = distance_bray_mucosal)

# Create PCoA plot for Mucosal Surface
PCoA_plot_mucosal <- plot_ordination(btphylo_atsacrifice_pruned_transformed_mucosal, mucosal_ordinate_PCoA_bray, color = "Treatment") +
  theme_bw() +
  stat_ellipse() # adds an ellipse to the scatter plot for confidence intervals
PCoA_plot_mucosal +
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange"))

# Convert sample data to a dataframe
ord_mucosal_df <- data.frame(sample_data(btphylo_atsacrifice_pruned_transformed_mucosal))
ord_mucosal_df$SeqCoastOrder_Cage_Mouse <- interaction(ord_mucosal_df$SeqCoastOrder, ord_mucosal_df$Cage, ord_mucosal_df$Mouse, drop = TRUE)
ord_mucosal_df$SeqCoastOrder_Cage <- interaction(ord_mucosal_df$SeqCoastOrder, ord_mucosal_df$Cage, drop = TRUE)



# Run PERMANOVA using adonis2 for Mucosal Surface
mucosal_perm_results <- adonis2(distance_bray_mucosal ~ Treatment + Sex + Biogeography + Treatment*Sex+SeqCoastOrder_Cage+SeqCoastOrder_Cage_Mouse,
                                data = ord_mucosal_df, 
                                permutations = 999,
                                by="terms")

# Print the results of PERMANOVA
print(mucosal_perm_results)

# Run pairwise Adonis for Mucosal Surface
mucosal_pairwise_results <- pairwise.adonis(
  x = distance_bray_mucosal,                        # Bray-Curtis distance matrix
  factors = ord_mucosal_df$Treatment,               # The grouping factor (Treatment)
  sim.function = "vegdist",                          # Use vegan distance function
  sim.method = "bray",                               # Bray-Curtis method
  p.adjust.m = "bonferroni",                         # Adjust p-values using Bonferroni correction
  perm = 999                                         # Number of permutations
)

# Print the results of pairwise Adonis for Mucosal Surface
print(mucosal_pairwise_results)




```
Luminal Contents beta diversity pairwise adonis
```{r}

# Subset phyloseq object to only have the data from Luminal Contents
btphylo_atsacrifice_pruned_transformed_luminal <- subset_samples(btphylo_atsacrifice_pruned_transformed, 
                                                                Microniche %in% "LuminalContents")

# Calculate Bray-Curtis dissimilarity distance matrix for Luminal Contents
distance_bray_luminal = phyloseq::distance(btphylo_atsacrifice_pruned_transformed_luminal, method = "bray")

# Perform a principal coordinate analysis (PCoA) on the distance matrix for Luminal Contents
luminal_ordinate_PCoA_bray = ordinate(btphylo_atsacrifice_pruned_transformed_luminal, 
                                       method = "PCoA", 
                                       distance = distance_bray_luminal)

# Create PCoA plot for Luminal Contents
PCoA_plot_luminal <- plot_ordination(btphylo_atsacrifice_pruned_transformed_luminal, luminal_ordinate_PCoA_bray, color = "Treatment") +
  theme_bw() +
  stat_ellipse() # adds an ellipse to the scatter plot for confidence intervals
PCoA_plot_luminal +
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange"))

# Convert sample data to a dataframe
ord_luminal_df <- data.frame(sample_data(btphylo_atsacrifice_pruned_transformed_luminal))
ord_luminal_df$SeqCoastOrder_Cage_Mouse <- interaction(ord_luminal_df$SeqCoastOrder, ord_luminal_df$Cage, ord_luminal_df$Mouse, drop = TRUE)
ord_luminal_df$SeqCoastOrder_Cage <- interaction(ord_luminal_df$SeqCoastOrder, ord_luminal_df$Cage, drop = TRUE)



# Run PERMANOVA using adonis2 for Luminal Contents
luminal_perm_results <- adonis2(distance_bray_luminal ~ Treatment + Sex + Biogeography + Treatment*Sex+SeqCoastOrder_Cage+SeqCoastOrder_Cage_Mouse, 
                                data = ord_luminal_df, 
                                permutations = 999,
                                by="terms")

# Print the results of PERMANOVA
print(luminal_perm_results)

# Run pairwise Adonis for Luminal Contents
luminal_pairwise_results <- pairwise.adonis(
  x = distance_bray_luminal,                        # Bray-Curtis distance matrix
  factors = ord_luminal_df$Treatment,               # The grouping factor (Treatment)
  sim.function = "vegdist",                          # Use vegan distance function
  sim.method = "bray",                               # Bray-Curtis method
  p.adjust.m = "bonferroni",                         # Adjust p-values using Bonferroni correction
  perm = 999                                         # Number of permutations
)

# Print the results of pairwise Adonis for Luminal Contents
print(luminal_pairwise_results)


```

Stool at sacrifice beta diversity pairwise adonis
```{r}

# Subset phyloseq object to only have the data from stool Contents
btphylo_atsacrifice_pruned_transformed_stool <- subset_samples(btphylo_atsacrifice_pruned_transformed, 
                                                                Microniche %in% "stool")

# Calculate Bray-Curtis dissimilarity distance matrix for stool Contents
distance_bray_stool = phyloseq::distance(btphylo_atsacrifice_pruned_transformed_stool, method = "bray")

# Perform a principal coordinate analysis (PCoA) on the distance matrix for stool Contents
stool_ordinate_PCoA_bray = ordinate(btphylo_atsacrifice_pruned_transformed_stool, 
                                       method = "PCoA", 
                                       distance = distance_bray_stool)

# Create PCoA plot for stool Contents
PCoA_plot_stool <- plot_ordination(btphylo_atsacrifice_pruned_transformed_stool, stool_ordinate_PCoA_bray, color = "Treatment", shape = "Sex") +
  theme_bw() +
  stat_ellipse(aes(group=Treatment)) # adds an ellipse to the scatter plot for confidence intervals
PCoA_plot_stool +
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange"))

# Convert sample data to a dataframe
ord_stool_df <- data.frame(sample_data(btphylo_atsacrifice_pruned_transformed_stool))
ord_stool_df$SeqCoastOrder_Cage_Mouse <- interaction(ord_stool_df$SeqCoastOrder, ord_stool_df$Cage, ord_stool_df$Mouse, drop = TRUE)
ord_stool_df$SeqCoastOrder_Cage <- interaction(ord_stool_df$SeqCoastOrder, ord_stool_df$Cage, drop = TRUE)


# Run PERMANOVA using adonis2 for stool Contents
stool_perm_results <- adonis2(distance_bray_stool ~ Treatment + Sex+Treatment*Sex+SeqCoastOrder_Cage+Seq, 
                              data = ord_stool_df, 
                              permutations = 999,
                                by="terms")


# Print the results of PERMANOVA
print(stool_perm_results)

# Run pairwise Adonis for stool Contents
stool_pairwise_results <- pairwise.adonis(
  x = distance_bray_stool,                        # Bray-Curtis distance matrix
  factors = ord_stool_df$Treatment,               # The grouping factor (Treatment)
  sim.function = "vegdist",                          # Use vegan distance function
  sim.method = "bray",                               # Bray-Curtis method
  p.adjust.m = "bonferroni",                         # Adjust p-values using Bonferroni correction
  perm = 999                                         # Number of permutations
)

# Print the results of pairwise Adonis for stool Contents
print(stool_pairwise_results)


```


```{r}
#######################Baseline########################
#Subset phyloseq object to only have the data from Baseline
btphylo_atsacrifice_pruned_transformed_baseline <- subset_samples(btphylo_atsacrifice_pruned_transformed, 
                                                               Treatment == c( "Baseline"))

#Calculate Bray-Curtis dissimilarity distance matrix
distance_bray_baseline = phyloseq::distance(btphylo_atsacrifice_pruned_transformed_baseline, method = "bray")

#Perform a principal coordinate analysis (PCoA) on the distance matrix
baseline_ordinate_PCoA_bray = ordinate(btphylo_atsacrifice_pruned_transformed_baseline, method = "PCoA", distance = "distance_bray_baseline")

PCoA_plot_baseline <- plot_ordination(btphylo_atsacrifice_pruned_transformed_baseline, baseline_ordinate_PCoA_bray, shape = "Sex", color = "Microniche")+
  theme_bw()+
  stat_ellipse()+ #adds an ellipse to a scatter plot to help visualize the distribution of points in the plot. The ellipse is drawn using a confidence interval, which can be specified using the level argument. By default, stat_ellipse() uses a 95% confidence interval.
   scale_color_manual(values = pastel_colors)  # Apply pastel colors

ord_Baseline_df <- data.frame(sample_data(btphylo_atsacrifice_pruned_transformed_baseline))

# Run PERMANOVA using adonis2
Baseline_perm_results <- adonis2(distance_bray_baseline ~ Microniche + Sex, data = ord_Baseline_df, permutations = 999)

Baseline_perm_results


#######################PBS########################
#Subset phyloseq object to only have the data from PBS
btphylo_atsacrifice_pruned_transformed_PBS <- subset_samples(btphylo_atsacrifice_pruned_transformed, 
                                                               Treatment == c( "PBS"))

#Calculate Bray-Curtis dissimilarity distance matrix
distance_bray_PBS = phyloseq::distance(btphylo_atsacrifice_pruned_transformed_PBS, method = "bray")

#Perform a principal coordinate analysis (PCoA) on the distance matrix
PBS_ordinate_PCoA_bray = ordinate(btphylo_atsacrifice_pruned_transformed_PBS, method = "PCoA", distance = "distance_bray_PBS")

PCoA_plot_PBS <- plot_ordination(btphylo_atsacrifice_pruned_transformed_PBS, PBS_ordinate_PCoA_bray, shape = "Sex", color = "Microniche")+
  theme_bw()+
  stat_ellipse()+ #adds an ellipse to a scatter plot to help visualize the distribution of points in the plot. The ellipse is drawn using a confidence interval, which can be specified using the level argument. By default, stat_ellipse() uses a 95% confidence interval.
   scale_color_manual(values = pastel_colors)  # Apply pastel colors

ord_PBS_df <- data.frame(sample_data(btphylo_atsacrifice_pruned_transformed_PBS))

# Run PERMANOVA using adonis2
PBS_perm_results <- adonis2(distance_bray_PBS ~ Microniche + Sex, data = ord_PBS_df, permutations = 999)

PBS_perm_results

#######################PBS########################
#Subset phyloseq object to only have the data from Bacteroides
btphylo_atsacrifice_pruned_transformed_Bacteroides <- subset_samples(btphylo_atsacrifice_pruned_transformed, 
                                                               Treatment == c( "Bacteroides"))

#Calculate Bray-Curtis dissimilarity distance matrix
distance_bray_Bacteroides = phyloseq::distance(btphylo_atsacrifice_pruned_transformed_Bacteroides, method = "bray")

#Perform a principal coordinate analysis (PCoA) on the distance matrix
Bacteroides_ordinate_PCoA_bray = ordinate(btphylo_atsacrifice_pruned_transformed_Bacteroides, method = "PCoA", distance = "distance_bray_Bacteroides")

PCoA_plot_Bacteroides <- plot_ordination(btphylo_atsacrifice_pruned_transformed_Bacteroides, Bacteroides_ordinate_PCoA_bray, shape = "Sex", color = "Microniche")+
  theme_bw()+
  stat_ellipse()+ #adds an ellipse to a scatter plot to help visualize the distribution of points in the plot. The ellipse is drawn using a confidence interval, which can be specified using the level argument. By default, stat_ellipse() uses a 95% confidence interval.
   scale_color_manual(values = pastel_colors)  # Apply pastel colors

ord_Bacteroides_df <- data.frame(sample_data(btphylo_atsacrifice_pruned_transformed_Bacteroides))

# Run PERMANOVA using adonis2
Bacteroides_perm_results <- adonis2(distance_bray_Bacteroides ~ Microniche + Sex, data = ord_Bacteroides_df, permutations = 999)

Bacteroides_perm_results

```


```{r}
# Access the taxonomy table
tax_table <- as.data.frame(tax_table(btphylo_nosacrifice))

# Extract the genera
genera <- tax_table$Genus

# Print unique genera
unique_genera <- unique(genera)
print(unique_genera)


```


#Plot relative abundance across gut
```{r}

#filter and Normalize the data for across the gut
wh02 = genefilter_sample(btphylo_atsacrifice, filterfun_sample(function(x) x > 1), A=0.1*nsamples(btphylo_atsacrifice))


btphylo_atsacrifice_pruned = prune_taxa(wh02, btphylo_atsacrifice)


#Aggregate Data at the Genus Level: You'll need to aggregate your OTU table at the genus level.
btphylo_atsacrifice_pruned_genus <- tax_glom(btphylo_atsacrifice_pruned, "Genus")

#Transform relative abundance
btphylo_atsacrifice_pruned_genus_norm100 <- transform_sample_counts(btphylo_atsacrifice_pruned_genus, function(x) x / sum(x) *100)

btphylo_atsacrifice_pruned_genus_norm100_relab_escherichia <- subset_taxa(btphylo_atsacrifice_pruned_genus_norm100, Genus == "Escherichia-Shigella")

bt_genus_acrossgut_relab_escherichia_df <- psmelt(btphylo_atsacrifice_pruned_genus_norm100_relab_escherichia)

bt_genus_acrossgut_relab_escherichia_lumen_df <- subset(bt_genus_acrossgut_relab_escherichia_df, Microniche == "LuminalContents")
bt_genus_acrossgut_relab_escherichia_mucosa_df <- subset(bt_genus_acrossgut_relab_escherichia_df, Microniche == "MucosalSurface")


escherichia_lumen_16s_plot_smooth_connected_points <- ggplot(bt_genus_acrossgut_relab_escherichia_lumen_df, aes(x = Biogeography, y = Abundance, color = Treatment, shape = Sex)) +
  geom_point() +
  geom_line(aes(group = interaction(Cage,Mouse)), size = 0.25) +  # Add this line to connect points for each MouseID
  stat_summary(fun.y = mean, size=2, geom = "line", aes(group = Treatment)) +
  scale_y_continuous(trans = 'identity') +
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +
  theme_bw() +
  labs(y = "Relative Abundance of Escherichia (%)", x = "Experiment Day", color = "Treatment Group")

escherichia_lumen_16s_plot_smooth_connected_points


ggsave("Ecoli_acrossgut_lumen.pdf", plot = escherichia_lumen_16s_plot_smooth_connected_points, device = "pdf", width = 6, height = 3)

escherichia_mucosa_16s_plot_smooth_connected_points <- ggplot(bt_genus_acrossgut_relab_escherichia_mucosa_df, aes(x = Biogeography, y = Abundance, color = Treatment, shape = Sex)) +
    geom_point(size = 3) +  # Set size of the points
   geom_line(aes(group = interaction(Cage,Mouse_.)), size = 0.25) +  # Add this line to connect points for each MouseID
  stat_summary(fun.y = mean, size=2, geom = "line", aes(group = Treatment)) +
    stat_summary(fun = median, geom = "crossbar", aes(group = Treatment), position = position_dodge(width = 0.5), width = 0.5, fatten = 3) +  # Add median crossbars, position_dodge to separate them
    scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +  # Consistent color scheme
    theme_bw() +
    labs(title = "Relative Abundance of Escherichia Across the Gut on the mucosa",
         x = "Biogeography",
         y = "Relative Abundance of Escherichia (%)")

escherichia_mucosa_16s_plot_smooth_connected_points

ggsave("Ecoli_acrossgut_mucoosa.pdf", plot = escherichia_mucosa_16s_plot_smooth_connected_points, device = "pdf", width = 6, height = 3)


```

#Plot relative abundance across the gut
```{r}

btphylo_atsacrifice

#filter and Normalize the data for across the gut
wh02 = genefilter_sample(btphylo_atsacrifice, filterfun_sample(function(x) x > 1), A=0.1*nsamples(btphylo_atsacrifice))


btphylo_atsacrifice_pruned = prune_taxa(wh02, btphylo_atsacrifice)


#Aggregate Data at the Genus Level: You'll need to aggregate your OTU table at the genus level.
btphylo_atsacrifice_pruned_genus <- tax_glom(btphylo_atsacrifice_pruned, "Genus")

#Transform relative abundance
btphylo_atsacrifice_pruned_genus_relab <- transform_sample_counts(btphylo_atsacrifice_pruned_genus, function(x) x / sum(x) *100)

#Extract and Prepare Data for Plotting:

btphylo_atsacrifice_pruned_genus_relab_df <- psmelt(btphylo_atsacrifice_pruned_genus_relab)
btphylo_atsacrifice_pruned_genus_relab_df_bacteroides <- btphylo_atsacrifice_pruned_genus_relab_df %>% filter(Genus == "Bacteroides")

# Assuming YourDataFrame has a column named "Microniche"
btphylo_atsacrifice_pruned_genus_relab_df_bacteroides_lumen <- btphylo_atsacrifice_pruned_genus_relab_df_bacteroides %>%
    filter(Microniche == "LuminalContents")

btphylo_atsacrifice_pruned_genus_relab_df_bacteroides_mucosalsurface <- btphylo_atsacrifice_pruned_genus_relab_df_bacteroides %>%
    filter(Microniche == "MucosalSurface")

##Plot across the gut

Bt_16s_plot_acrossgut_biogeography_lumen <- ggplot(btphylo_atsacrifice_pruned_genus_relab_df_bacteroides_lumen, aes(x = Biogeography, y = Abundance, color = Treatment, shape = Sex)) +
    geom_point(size = 3) +  # Set size of the points
   geom_line(aes(group = interaction(Cage,Mouse)), size = 0.25) +  # Add this line to connect points for each MouseID
  stat_summary(fun.y = mean, size=2, geom = "line", aes(group = Treatment)) +
    stat_summary(fun = median, geom = "crossbar", aes(group = Treatment), position = position_dodge(width = 0.5), width = 0.5, fatten = 3) +  # Add median crossbars, position_dodge to separate them
    scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +  # Consistent color scheme
    theme_bw() +
    labs(x = "Biogeography",
         y = "Relative Abundance of Bacteroides (%)")+
  coord_cartesian(ylim = c(0, 70))  # Zoom into y-axis range


Bt_16s_plot_acrossgut_biogeography_lumen

ggsave("Bt_16s_plot_acrossgut_biogeography_lumen.pdf", plot = Bt_16s_plot_acrossgut_biogeography_lumen, device = "pdf", width = 6, height = 3)


#stats for bacteroides across gut in the lumen

btphylo_atsacrifice_pruned_genus_relab_df_bacteroides_lumen$MouseID <- paste(btphylo_atsacrifice_pruned_genus_relab_df_bacteroides_lumen$Cage, btphylo_atsacrifice_pruned_genus_relab_df_bacteroides_lumen$Mouse_., sep = "_")


btphylo_atsacrifice_pruned_genus_relab_df_bacteroides_lumen$Treatment <- factor(btphylo_atsacrifice_pruned_genus_relab_df_bacteroides_lumen$Treatment, levels=c("PBS", "Bacteroides", "Baseline"))

bacteroides_relabundance_acrossgut_lumen_model <- lmer(Abundance ~ Treatment + Biogeography + Sex + Treatment*Biogeography+(1|MouseID), data = btphylo_atsacrifice_pruned_genus_relab_df_bacteroides_lumen)

summary(bacteroides_relabundance_acrossgut_lumen_model)


```

#Plot across Microniches
```{r}


Bt_16s_plot_acrossgut_biogeography_mucosa <- ggplot(btphylo_atsacrifice_pruned_genus_relab_df_bacteroides_mucosalsurface, aes(x = Biogeography, y = Abundance, color = Treatment, shape = Sex)) +
    geom_point(size = 3) +  # Set size of the points
   geom_line(aes(group = interaction(Cage,Mouse_.)), size = 0.25) +  # Add this line to connect points for each MouseID
  stat_summary(fun.y = mean, size=2, geom = "line", aes(group = Treatment)) +
    stat_summary(fun = median, geom = "crossbar", aes(group = Treatment), position = position_dodge(width = 0.5), width = 0.5, fatten = 3) +  # Add median crossbars, position_dodge to separate them
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange")) +  # Consistent color scheme
    theme_bw() +
    labs(x = "Biogeography",
         y = "Relative Abundance of Bacteroides (%)")+
    coord_cartesian(ylim = c(0, 70))  # Zoom into y-axis range


Bt_16s_plot_acrossgut_biogeography_mucosa

ggsave("Bt_16s_plot_acrossgut_biogeography_mucosa.pdf", plot = Bt_16s_plot_acrossgut_biogeography_mucosa, device = "pdf", width = 6, height = 3)

#stats for bacteroides across gut in the lumen

btphylo_atsacrifice_pruned_genus_relab_df_bacteroides_mucosalsurface$MouseID <- paste(btphylo_atsacrifice_pruned_genus_relab_df_bacteroides_mucosalsurface$Cage, btphylo_atsacrifice_pruned_genus_relab_df_bacteroides_mucosalsurface$Mouse_., sep = "_")


btphylo_atsacrifice_pruned_genus_relab_df_bacteroides_mucosalsurface$Treatment <- factor(btphylo_atsacrifice_pruned_genus_relab_df_bacteroides_mucosalsurface$Treatment, levels=c("Baseline", "PBS", "Bacteroides"))


btphylo_atsacrifice_pruned_genus_relab_df_bacteroides_mucosalsurface_model <- lmer(Abundance ~ Treatment + Biogeography + Sex + Treatment*Biogeography + (1|MouseID), data = btphylo_atsacrifice_pruned_genus_relab_df_bacteroides_mucosalsurface)

summary(btphylo_atsacrifice_pruned_genus_relab_df_bacteroides_mucosalsurface_model)





```


#Plot beta diversity between sexes
```{r}
# Compute ordination (PCoA) using a distance matrix (e.g., Morisita Horn)

btphylo_atsacrifice

btphylo_atsacrifice_normalized <- transform_sample_counts(btphylo_atsacrifice, function(x) x / sum(x) *100)

btphylo_atsacrifice_normalized_horn <- phyloseq::distance(btphylo_atsacrifice_normalized, method = "horn")

btphylo_atsacrifice_normalized_ord_pcoa_horn <- ordinate(btphylo_atsacrifice_normalized, method = "PCoA", distance = "horn")

# Plot the beta diversity between sexes, all together
betahorn_diversity_acrossgut_lumen_plot1 <- plot_ordination(btphylo_atsacrifice_normalized, btphylo_atsacrifice_normalized_ord_pcoa_horn, color = "Sex", shape = "Sex") +
  geom_point(size = 3) +  # Adjust size of points
  theme_bw() +
  scale_color_manual(values = c("F" = "lightgreen", "M" = "lightblue"))+
  stat_ellipse(aes(group = Sex), linetype = 1)


betahorn_diversity_acrossgut_lumen_plot1

ggsave("betahorn_diversity_betwensexes_lumen_plot.pdf", plot = betahorn_diversity_acrossgut_lumen_plot1, device = "pdf", width = 6, height = 4)

# Plot the beta diversity between sexes, for each treatment
betahorn_diversity_acrossgut_lumen_plot2 <- plot_ordination(btEstablishmentWTmice_phyloseq_nobaselinemice_atsacrifice_lumen, ord_pcoa_horn, color = "Sex", shape = "Sex") +
  geom_point(size = 3) +  # Adjust size of points
  theme_bw() +
  facet_wrap(~Treatment)+
  scale_color_manual(values = c("F" = "lightgreen", "M" = "lightblue"))+
  stat_ellipse(aes(group = Sex), linetype = 1)


betahorn_diversity_acrossgut_lumen_plot2

ggsave("betahorn_diversity_betwensexes_lumen_plot2.pdf", plot = betahorn_diversity_acrossgut_lumen_plot2, device = "pdf", width = 6, height = 4)

# Compute ordination (PCoA) using a distance matrix (e.g., Bray-Curtis)
ord_pcoa_bray <- ordinate(btEstablishmentWTmice_phyloseq_nobaselinemice_atsacrifice_lumen, method = "PCoA", distance = "bray")

# Plot the beta diversity between sexes, all together
betabray_diversity_acrossgut_lumen_plot3 <- plot_ordination(btEstablishmentWTmice_phyloseq_nobaselinemice_atsacrifice_lumen, ord_pcoa_bray, color = "Sex", shape = "Sex") +
  geom_point(size = 3) +  # Adjust size of points
  theme_bw() +
  scale_color_manual(values = c("F" = "lightgreen", "M" = "lightblue"))+
  stat_ellipse(aes(group = Sex), linetype = 1)


betabray_diversity_acrossgut_lumen_plot3

ggsave("betabray_diversity_betwensexes_lumen_plot3.pdf", plot = betabray_diversity_acrossgut_lumen_plot3, device = "pdf", width = 6, height = 4)

# Plot the beta diversity between sexes, for each treatment
betabray_diversity_acrossgut_lumen_plot4 <- plot_ordination(btEstablishmentWTmice_phyloseq_nobaselinemice_atsacrifice_lumen, ord_pcoa_bray, color = "Sex", shape = "Sex") +
  geom_point(size = 3) +  # Adjust size of points
  theme_bw() +
  facet_wrap(~Treatment)+
  scale_color_manual(values = c("F" = "lightgreen", "M" = "lightblue"))+
  stat_ellipse(aes(group = Sex), linetype = 1)


betahorn_diversity_acrossgut_lumen_plot4

ggsave("betabray_diversity_betwensexes_lumen_plot4.pdf", plot = betabray_diversity_acrossgut_lumen_plot4, device = "pdf", width = 6, height = 4)
```

```{r}

# Plot the beta diversity with regions all togther but treatments separated
betahorn_diversity_acrossgut_lumen_plot2 <- plot_ordination(btEstablishmentWTmice_phyloseq_nobaselinemice_atsacrifice_lumen, ord_pcoa_horn, color = "Biogeography", shape = "Sex") +
  geom_point(size = 3) +  # Adjust size of points
  theme_bw() +
  facet_wrap(~Treatment, nrow = 1, ncol = 3) +  # Facet by Biogeography with specified layout
  scale_color_manual(values = c("SmallIntestine" = "darkblue", "Cecum" = "royalblue", "ProximalColon" = "lightblue", "DistalColon" = "#D1E8EE"))+
  stat_ellipse(aes(group = Biogeography), linetype = 1)


betahorn_diversity_acrossgut_lumen_plot2

ggsave("betahorn_diversity_acrossgut_pertreatment_lumen_plot.pdf", plot = betahorn_diversity_acrossgut_lumen_plot2, device = "pdf", width = 8, height = 2)


# Plot the beta diversity with biogeography faceted
betahorn_diversity_acrossgut_lumen_plot <- plot_ordination(btEstablishmentWTmice_phyloseq_nobaselinemice_atsacrifice_lumen, ord_pcoa_horn, color = "Treatment", shape = "Sex") +
  geom_point(size = 3) +  # Adjust size of points
  facet_wrap(~Biogeography, nrow = 1, ncol = 4) +  # Facet by Biogeography with specified layout
  theme_bw() +
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange"))+
  stat_ellipse(aes(group = Treatment), linetype = 1)


betahorn_diversity_acrossgut_lumen_plot

ggsave("betahorn_diversity_acrossgut_lumen_plot.pdf", plot = betahorn_diversity_acrossgut_lumen_plot, device = "pdf", width = 8, height = 2)



# Compute ordination (PCoA) using a distance matrix (e.g., Morisita)

ord_pcoa_bray <- ordinate(btEstablishmentWTmice_phyloseq_nobaselinemice_atsacrifice_lumen, method = "PCoA", distance = "bray")

# Plot the beta diversity
betabray_diversity_acrossgut_lumen_plot <- plot_ordination(btEstablishmentWTmice_phyloseq_nobaselinemice_atsacrifice_lumen, ord_pcoa_bray, color = "Treatment", shape = "Sex") +
  geom_point(size = 3) +  # Adjust size of points
  theme_bw() +
  facet_wrap(~Biogeography, nrow = 1, ncol = 4) +  # Facet by Biogeography with specified layout
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange"))+
  stat_ellipse(aes(group = Treatment), linetype = 1)


betabray_diversity_acrossgut_lumen_plot


```

#Plot beta diversity across the gut, coloring by treatment group

```{r}
desired_timepoints <- c("Sacrifice", "LPS", "Baseline_6")
desired_treatments <- c("Baseline")

phylo_LPSandSacrifice <- subset_samples(btEstablishmentWTmice_phyloseq, 
                                        Timepoint %in% desired_timepoints & 
                                        (Treatment %in% desired_treatments | Timepoint != "Baseline_6"))

# Extract the sample data from the phyloseq object
sample_data_df <- as(sample_data(phylo_LPSandSacrifice), "data.frame")

sample_data_df <- sample_data_df %>% 
  mutate(Biogeography = case_when(
    is.na(Biogeography) ~ "Fecal",
    TRUE ~ Biogeography  # Keep all other values as is
  ))

#Factor order of biogrography
sample_data_df$Biogeography <- factor(sample_data_df$Biogeography, levels = c("SmallIntestine", "Cecum", "ProximalColon", "DistalColon", "Fecal"))


# Update the sample data in the phyloseq object
sample_data(phylo_LPSandSacrifice) <- sample_data_df

#filter the data for beta diverity analysis


#selecting taxa that are present in more than 1 count in at least 10% of the samples. This is a common way to filter out low-abundance taxa that might be less informative or represent sequencing noise
wh03 = genefilter_sample(phylo_LPSandSacrifice, filterfun_sample(function(x) x > 1), A=0.1*nsamples(phylo_LPSandSacrifice))


#removes samples that have zero total counts. 
phylo_LPSandSacrifice_toprune <- prune_samples(sample_sums(phylo_LPSandSacrifice) > 0, phylo_LPSandSacrifice)

#applying the filter criteria from step 1 to prune the taxa from the phylo_LPSandSacrifice_toprune object. This results in phylo_LPSandSacrifice_pruned, which contains only the taxa that meet your filtering criteria
phylo_LPSandSacrifice_pruned = prune_taxa(wh03, phylo_LPSandSacrifice_toprune)




#This is the ordination with the normalized data. You usually shouldn't calculate beta diversity distances on normalized data.
phylo_LPSandSacrifice_pruned_ord_horn = ordinate(phylo_LPSandSacrifice_pruned, method="PCoA", distance = "horn")

#PCoA is 35.3%


phylo_LPSandSacrifice_beta_horn <- plot_ordination(phylo_LPSandSacrifice, phylo_LPSandSacrifice_pruned_ord_horn, color = "Biogeography", shape = "Sex") +
  geom_point(size = 2) +  # Adjust size of points
  theme_bw() +
  facet_wrap(~Treatment, nrow = 1, ncol = 3) +  # Facet by Biogeography with specified layout
  scale_color_manual(values = c("Fecal" = "darkgreen", "SmallIntestine" = "darkblue", "Cecum" = "royalblue", "ProximalColon" = "lightblue", "DistalColon" = "#D1E8EE"))+
  stat_ellipse(aes(group = Biogeography), linetype = 1)

phylo_LPSandSacrifice_beta_horn

ggsave("phylo_LPSandSacrifice_beta_horn_acrosstreatment.pdf", width = 10, height = 3)


phylo_LPSandSacrifice_no_fecal <- subset_samples(phylo_LPSandSacrifice, Biogeography != "Fecal")
phylo_LPSandSacrifice_no_fecal_ord_horn <- ordinate(phylo_LPSandSacrifice_no_fecal, method = "PCoA", distance = "horn")

phylo_LPSandSacrifice_acrossgut_beta_horn <- plot_ordination(phylo_LPSandSacrifice_no_fecal, phylo_LPSandSacrifice_no_fecal_ord_horn, color = "Treatment", shape = "Microniche") +
  geom_jitter(size = 2) +  # Add jitter to points
  theme_bw() +
  facet_wrap(~Biogeography, nrow = 1, ncol = 4) +  # Facet by Biogeography with specified layout
  scale_color_manual(values = c("Baseline" = "lightgreen", "PBS" = "lightblue", "Bacteroides" = "orange"))+
  stat_ellipse(aes(group = Treatment), linetype = 1)

phylo_LPSandSacrifice_acrossgut_beta_horn

ggsave("phylo_LPSandSacrifice_beta_horn_acrossgut.pdf", width = 10, height = 3)


```



##### Running statistics forphylo_LPSandSacrifice_acrossgut_beta_horn
```{r}


# Assuming phylo_LPSandSacrifice_pruned is your phyloseq object
biogeographies <- c("SmallIntestine", "Cecum", "ProximalColon", "DistalColon")
results <- list()
pairwise_results <- list()


for (biogeo in biogeographies) {
  # Subset the phyloseq object
  phy_subset <- subset_samples(phylo_LPSandSacrifice_pruned, Biogeography == biogeo)
  
  # Create a distance matrix
  dist_matrix <- distance(phy_subset, method = "horn") # Replace 'bray' with your preferred distance method

  # Convert sample data to a data frame and ensure the presence of Treatment and Sex columns
  df_sample_data <- data.frame(sample_data(phy_subset))

  # Check if 'Treatment' and 'Sex' columns exist in the data frame
  if(!("Treatment" %in% colnames(df_sample_data)) || !("Sex" %in% colnames(df_sample_data))) {
    stop("The required columns 'Treatment' and 'Sex' are not present in the sample data.")
  }

  # Run PERMANOVA using adonis2
  perm_result <- adonis2(dist_matrix ~ Treatment + Sex + Microniche, data = df_sample_data, permutations = 999)
  results[[biogeo]] <- perm_result
  
   # Perform pairwise Adonis
  pairwise_result <- pairwise.adonis(dist_matrix, df_sample_data$Treatment)
  pairwise_results[[biogeo]] <- pairwise_result
}

# View results
results
pairwise_results

```

#Run another permanova for across treatments
```{r}

desired_timepoints2 <- c("Sacrifice", "LPS")

phylo_LPSandSacrifice2 <- subset_samples(btEstablishmentWTmice_phyloseq, 
                                        Timepoint %in% desired_timepoints2)
sample_data_df <- sample_data_df %>% 
  mutate(Biogeography = case_when(
    is.na(Biogeography) ~ "Fecal",
    TRUE ~ Biogeography  # Keep all other values as is
  ))

#Factor order of biogrography
sample_data_df$Biogeography <- factor(sample_data_df$Biogeography, levels = c("SmallIntestine", "Cecum", "ProximalColon", "DistalColon", "Fecal"))


# Update the sample data in the phyloseq object
sample_data(phylo_LPSandSacrifice2) <- sample_data_df

#filter the data for beta diverity analysis


#selecting taxa that are present in more than 1 count in at least 10% of the samples. This is a common way to filter out low-abundance taxa that might be less informative or represent sequencing noise
wh04 = genefilter_sample(phylo_LPSandSacrifice2, filterfun_sample(function(x) x > 1), A=0.1*nsamples(phylo_LPSandSacrifice))


#removes samples that have zero total counts. 
phylo_phylo_LPSandSacrifice2_toprune <- prune_samples(sample_sums(phylo_LPSandSacrifice2) > 0, phylo_LPSandSacrifice2)

#applying the filter criteria from step 1 to prune the taxa from the phylo_LPSandSacrifice_toprune object. This results in phylo_LPSandSacrifice_pruned, which contains only the taxa that meet your filtering criteria
phylo_phylo_LPSandSacrifice2_pruned = prune_taxa(wh04, phylo_phylo_LPSandSacrifice2_toprune)

library(phyloseq)
library(vegan)  # For adonis2
# Assuming the pairwise.adonis function is correctly loaded from its source

Treatments <- c("Baseline", "PBS", "Bacteroides")
results2 <- list()
pairwise_results <- list()

for (Treat in Treatments) {
  # Subset the phyloseq object based on the Treatment
  phy_subset <- subset_samples(phylo_phylo_LPSandSacrifice2_pruned, Treatment == Treat)

  # Ensure there are enough samples to analyze
  if (nsamples(phy_subset) > 1) {
    # Convert sample data to a data frame
    df_sample_data <- as.data.frame(sample_data(phy_subset))

    # Check if the necessary columns are present
    if (!all(c("Biogeography", "Sex", "Microniche") %in% colnames(df_sample_data))) {
      stop("One or more required columns (Biogeography, Sex, Microniche) are missing in the sample data.")
    }

    # Create a distance matrix
    dist_matrix <- phyloseq::distance(phy_subset, method = "horn")

    # Run PERMANOVA using adonis2
    perm_result <- adonis2(dist_matrix ~ Biogeography + Sex + Microniche, 
                           data = df_sample_data, permutations = 999)
    results[[Treat]] <- perm_result

    # Perform pairwise Adonis - modify as per your actual requirements
    pairwise_result2 <- pairwise.adonis(dist_matrix, df_sample_data$Biogeography)
    pairwise_results2[[Treat]] <- pairwise_result2
  } else {
    cat("Not enough samples found for Treatment:", Treat, "\n")
  }
}

# View results
print(results2)
print(pairwise_results)



```



#Run an analysis that identifies what microbes are most different between each region
```{r}

library(phyloseq)

 (!requireNamespace("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
  
BiocManager::install("ANCOMBC", force = TRUE)

library(ANCOMBC)
library(dplyr)
library(nloptr)

#loading example data so that i can see what this looks like
data(dietswap)

#Data from samples collected across gut at sacrifice
btphylo_atsacrifice <- subset_samples(btEstablishmentWTmice_phyloseq, Timepoint == "Sacrifice")

#wh0_btphylo_atsacrifice = genefilter_sample(btphylo_atsacrifice, filterfun_sample(function(x) x > 1), A=0.1*nsamples(btphylo_atsacrifice))

#btphylo_atsacrifice_pruned = prune_taxa(wh0_btphylo_atsacrifice, btphylo_atsacrifice)


btphylo_atsacrifice_relab100<- transform_sample_counts(btphylo_atsacrifice, function(x) x / sum(x) *100)

# Ensure that the taxonomy table has a 'Genus' level
if ("Genus" %in% rank_names(btphylo_atsacrifice_relab100)) {
    # Aggregate by Genus
    btphylo_atsacrifice_relab100_genus <- tax_glom(btphylo_atsacrifice_relab100, "Genus")
} else {
    stop("The taxonomy table does not contain a 'Genus' level. Please check your data.")
}

# Extract the data from the phyloseq object
ASV_data <- as(otu_table(btphylo_atsacrifice_relab100_genus), "matrix")
tax_table <- as.data.frame(tax_table(btphylo_atsacrifice_relab100_genus))
sample_data <- sample_data(btphylo_atsacrifice_relab100_genus)

# Convert data to a dataframe for ANCOMBC
ASV_data_df <- data.frame(t(ASV_data))

# Check if all ASVs in df_data are present in tax_table
if (!all(colnames(ASV_data_df) %in% rownames(tax_table))) {
    stop("Some ASVs in df_data are not present in tax_table.")
}

genus_names <- tax_table[match(colnames(ASV_data_df), rownames(tax_table)), "Genus"]

# Replace missing or NA genus names with a placeholder or unique identifier
genus_names[is.na(genus_names)] <- paste("NA", seq(along=genus_names[is.na(genus_names)]), sep="_")

colnames(ASV_data_df) <- genus_names



sample_data$SampleID <- rownames(sample_data)
grp.names <- as.vector(sample_data$Treatment)

df_metadata <- data.frame(sample_data)

#feature.table: Data frame or matrix representing observed OTU table with OTUs (or taxa) in rows and samples in columns.
df_data

#meta.data: Data frame or matrix of all variables and covariates of interest.
sample_data

#sample.var: Character. The name of column storing sample IDs.
SampleID

#group.var: Character. The name of the main variable of interest. ANCOM-BC v1.0 only supports discrete group.var and aims to compare the change of absolute abundance across different levels of group.var.
grp.names


#zero.cut: Numerical fraction between 0 and 1. Taxa with proportion of zeroes greater than zero.cut are not included in the analysis.
0.01

#lib.cut: Numeric. Samples with library size less than lib.cut are not included in the analysis.


#neg.lb: Logical. TRUE indicates a taxon would be classified as a structural zero in the corresponding experimental group using its asymptotic lower bound.

ANCOM_BC(feature.table, 
grp.name, 
grp.ind, 
struc.zero, 
adj.method, tol.EM, max.iterNum, perNum, alpha)






#Run ANCOM-BC
ancombc(
  data = NULL,
  assay.type = NULL,
  assay_name = "counts",
  rank = NULL,
  tax_level = NULL,
  phyloseq = btphylo_atsacrifice_relab100_genus,
  formula,
  p_adj_method = "holm",
  prv_cut = 0.1,
  lib_cut = 0,
  group = NULL,
  struc_zero = FALSE,
  neg_lb = FALSE,
  tol = 1e-05,
  max_iter = 100,
  conserve = FALSE,
  alpha = 0.05,
  global = FALSE,
  n_cl = 1,
  verbose = FALSE
)
```


#Using biomehorizon to look at differential abundance across time


```{r}

## Install devtools
install.packages("devtools")


devtools::install_github("blekhmanlab/biomehorizon")
library(biomehorizon)  
otusample_diet <- otusample_diet

otusample_diet %>%
	arrange(desc(MCT.f.0002)) %>%
	select(1:7) %>%
	head()

head(metadatasample_diet)

metadatasample_diet <- metadatasample_diet

length(metadatasample_diet$subject[metadatasample_diet$subject == "MCTs01"])

paramList <- prepanel(otudata = otusample_diet, metadata = metadatasample_diet, subj = "MCTs01")


library(dplyr)

## Retrieve samples from MCTs01
otusample_subj1 <- otusample_diet %>%
	select(taxon_id, as.character((metadatasample_diet %>% filter(subject=="MCTs01"))$sample))
	

otusample_subj1 %>%
	filter(taxon_id == "taxon 2") %>%
	select(-taxon_id) %>%
	as.numeric()

paramList <- prepanel(otudata = otusample_diet, metadata = metadatasample_diet, subj = "MCTs01",
thresh_prevalence = 90, thresh_abundance = 1.5)


paramList <- prepanel(otudata = otusample_diet, metadata = metadatasample_diet, subj = "MCTs01",
otulist = c("taxon 1", "taxon 2", "taxon 10", "taxon 14"))

## Basic plot using default filtering thresholds
paramList <- prepanel(otudata = otusample_diet, metadata = metadatasample_diet, subj = "MCTs01")

horizonplot(paramList)

```



